Opposite and complementary roles of the two calcium thresholds for inducing LTP and LTD in models of striatal projection neurons

Version 1

Abstract

Synaptic plasticity has been shown to occur when calcium, flowing into the synapse due to incoming stimuli, surpasses a threshold level. This threshold level is modifiable through a process called metaplasticity. Some neurons, such as the striatal projection neurons, use different sources of calcium as the signal for synaptic strengthening (long-term potentiation, LTP) or weakening (long-term depression, LTD), resulting in them having two thresholds for inducing plasticity. In this study, we show opposite and complementary roles of metaplasticity in these two thresholds for inducing LTP and LTD on learning how to solve the linear and nonlinear feature binding problem (FBP and NFBP). In short, metaplasticity in one threshold (e.g. LTD) allows synaptic plasticity of the opposite type (e.g. LTP) to be properly expressed. This happens because metaplasticity in the LTD threshold protects strengthened synapses from weakening, thus allowing them to persistently increase during learning (and encode learned patterns). Similarly, metaplasticity in the LTP threhsold prevents weakened synapses from strengthening, thus allowing them to persistently decrease. Metaplasticity in both thresholds is necessary when synapses are clustered and the neuron needs to rely on supralinear dendritic integration for learning.

Introduction

Learning is thought to be implemented by several processes in the nervous system, one of which is synaptic plasticity - the modification of synapses based on incoming stimuli to the neuron (Citri and Malenka, 2007; Kandel et al., 2014; Takeuchi et al., 2014; Abraham et al., 2019). Most often, incoming stimuli cause calcium influx in the synapse, an important “second messenger” ion that triggers synaptic plasticity. Importantly, not all levels of calcium can elicit synaptic plasticity - threshold levels of calcium have been reported above which plasticity is triggered (Yang et al., 1999; Nishiyama et al., 2000; Cormier et al., 2001; Shindou et al., 2011). These calcium thresholds for inducing plasticity are themselves modifiable (plastic) through a process called metaplasticity (Abraham, 2008). Metaplasticity regulates the conditions under which plasticity occurs, i.e. under which synapses are modified.

In many neurons, such as hippocampal CA1 neurons, Purkinje neurons, cerebellar Golgi cells, and projection neurons from the core of the nucleus accumbens, the calcium signal for synaptic strengthening (long-term potentiation, LTP) and for synaptic weakening (long-term depression, LTD) comes from different sources (Nishiyama et al., 2000; Kohda et al., 1995; Binda et al., 2016; Locatelli et al., 2021; Ji and Martin, 2012; Cho et al., 2001). In the direct-pathway striatal projection neurons (dSPNs), a model of which is used in this study, calcium from NMDA receptors (NMDARs) triggers LTP when followed by activation of D1 dopamine receptors (D1Rs), while calcium from voltage-gated L-type channels (Cav1.3) triggers LTD if paired with inactivation of D1Rs (Shen et al., 2008; Yagishita et al., 2014; Fisher et al., 2017; Fino et al., 2010; Shindou et al., 2011; Iino et al., 2020). A threshold for eliciting LTD in SPNs has been experimentally quantified (Shindou et al., 2011). Although a similar threshold for LTP in SPNs has not been reported yet, presumably such a threshold exists, because weaker stimulation protocols do not induce LTP (Fino et al., 2005). (However, probably due to the local striatal microcircuitry, a range of different stimulation protocols can induce a range of different plasticity outcomes, as reviewed in Reynolds and Wickens (2002). An LTP threshold for calcium has been quantified in hippocampal CA1 neurons (Cormier et al., 2001).

By regulating when plasticity can be induced, metaplasticity has been proposed to prepare neural circuits and networks for learning, stabilize synaptic weights (prevent them from reaching very high or very low values) and to contribute to cognitive effects of disease when it is dysregulated (Hulme et al., 2013). All metaplasticity models so far have only one modifiable threshold, and typically, synaptic activity above the threshold induces LTP, while activity below the threshold induces LTD (Yger and Gilson, 2015; Bienenstock et al., 1982a; Benuskova and Abraham, 2006). Using a multicompartment model of an SPN, in this article we study what happens when there are two separate calcium thresholds for LTP and LTD, and what their role is in learning. We use a simple learning rule that models plasticity of the dSPN cortico-striatal synapse, to which we add metaplasticity in the two thresholds (Fig. 1A), and apply it to both the linear and nonlinear feature binding problem (FBP and NFBP). The results show that metaplasticity in both thresholds is necessary for successful learning of both tasks, with a specific, opposite and complementary role of each threshold.

Synaptic plasticity of the cortico-striatal synapse, the tasks for learning, and the dopamine signals for the two types of patterns in the tasks.

(A) Scheme describing the requirements for synaptic plasticity at the cortico-striatal synapse (adapted from Shen et al. (2008) by adding the calcium thresholds, which are the focus of this study). (B) The FBP and the NFBP, illustrated using features common to the visual system. SPNs might instead receive, e.g. sensory- and motor-related features. (C) Dopamine signals from the midbrain are assumed to arrive in the striatum after every pattern. Dopamine peaks are emitted after the relevant patterns, and dopamine pauses after the irrelevant patterns. (The midbrain and its projections are not explicitly modeled in this study, dopamine is programatically provided to the SPN a certain time after pattern arrival.) This figure is adapted from Figure 1 from Khodadadi et al. (2025).

Before presenting the results, we introduce the FBP and NFBP. The FBP refers to combining (binding) an object’s individual features, such as color, shape or motion, into a unified object (Roskies, 1999; Treisman, 1999; von der Malsburg, 1999; Hardcastle, 2017). Feature binding is a problem because features such as color, shape and motion, are each processed in different brain regions, and possibly at different times, and need to be combined to represent/construct the perception of the object they correspond to. By applying the simple plasticity and metaplasticity rule to the FBP and NFBP, we in fact study whether a single SPN can perform feature binding, and focus on the role of metaplasticity in learning. In its simplest form, the linear feature binding problem (FBP) consists of discriminating between two patterns, meaning that a neuron should spike when activated by one of them, and remain silent when activated by the other (Fig. 1B, top). Each pattern consists of two features, e.g. shape and color, and one of the features can have two values (e.g. ‘red’ and ‘yellow’ for the color feature, as opposed to just ‘strawberry’ for the shape feature in Fig. 1B). Each feature excites the neuron equally, and the term “linear” refers to the computational complexity of the problem, meaning that a neuron able to perform simple linear summation (integration) of incoming stimuli, such as synaptic potentials summated (added) at the soma, can solve the problem. The nonlinear feature binding problem (NFBP) consists of discriminating between four patterns, so that the neuron spikes for two patterns and remains silent for the other two (Fig. 1B, bottom). Each pattern also consists of two features, and in the NFBP both features can have two values (‘strawberry’ or ‘banana’ for shape, and ‘red’ or ‘yellow’ for color, shown in the second table in Fig. 1B). Again, each feature excites the neuron equally, and the additional layer of complexity comes from the requirement to spike for two different patterns whose features do not overlap (‘red strawberry’ and ‘yellow banana’). One of the theoretical solutions to the NFBP requires supralinear integration of inputs (Tran-Van-Minh et al., 2015). SPNs exhibit such supralinearities in the form of prolonged dendritic voltage elevations called plateau potentials, demonstrated in the next section. (Supralinear integration of inputs can also solve the simpler FBP.)

As mentioned above, in addition to calcium influx, dopamine signals are also required for plasticity in SPNs. Because of this, we assume that after a pattern is presented, dopamine signals are emitted from the midbrain, and they indicate whether the pattern was “rewarding” or not (Fig. 1C). The patterns for which the SPN should learn to spike are called relevant patterns, and those for which it should be silent are called irrelevant patterns. We assume that dopamine peaks are emitted for relevant patterns, and dopamine pauses for irrelevant patterns (Fig. 1C). After stimulation of the synapse, if the required calcium signal has reached its corresponding threshold, the dopamine signal determines the direction of plasticity. Dopamine peaks are assumed to activate the D1Rs, enabling LTP, and dopamine pauses are assumed to deplete dopamine enough so that D1Rs are not activated, enabling LTD.

Results

Linear and supralinear integration of inputs by SPNs and corresponding calcium elevations

Figure 2 compares linear integration of synaptic inputs distributed across the dendrites with supra-linear integration of clustered synaptic inputs on one dendrite (Fig. 2A). The number of synaptic inputs is varied from 1 to 30 synapses, and the evoked somatic and spine voltage, as well as the spine [Ca]NMDA and [Ca]L-type are shown in Figs. 2B and 2C. A comparison of the amplitudes of these four quantities between linear and supralinear integration is shown in Fig. 2D. With linear integration, the somatic voltage amplitude increases linearly with the number of synapses (Fig. 2B1 and blue line in Fig. 2D1). Conversely, the spine voltage amplitude remains approximately the same, because spines are distributed across the dendrites and are activated in the same way (Fig. 2B2 and blue line in Fig. 2D2). The similar amplitudes of the spines’ voltage result in similar [Ca]NMDA and [Ca]L-type elevations (Figs. 2B3, B4 and blue lines in Figs. 2D3, D4).

Linear versus supralinear integration of synaptic inputs by SPNs.

(A) Illustrations of the two scenarios of linear integration of distributed synaptic inputs by the soma, and supralinear integration of clustered synaptic inputs in a dendrite. In both scenarios, each synapse is activated with 3 incoming spikes within a window of 35 milliseconds. (B1-B4) Somatic voltage, spine voltage, spine [Ca]NMDA and spine [Ca]L-type evoked in the linear integration scenario (randomly distributed across the dendrites). Line color indicates the number of distributed synapses. (C1-C4) Somatic voltage, spine voltage, spine [Ca]NMDA and spine [Ca]L-type evoked in the supralinear integration scenario (synaptic cluster placed on one dendrite, in a 20-micrometer region approximately 120 micrometers away from the soma). Similarly, line color represents the size of the synaptic cluster, which is varied from 1 to 30 synapses. (D1-D4) The amplitude of the somatic voltage, spine voltage, spine [Ca]NMDA and [Ca]L-type compared between the cases of linear and supralinear integration. Results are averages over 20 trials, and in the case of synaptic clusters, over 8 different dendrites (clusters located in a 20-micrometer in a dendritic region starting at approximately between 120 micrometers from the soma).

On the other hand, when synapses are clustered closely together on a dendrite, SPNs can perform supralinear integration of synaptic inputs through plateau potenitals (Du et al., 2017; Plotkin et al., 2011). Plateau potentials are evoked by activating synaptic NMDARs, and, in this study, in addition by glutamate spillover which activates extrasynaptic NMDARs. We included glutamate spillover because it robustly provides an all-or-none jump in the (somatic) voltage amplitude Trpevski et al. (2023). This all-or-none supralinearity is important for solving the NFBP (Tran-Van-Minh et al., 2015). Fig. 2C1 shows the somatic voltage elevations when varying the size of the synaptic cluster from 1 to 30 synapses, while Fig. 2C2 shows the spine voltage in one spine of the synaptic cluster. Similarly, Figs. 2C3, C4 show the spine [Ca]NMDA and [Ca]L-type that result from stimulating the synaptic cluster. As explained in the Methods section, calcium from these two different sources is tracked separately in the SPN model. Also, [Ca]NMDA diffuses in the neuron, while [Ca]L-type does not, since it is assumed to be localized in a microdomain. As the cluster size is increased, increasingly larger and longer voltage elevations are produced, as a result of activating more NMDARs (Figs. 2C1, C2, traces for cluster sizes below 20 synapses). Once glutamate spillover occurs (here at a cluster size of 20 synapses), an all-or-none plateau potential appears, causing an all-or-none jump in the calcium concentration, as well (Figs. 2C1-C4).

This all-or-none behavior is also visible as a jump in the amplitudes of the somatic and spine voltage, and the spine [Ca]NMDA and [Ca]L-type in Figs. 2D1-D4. The somatic and spine plateau amplitude shows roughly linear increases before and after glutamate spillover, with the largest non-linearity being caused by the plateau (Figs. 2D1-D2, red lines). The plateau also causes a jump in [Ca]NMDA and [Ca]L-type, as well (Figs. 2D3-D4, red lines). The jump in all four quantities in Fig. 2D is an indication of a “threshold supralinearity”. When a threshold level of stimulation is achieved, the neuron’s somatic voltage amplitude is suddenly much larger, despite it being stimulated with only one additional synapse in the synaptic cluster.

Also, the comparison of linear and supralinear integration in Fig. 2D1 shows that, once a plateau appears, the same number of synapses trigger a much larger somatic voltage elevation when they are clustered compared to when they are randomly distributed across the dendrites. (It also shows a small difference between linear and supralinear integration when the number of stimulated synapses is less than 10. This arises because the synaptic clusters are always placed at approximately the same distance from the soma, while the distributed synapses can be randomly placed further away.)

Synaptic plasticity rule with metaplasticity solves both the FBP and NFBP

The synaptic plasticity rule is simple, and is given with Eq. 1. The synaptic weight w is increased if, during pattern presentation, [Ca]NMDA reaches a level above the LTP threshold, θLTP, and a dopamine peak arrives afterwards. Similarly, the weight is decreased if [Ca]L-type goes above the LTD threshold, θLTD, and a dopamine pause arrives afterwards. Synapses can be increased to a maximal level of wmax, and decreased to a minimal level of wmin, and weight updates are “multiplicative”, i.e. proportional to the current weight value. This formulation is very similar to existing plasticity rules (Rubin et al., 2001; Kistler, 2002; Gütig et al., 2003). The weight updates are also proportional to a learning rate η, and the time step for advancing the simulation, Δt. This rule is local, meaning that it depends on calcium levels locally at the synapse, and somatic spiking is not necessary for plasticity (similarly to Khodadadi et al. (2025)). This is beneficial for distal synapses, as backpropagating action potentials in general do not affect the local voltage which triggers calcium influx at distant dendritic locations. Also, at least in hippocampal CA3 pyramidal neurons, it has been reported that local dendritic depolarizations are necessary for plasticity, and somatic spiking is not, consistent with the fact that local calcium levels, driven by local voltage, are the signals triggering plasticity (Brandalise et al., 2016). Metaplasticity is also implemented in a simple way, given with Eq. 2. It only occurs if synaptic plasticity also occurs. During one update step, each threshold moves closer to the amplitude (maximum) of the calcium level evoked by an arriving pattern, with a rate ηθ (θLTP follows the amplitude of [Ca]NMDA, and θLTD follows the amplitude of [Ca]L-type). This implementation is very similar to the sliding threshold in the well-studied BCM learning rule, where the threshold follows a quantity derived from local synaptic activity (Bienenstock et al., 1982a). When a synapse is modified, both thresholds are updated. When using supralinear integration of inputs, we have added an upper threshold for LTP, ΘLTP (Eq. 3). This threshold serves to stop synaptic strengthening once synapses are strong enough to evoke a plateau. With it, synapses are not strengthened to the maximal value, wmax, thus avoiding strong supralinear integration for the irrelevant patterns. (This is important for solving the NFBP, as will be seen below.) The upper threshold for LTP only plays a role in synaptic plasticity, and is not taken into account in metaplasticity. This means that metaplasticity is performed in the same way as in Eq. 2, and uner the conditions in Eq. 1. In summary, each synapse is updated based on its local calcium levels, and has its own LTP and LTD thresholds. During learning, patterns in the FBP and NFBP are presented in random order, and after some time, each pattern is followed by a dopamine feedback signal (a dopamine peak or pause), during which plasticity occurs (Fig. 3A). The learning rule is always active (always “on”), avoiding separate training and testing phases sometimes used in other plasticity rules.

Learning outcome on the FBP and NFBP.

(A) The stimulation protocol used. Patterns are presented in random order. A pattern is represented with synaptic inputs arriving within a window of 35 ms, 3 spikes per synapse. The dopamine signal, during which synapses are updated, is given 400 ms after the start of a pattern, and lasts for 20 ms. The next pattern arrives 600 ms after the start of the current pattern. (B1) Illustration of the setup for the FBP with linear integration. Each feature is represented by 15 synapses distributed across the dendrites. (B2) Illustration of the setup for the FBP with supralinear integration. Each feature is represented by 10 synapses clustered in one dendrite. (B3) Illustration of the setup for the NFBP (only supralinear integration). Each feature is represented by 10 clustered synapses, and clusters are located on two different dendrites. (C1-C3) The somatic and dendritic voltages in the FBP with linear (C1) and supralinear (C2) integration, and the NFBP (C3), before and after learning (gray and black traces, respectively), elicited by the relevant and irrelevant pattern(s). (D1-D3) The performance on the FBP with linear (D1) and supralinear (D2) integration, and on the NFBP (D3). In (D1) and (D2) the performance is averaged over 50 trials, and in (D2) the cluster was placed at random in one of 8 dendrites in each trial. In (D3) the performance is averaged over 12 trials per input configuration, which are divided into two groups (with 18 and 13 input configurations, shown in Figure 3–figure supplement 1). (This totals to 216 and 156 trials in the two groups, respectively.) Each point in the plots is the average score over the 20 previous patterns, and the shaded area shows the standard deviation. The dashed lines are the threshold scores for solving the FBP and NFBP, which are 75% and 87.5%, respectively. (See the Methods section for what these threshold scores represent). Figure 3—figure supplement 1. All input configurations used in the NFBP.

Figure 3 shows the outcome of learning on the FBP and NFBP. For the FBP we have tested both linear integration of randomly distributed synaptic inputs (Figs. 3B1-D1) and supralinear integration of clustered synapses (Figs. 3B2-D2), while for the NFBP we tested only supralinear integration of clustered synapses in two different dendrites (Figs. 3B3-D3), since only supralinear integration can solve the NFBP (Tran-Van-Minh et al., 2015). Figs. 3C1-C3 show the voltage before and after learning for each pattern in the three scenarios. The SPN is silent before learning, and at the end of learning it has learned to spike to the relevant pattern(s). With supralinear integration, it does this by learning to evoke plateau potentials for the relevant pattern(s). For example, after learning (black traces), ‘red strawberry’ in Figs. 3C2 and 3C3 elicits a plateau in dendrite 1, which triggers somatic spiking. For the NFBP, the two relevant patterns evoke plateaus on separate dendrites (‘red strawberry’ on dendrite 1 and ‘yellow banana’ on dendrite 2, Fig. 3C3). Figs. 3D1, D2 show the performance on the FBP with linear and supralinear integration averaged over 50 trials, indicating that both modes of integration can solve the task. For the NFBP, in addition to the input configuration in Fig. 3B3, with three features on two dendrites, we also tested all possible configurations of four features on two dendrites which allow storage of the relevant patterns on separate dendrites. (These input configurations are listed in Figure 3–figure supplement 1.) Fig. 3D3 shows that when up to three features are present in one dendrite, the NFBP is successfully solved. When four features are present in one or both dendrites, usually only one relevant pattern is learned. Solving the NFBP in this case requires additional mechanisms such as branch plasticity, as suggested by another study with more abstract neuron models and plasticity rules (Legenstein and Maass, 2011).

Before separately examining the effects of each calcium threshold on learning, it is first necessary to understand how the learning rule successfully solves the tasks by modifying the synaptic weights and thresholds. Figure 4 shows how the synaptic weights and the two calcium thresholds change during learning: they show the same qualitative behavior for both the FBP and NFBP. For the FBP, the synapses carrying the features ‘red’ and ‘strawberry’, which together form the relevant pattern, are strengthened (Figs. 4B1, B2). The synapses for ‘yellow’ only weaken, since they are only activated as part of the irrelevant pattern (which is followed by a dopamine pause, a signal for LTD). For the NFBP, where each relevant pattern is stored in a separate dendrite, the outcome is similar: the synapses for ‘red’ and ‘strawberry’ on dendrite 1 and the synapses for ‘yellow’ and ‘banana’ on dendrite 2 are strengthened (Figs. 4B3, B4). The remaining synaptic cluster, belonging to one irrelevant pattern in each dendrite, is weakened (the (the synapses for ‘yellow’ on dendrite 1 and the synapses for ‘strawberry’ on dendrite 2). In all cases, weakening stops when the evoked [Ca]L-type falls below the LTD threshold. For the FBP with linear integration at the soma, many of the randomly distributed synapses for the relevant pattern are strengthened almost to the maximal value wmax (Fig. 4B1). On the other hand, for the FBP with supralinear integration and the NFBP, the clustered synapses for the relevant pattern(s) are not strengthened to their maximal values. Instead, they stabilize once they are strong enough to evoke a plateau (Figs. 4B2-B4). This is due to the upper threshold for LTP, ΘLTP, shown with dashed lines in Figs. 4C2-C4. (The stabilization takes about twice as long in the NFBP simply because there are twice as many patterns in the task.) When plateaus are evoked by the relevant patterns, they drive somatic spiking (as indicated above in Figs. 3C2, C3). An irrelevant pattern stimulates only half of the strengthened synapses on one dendrite. This is not enough for a plateau potential, and even though a supralinear voltage elevation is elicited, no somatic spiking is triggered (Fig. 3C2, ‘yellow strawberry’ on dendrite 1 for the FBP and Fig. 3C3, ‘yellow strawberry’ on dendrite 1 and ‘red banana’ on dendrite 2 for the NFBP). Nevertheless, these voltage elevations produce high calcium of both types, but, as will be seen below, the synapses involved are protected from weakening as a result of metaplasticity.

The synaptic weights and calcium thresholds during learning of the FBP and NFBP.

(A) Illustrations of the setups for the FBP with linear (A1) and supralinear (A2) integration, and the NFBP (A3) (same as in Fig. 3). (B-D) The evolution of synaptic weights (B), LTP thresholds (C) and LTD thresholds (D) in the FBP with linear (B1-D1) and supralinear (B2-D2) integration, and the NFBP (B3-D3 and B4-D4). In (B) all synaptic weights for the features are shown, while in (C) and (D) only one synapse per feature is chosen to show its calcium threshold (solid lines). Dots in (C) and (D) represent the amplitudes of [Ca]NMDA and [Ca]L-type, respectively, during pattern presentation. For clarity, the calcium amplitudes for some patterns are omitted. (C) and (D) are shown in larger scale in Figure 4–figure supplement 1, with much more detail in the calcium amplitudes. Arrows show the moment in time when the weakened synapses become low enough for their calcium levels to mostly stay below the calcium thresholds. Figure 4—figure supplement 1. Calcium thresholds and calcium amplitudes from Fig. 4 in greater detail. Figure 4—figure supplement 2. The effect of having no upper LTP threshold on supralinear integration. Figure 4—figure supplement 3. The effect of the learning rate on learning. Figure 4—figure supplement 4. The effect of the metaplasticity rate on learning.

The LTP and LTD thresholds for just one synapse from each feature are shown in Figs. 4C and 4D, respectively (the same plots are shown with larger panels in Figure 4–figure supplement 1 for better clarity). As mentioned above, they show the same behavior for both the FBP and NFBP, the only difference being that they stabilize at different values due to the different calcium levels evoked by distributed and clustered synaptic inputs. Also, the LTP and LTD thresholds show the same behavior when compared to each other. For example, looking at the strengthened synapses first, they increase to follow the calcium amplitudes evoked by each pattern presentation, and stabilize at these values towards the end of the simulation (thresholds for ‘red’ and ‘strawberry’ tend to the dots of the same color above them in Figs. 4C1-C3 and 4D1-D3, and the thresholds for ‘yellow’ and ‘banana’ tend to the dots of the same color above them in Figs. 4C4, D4; this is better visible in Figure 4–figure supplement 1 and also highlighted with the light blue regions in Figure 4–figure supplement 1B3). In the case of linear integration they stabilize at low values, because distributed synapses do not have cooperative voltage effects (due to the large distances between them), and low calcium levels are evoked as a result (Figs. 4C1, D1). Instead, with supralinear integration, the plateaus evoked by the clustered synapses cause much larger voltage and calcium elevations, so the thresholds rise to much higher levels (Figs. 4C2-C4 and 4D2-D4).

We next look at the thresholds for the weakened synapses (‘yellow’ synapse in Figs. 4C1-C3 and 4D1-D3 and ‘strawberry’ synapse in Figs. 4C4, D4). At the beginning, the thresholds increase. Later, once the synapses are weakened enough so that the evoked [Ca]L-type falls below the LTD threshold, they reach a stable level. For example, in Figs. 4C2, C3 and 4D2, D3, at the beginning of learning, yellow dots can be seen above the ‘yellow’ thresholds, while later they almost always remain below the thresholds. Similarly, in Figs. 4C4, D4, light red dots are at first seen above the ‘strawberry’ thresholds, while later they are found below them. are at first seen above the ‘strawberry’ thresholds, while later they are found below them. This transition of the [Ca]L-type amplitudes from above to below the LTD thresholds is marked with the arrows in Figs. 4B2-4-D2-4, corresponding to the moment in time where synapses are weakened enough. (This is more clearly visible in Figure 4–figure supplement 1, and it is also highlighted with the green and brown regions in Figure 4–figure supplement 1B3, and the pink region where yellow dots are absent.) We mention that in the FBP, where the weakened synapses are only activated by the irrelevant pattern, both of their thresholds are only updated during LTD. On the other hand, in the NBFP the weakened synapses and their thresholds can be updated during both LTP and LTD, because these synapses are also activated by a relevant pattern. (However, in Fig. 4 this only happens at the very beginning, when the weakened synapses also undergo LTP. As they are being weakened, their LTP threshold is raised, so [Ca]NMDA amplitudes later do not go above the LTP threshold.) In summary, once a calcium amplitude falls below its threshold, one of the necessary conditions to trigger plasticity and metaplasticity is not met, and the thresholds, as well as the synapses, are not updated anymore. This will be much more evident in the next section, where we study the role of metaplasticity.

We stress once again that with supralinear integration, one of the strengthened synaptic clusters in each dendrite is activated by both a relevant and an irrelevant pattern (the ‘strawberry’ cluster in Figs. 4B2, B3 and the ‘yellow’ cluster in Fig. 4B4). Importantly, the LTD threshold for these synapses is raised because of the plateaus (Figs. 4D2-D4: ‘strawberry’ synapse in Figs. 4D2, D3 and ‘yellow’ synapse in Fig. 4D4). Because of this, the high [Ca]L-type levels in these synapses evoked from an irrelevant pattern are below the LTD threshold and are not enough to weaken them (e.g. light red dots in Figs. 4D2, D3 just above the ‘yellow’ threshold after 100 s are far below their ‘strawberry’ threshold; similarly yellow dots in Fig. 4D4 located just above the ‘strawberry’ threshold after 100s are far below their ‘yellow’ threshold; dots also highlighted in the pink region of Figure 4–figure supplement 1B3). In this way, these strengthened synapses are protected from weakening, allowing them to store one part of the relevant pattern. This effect will also be much more evident in the next sections, where we look at the effects of metaplasticity on learning.

Briefly, we have also examined the effects of parameters in the learning rule. In supralinear integration, the upper plasticity threshold ΘLTP prevents the strengthened synapses to saturate at their maximal levels. Without it, the FBP is still solved, even though weights attain their maximal values (Figure 4–figure supplements 2B1 and 2E1). It is successfully solved because the features innervate only one dendrite, and the saturated weights do not cause somatic spiking when activated by the irrelevant pattern (Figure 4–figure supplement 2A1). However, the NFBP is not solved without an upper threshold ΘLTP, because the features innervate two dendrites, and the maximally strengthened synapses cause supralinear voltage elevations in the dendrites for the irrelevant patterns, which trigger somatic spiking (Figure 4–figure supplements 2A2, 2B2, 2B3 and 2E2). The maximally strengthened synapses also drive more weakening in the weakened synapses, compared to when an upper threshold is present (Figure 4–figure supplement 2B). This is due to the elevated [Ca]L-type evoked when an irrelevant pattern activates the maximally strengthened synapses (‘strawberry synapses in Figure 4–figure supplements 2B1, B2 and ‘yellow’ synapses in Figure 4–figure supplement 2B3). The elevated [Ca]L-type for the irrelevant patterns is also visible in the elevated LTD thresholds for the weakened synapses (‘yellow’ synapse in Figure 4–figure supplement 2D1, D2 and ‘strawberry’ synapse in Figure 4–figure supplement 2D3, cf. Figs. 4D2-D4). (The LTP thresholds for the weakened synapses are also elevated, due to higher [Ca]NMDA produced by the maximally strengthened synapses. This is visible in Figure 4–figure supplement 2C when compared to Figs. 4C2-C4)

The learning rate η has an effect on the speed of learning - more pattern presentations are required with lower learning rates, and vice versa (Figure 4–figure supplement 3). The rate of threshold adaptation (metaplasticity rate) ηθ has no significant effect on learning to solve the FBP and NFBP (Figure 4–figure supplement 4).

Metaplasticity in the LTD threshold prevents weakening of strengthened synapses, protecting learned patterns

The role of each threshold is easily examined with this plasticity and metaplasticity learning rule by turning metaplasticity off, i.e. fixing a threshold’s value during learning. We focus on the NFBP because it is the harder task, but we also check metaplasticity’s effect on the FBP. Fixing the LTD threshold (to the initial low value in the simulation) shows that the synapses encoding the relevant patterns cannot strengthen properly (Figs. 5C1, C2). Before we show why this is so, it is important to remember that in the NFBP, a relevant pattern shares one feature with both irrelevant patterns (e.g. ‘red strawberry’ shares the color feature with ‘red banana’ and the shape feature with ‘yellow strawberry’). Because of this, half of the synapses activated by a relevant pattern are also always activated when an irrelevant pattern arrives.

Learning outcome on the NFBP without metaplasticity in the LTD threshold.

(A) An illustration of the setup for the NFBP (same as in Fig. 3B3). (B) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for all four patterns. (C-E) The evolution of synaptic weights (C), LTP thresholds (D) and LTD thresholds (E) for the synapses in each dendrite. (E) shows the fixed LTD thresholds. In (C) all synapses are shown, and in (D, E) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (D) and [Ca]L-type (E) during pattern presentation, omitting some patterns for clarity. (F) The performance on the NFBP without metaplasticity in the LTD threshold. Figure 5—figure supplement 1. The effect of a fixed LTD threshold on solving the FBP.

As learning progresses, the relevant patterns cause strengthening in the synapses that they activate. However, because there is no metaplasticity in the LTD threshold, irrelevant patterns always cause higher [Ca]L-type than the fixed LTD threshold (Figs. 5E1, E2 – LTD thresholds stay fixed during learning, and [Ca]L-type is always above θLTD). This means that when the strengthened synapses are activated by an irrelevant pattern, they will always get weakened. Conversely, when metaplasticity in the LTD threshold is active, it follows the increased levels of [Ca]L-type elicited by a plateau (evoked due to previous synaptic strengthening, as shown previously in Figs. 4D2–D4). This elevated LTD threshold protects the strengthened synapses from weakening when they are activated by an irrelevant pattern. As mentioned above, they are protected because an irrelevant pattern, activating only half of the strengthened synapses, does not evoke a plateau, resulting in [Ca]L-type lower than the elevated LTD threshold in Figs. 4D2 - 4D4 (e.g. light red dots in pink region in Figure 4–figure supplement 1B3 are below the elevated ‘strawberry’ threshold).

Without metaplasticity in the LTD threshold, strengthened synapses are not protected from weakening. As a result, the synaptic weights for the relevant pattern cannot stabilize (Figs. 5C1, 5C2), and no plateaus are evoked to trigger somatic spiking (Fig. 5B), so the NFBP cannot be learned (Fig. 5F). The LTP threshold follows the level of [Ca]NMDA and also does not stabilize (Figs. 5D1, 5D2). Also, because the LTD threshold is fixed to a low value, the synapses belonging to the weakened feature in each dendrite are almost weakened to the minimal value wmin. In summary, by protecting synapses from weakening, the threshold for weakening allows proper synaptic strengthening.

In the FBP, metaplasticity in the LTD threshold has the same effect. However, due to there being only two patterns in the FBP, the synapses for only one feature in the relevant pattern cannot stabilize (the ‘strawberry’ synapses, Figure 5–figure supplement 1B). This is because the synapses for the feature ‘red’ are only activated by the relevant pattern and, as a result, can only ever be strengthened. Phrased differently, they are never activated by an irrelevant pattern, which would cause them to weaken and be influenced by the fixed LTD threshold. Because the ‘strawberry’ synapses do not stabilize, the clustered synapses in the FBP with supralinear integration are not strong enough to evoke plateaus, and the FBP is not learned (Figure 5–figure supplements 1A2 and 1E2). On the other hand, in the FBP with linear integration, despite these synapses not stabilizing, the strengthening is enough to drive somatic spiking (Figure 5–figure supplement 1A1). The FBP with linear integration is solved but with a lower performance (Figure 5–figure supplement 1E1).

Metaplasticity in the LTP threshold prevents strengthening of weakened synapses, promoting the weakening of unnecessary synapses

On the other hand, the LTP threshold has an opposite effect on learning. If this threshold is fixed, the feature belonging to the irrelevant pattern in each dendrite cannot be weakened properly (Figs. 6C1, C2). Because of the low value of the LTP threshold, whenever an irrelevant pattern activates these synapses, the evoked [Ca]NMDA is still greater than the threshold (Figs. 6D1, D2), triggering synaptic strengthening even if synapses have been previously weakened. This means that when metaplasticity in the LTP threshold is active, by following the levels of [Ca]NMDA (e.g. LTP threshold for ‘yellow’ in Figs. 4C1, C2, and for ‘strawberry’ in Fig. 4C3), it prevents these synapses from strengthening when they are activated by the relevant pattern (which can only be stored fully in the other dendrite). As a result, the synapses for ‘yellow’ in Figs. 4B1, B2, and for ‘strawberry’ in Fig. 4B3 are successfully weakened, while in Figs. 6C1, C2 they are not. (The LTD threshold behaves as usual for the strengthened synapses, protecting them from weakening, as shown in Figs. 6C1, C2 and 6E1, E2. However, for the weakened synapses, it does not stabilize.)

Learning outcome on the NFBP without metaplasticity in the LTP threshold.

(A) An illustration of the setup for the NFBP (same as in Fig. 3B3). (B) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for all four patterns. (C-E) The evolution of synaptic weights (C), LTP thresholds (D) and LTD thresholds (E) for the synapses in each dendrite. (D) shows the fixed LTP thresholds, and the dashed lines show the upper LTP threshold, ΘLTP. In (C) all synapses are shown, and in (D, E) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (D) and [Ca]L-type (E) during pattern presentation, omitting some patterns for clarity. (F) The performance on the NFBP without metaplasticity in the LTP threshold. Figure 6—figure supplement 1. The effect of a fixed LTP threshold on solving the FBP.

Without metaplasticity in the LTP threshold, the NFBP is on average not solved (Fig. 6F). Despite both relevant patterns being stored by each dendrite, the performance is lowered because the SPN sometimes also spikes for the irrelevant patterns (Fig. 6B). Also, in the input configurations containing four features in one dendrite, the average performance even drops below 50 % (green trace in Fig. 6F). In summary, by preventing synapses from strengthening, the threshold for strengthening allows proper synaptic weakening.

In the FBP, a fixed LTP threshold does not have an effect on learning (Figure 6–figure supplement 1). The synaptic weights behave the same as with metaplasticity in the LTP threshold being active (Figure 6–figure supplement 1B, cf. Figs. 4B1, B2). As a result, the FBP is solved with both linear and supralinear integration (Figure 6–figure supplements 1A and 1E). This is also due to there being only two patterns in the FBP. The weakened ‘yellow’ synapses, whose proper weakening could be affected by the fixed LTP threshold, are only activated during the irrelevant pattern (‘yellow strawberry’). Because they are never activated by the relevant pattern, they can never strengthen, and as a result they remain unaffected by the fixed LTP threshold.

Both thresholds need to be updated during one synaptic modification in order to solve the NFBP

The rule is formulated so that during one synaptic modification (e.g. LTP) the thresholds for both LTP and LTD are updated. Alternatively, it could be formulated so that during LTP, only the threshold for LTP is updated, and during LTD, only that for LTD is updated. We test this variant in this section, and the results from the previous two sections already give an intuition how this variant would work. Updating only the corresponding threshold during one synaptic modification means that the other threshold is, in effect, kept fixed during that plasticity update, making metaplasticity only partially active. The effects are seen in Figs. 7C1, C2. Synapses are neither properly strength-ened, nor weakened, at least within the length of the simulation. For example, one of the relevant patterns (‘red strawberry’) shows a trend of strengthening, but is also being weakened (Fig. 7C1). This is because as the LTD thresholds for ‘red’ and ‘strawberry’ move higher (Fig. 7E1), LTD in these synapses is (sometimes) prevented, and a trend of synaptic strengthening appears. Also, as the LTP threshold for ‘yellow’ increases (Fig. 7D1), it prevents these synapses from strengthening, and as a result, they weaken to a degree.

Learning outcome on the NFBP with partial metaplasticity, where the LTP threshold is updated only during LTP, and the LTD threshold is updated only during LTD.

(A) An illustration of the setup for the NFBP (same as in Fig. 3B3). (B) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for all four patterns. (C-E) The evolution of synaptic weights (C), LTP thresholds (D) and LTD thresholds (E) for the synapses in each dendrite. In (C) all synapses are shown, and in (D, E) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (D) and [Ca]L-type (E) during pattern presentation, omitting some patterns for clarity. (F) The performance on the NFBP with partial metaplasticity. Figure 7—figure supplement 1. The effect of partial metaplasticity on solving the FBP.

However, since the LTD threshold is not updated during LTP, it does not follow the [Ca]L-type levels generated by the relevant patterns. It is updated only during LTD, following the [Ca]L-type levels from irrelevant patterns. For example, in dendrite 1, the ‘red banana’ activates only the cluster representing ‘red’, evoking little [Ca]L-type and, as a result, the LTD threshold for the ‘red’ synapse is lower (Fig. 7E1). On the other hand, the ‘yellow strawberry’ activates two synaptic clusters (Fig. 7A1), causing a larger voltage elevation and consequently larger [Ca]L-type, so the LTD thresholds for these synapses move higher (Fig. 7E1). However, because of the trend of synaptic strengthening in the ‘red’ and ‘strawberry’ synapses, the irrelevant patterns sometimes still evoke [Ca]L-type higher than the LTD thresholds. As a result, the synapses for ‘red’ and ‘strawberry’ in dendrite 1 are not protected from weakening (when activated as part of the irrelevant patterns). Also, as the LTD threshold for ‘yellow’ increases in the first 70 s of the simulation (Fig. 7E1), the synapses for ‘yellow’ show a trend of increasing in this period, as well, before being decreased afterwards (when the ‘yellow’ LTP threshold is increased).

On the other hand, in dendrite 2 the relevant pattern is not strengthened, and it is not clear whether more learning trials will cause both relevant patterns to be stored in the two dendrites. On average, the NFBP is not solved in this case (Fig. 7F).

Finally, this variant is similar to not having metaplasticity at all (Fig. 8). Without metaplasticity, no synapses stabilize, i.e. they are neither strengthened nor weakened (Fig. 8C1, C2).

Learning outcome on the NFBP without metaplasticity (the LTP and the LTD thresholds are fixed).

(A) An illustration of the setup for the NFBP (same as in Fig. 3B3). (B) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for all four patterns. (C) The evolution of synaptic weights in both dendrites. (D, E) The fixed LTP thresholds (D) and LTD thresholds (E) in each dendrite, with calcium amplitudes during pattern presentation shown with dots (omitting some patterns for clarity). (F) The performance on the NFBP without metaplasticity. Figure 8—figure supplement 1. The effect of having no metaplasticity on solving the FBP.

For the FBP, the results in the previous two sections again give an intuition for how partial meta-plasticity would work. As shown in Figure 6–figure supplement 1, even when the LTP threshold is fixed, it does not have an effect on solving the FBP. Similarly, neither does it have an effect when it is partially active: the LTP threshold is merely less elevated than compared to when metaplasticity is always active (Figure 7–figure supplement 1C, cf. Figs. 4C1, C2). On the other hand, partial meta-plasticity in the LTD threshold is not enough to protect strengthened synapses from weakening. The LTD threshold does not follow [Ca]L-type fast enough (Figure 7–figure supplement 1D), and as a result the ‘strawberry’ synapses are not strengthened properly (even though, for example, their weights stabilize when using supralinear integration, Figure 7–figure supplement 1B). Similarly to Figure 5–figure supplement 1, this does not affect learning to solve the FBP with linear integration (Figure 7–figure supplement 1A1 and 1E1), but it does affect learning of the FBP with supralinear integration (Figure 7–figure supplement 1A2 and 1E2).

The effects of having no metaplaticity on the FBP are similar to those of having partial meta-plasticity (Figure 8–figure supplement 1). Again, the synapses encoding one of the features in the relevant pattern cannot stabilize (those for ‘strawberry’, Figure 8–figure supplement 1B). This only moderately affects the performance on the FBP with linear integration (Figure 8–figure supplement 1A1 and 1E1), but is necessary for solving the FBP with supralinear integration (Figure 8–figure supplement 1A2 and 1E2). In summary, this shows that in the FBP, metaplasticity is important when a feature is represented with clustered synapses. It allows their synaptic weights to stabilize, and to evoke plateau potentials that can drive somatic spiking. Conversely, when a feature activates distributed synapses across the dendrites, the FBP can still be solved without metaplasticity, even if some synapses do not stabilize.

Discussion

In this article we studied the effects of metaplasticity in two separate calcium thresholds for inducing synaptic plasticity, one of which is a threshold for LTP, and the other for LTD. Plasticity occurs when calcium levels from different sources, implicated in synaptic plasticity in SPNs, reach levels above the thresholds. We studied the effects of metaplasticity on how well the SPN learns the FBP and NFBP, tasks which require linear and supralinear summation of synaptic inputs, respectively. The results showed that:

Without metaplasticity, the NFBP cannot be solved. It is also necessary for solving the FBP with supralinear integration, but it is not necessary for solving the FBP with linear integration.

Metaplasticity in the LTD threshold protects learned patterns from being weakened, allowing the LTP process to properly strengthen (encode) the patterns. In effect, the threshold for weakening allows proper synaptic strengthening.

Metaplasticity in the LTP threshold prevents unnecessary features from being strengthened, allowing the LTD process to persistently weaken them. In effect, the threshold for strengthening allows proper synaptic weakening.

For successful learning of the NFBP, metaplasticity in both thresholds needs to occur during one synaptic modification, irrespective of whether the modification is LTP or LTD.

In summary, the two calcium thresholds have opposite and complementary roles in regulating the strengthening and weakening of synapses. The ultimate effects of metaplasticity are unexpected, at least at first glance, since the threshold for one plasticity type (e.g. LTD) allows proper expression of the other plasticity type (e.g. LTP). Metaplasticity is particularly important for clustered synapses, where the cooperative, supralinear voltage elevations can cause large calcium influx - it allows synpatic plasticity to adapt to this large calcium influx. By regulating plasticity in clustered synapses, metaplasticity allows neurons to perform computationally difficult tasks (such as the NFBP) that rely on supralinear integration by the clustered synapses. Simpler tasks that can be solved with linear integration of synaptic inputs (such as the FBP) do not require metaplasticity.

This study also provides a basis for exploring other metaplasticity rules. Here we only considered metaplasticity that is activated as a consequence of synaptic plasticity. However, stimulation that does not cause synaptic plasticity can also activate metaplasticity, affecting the induction of future plasticity (Abraham, 2008). Also, we only considered metaplasticity evoked in the same synapses activated by presynaptic spikes (homosynaptic metaplasticity). However, heterosynaptic metaplasticity mechanisms also exist, where the evoking stimuli originate from different synapses or different neurons (Abraham, 2008; Abraham et al., 2019; Hulme et al., 2014). Furthermore, as detailed calcium models for other neurons are developed, the contributions of different calcium sources in triggering synaptic plasticity can be studied using this simple learning rule as a basis. Regarding SPNs specifically, more realistic metaplasticity rules can also be developed once future experimental findings appear.

We have made two assumptions when formulating the learning rule. The first is that metaplasticity operates in cortico-striatal synapses, which is not a strong assumption, since metaplasticity has been reported for many synapses. The second – a major assumption – is the upper threshold for synaptic strengthening, which means that high [Ca]NMDA during LTP conditions (a dopamine peak) prevents synaptic plasticity (blocks LTP). It is an open question whether this assumption is true. Nevertheless, there are some molecular mechanisms which could form the basis for such an upper threshold. In the neuromodulatory branch of the signaling network in Fig. 1 (the branch activated by dopamine), the enzyme adenylyl cyclase 5, which transduces the dopamine signal from the D1R, is inhibited by calcium. This indicates that high [Ca]NMDA might prevent synaptic strengthening already very early in the signaling network. Further downstream in the signaling network, calcium activates PP2B through calmodulin, and PP2B inactivates PKA, an enzyme necessary for synaptic strengthening (Church et al., 2021; Yagishita et al., 2014). This is another route through which high [Ca]NMDA could prevent LTP. Due to the complexity of the signaling network, the influence of high [Ca]NMDA on synaptic plasticity requires further study, and kinetic models of the cortico-striatal synaptic signaling network which incorporate the many molecular interactions might provide some answers (Lindskog et al., 2006; Nakano et al., 2010; Nair et al., 2015).

In summary, at least when it comes to solving the NFBP, the proposed learning rule highlights the importance of deciding when to stop updating the synaptic weights. The two processes (mechanisms) that regulate when synaptic plasticity occurs are the upper plasticity threshold for LTP and metaplasticity. Without the upper plasticity threshold, all strengthened synapses would reach their maximally attainable synaptic weight. Without metaplasticity, the synaptic weights would not stabilize. Important future questions relate to the various such possible mechanisms and their effects on learning, as it is likely that different types of synapses possess different metaplasticity mechanisms, which support specific functions in various brain regions. Some examples of different metaplasticity mechanisms are known in the hippocampus. In the Schaffer collateral synapses to the CA1 area, prior activation of NMDARs (which does not induce long-term plasticity) temporarily increases the LTP threshold and prevents LTP, while prior activation of mGluRs facilitates LTP (Huang et al., 1992; Cohen and Abraham, 1996). In the synapses from the perforant path to the dentate gyrus, prior stimulation at theta frequency typical for the hippocampus, lowers the threshold for both LTP and LTD (Christie et al., 1995; Christie and Abraham, 1992). Additionally, prior stimulation at a high frequency also lowers the LTD threshold in both unchanged and in previously strengthened synapses (Holland and Wagner, 1998).

Finally, we have recently participated in another study asking whether SPNs could solve the NFBP, using a different learning rule that incorporates more biological complexity/detail (Khodadadi et al., 2025). That study focused on dendritic integration which was much less supralinear, i.e. the dendrites were in a regime where they produced so-called “boosting” nonlinearities, that are likely common physiologically. Completely solving the NFBP required both excitatory and inhibitory synaptic plasticity, regardless of whether the synaptic clusters were closer or further away from the soma. On the other hand, the all-or-none plateau potentials in this study achieve a very high score on the NFBP with excitatory plasticity alone. This indicates that with a threshold non-linearity, the SPNs might completely solve the task without requiring inhibitory synapses and even for very distally located synaptic clusters. Whether this is true will be explored in another study, as theoretical considerations suggest that the NFBP can be solved with excitatory synapses only (Tran-Van-Minh et al., 2015).

Methods

Plasticity rule

We highlight again that, similar to Khodadadi et al. (2025), the learning rule is:

local: only signals at the local synaptic site are used to trigger plasticity, calcium- and reward-based:

local calcium levels trigger plasticity, the type of which is decided by the dopamine reward signal, and

always “on”: no separate training and testing phases are needed because learning is regulated by metaplasticity.

Differently from Khodadadi et al. (2025), the learning rule has metaplasticity in both calcium thresholds, and is also (much) simpler, making it suitable for studying the thresholds’ roles in synaptic plasticity.

The parameters used in the plasticity rule are given in Table 1. The values of wmax and wmin are not varied throughout the simulations. They describe the maximal and minimal values that the synaptic weight can take on in the model. Whether these values are reached depends on the metaplasticity mechanisms (the values of the calcium thresholds), and, in the case of supralinear integration, the upper plasticity threshold ΘLTP, which is a fixed parameter that can be set at any value in the region of the supralinearity (the jump) in [Ca]NMDA in Fig. 2D3 (we used 40 μM). The learning rate η and the metaplasticity rate ηθ are also kept fixed throughout all simulations, except in Figure 4–figure supplements 3 and 4, where they are varied to study their effect on learning. Finally, the LTP and LTD thresholds, θLTP and θLTD, are initialized to low values, θLTP, 0 and θLTD, 0, which are also the minimal values that these thresholds can take on. They are initialized to low values in order to showcase learning in the synapses. Otherwise, if they are set to high values, they may act to protect synapses from weakening (high θLTD, 0) or prevent them from strengthening (high θLTP, 0), as is shown with Figs. 5 and 6. The synaptic weights are initialized uniformly at random from the interval [0.3, 0.35].

Neuron model

We use a morphologically realistic, multicompartmental model of a direct-pathway striatal projection neuron from the collection of 71 such models in Lindroos and Hellgren Kotaleski (2020) (index 34 in the collection). As detailed in Lindroos and Hellgren Kotaleski (2020), the ionic conductances in all models in the collection have been selected with a search process to produce frequency-current (F-I) responses and dendritic calcium elevations arising from backpropagating action potentials that match experimental data, and in addition produced a suitable number of spikes to injected current, no voltage oscillations leading up to the spikes, proper afterhyperpolarization behavior, and an F-I slope within three standard deviations of the experimentally determined mean.

The main contributor to supralinear integration in the SPN model is the activation of NMDARs. This has been shown, for example, in Fig. S7 in Du et al. (2017), where supralinear integration with clustered synapses is similar in an active and a passive dendrite in the SPN. This means that the NMDA synapses, and not the SPN’s ion channels, are causing it. SPN–specific ion channels do have a large effect on the SPN’s elecrophysiological behavior – the lowered resting potential of around −85 mV is a result of the strongly inwardly rectifying K+ channels (Nisenbaum and Wilson, 1995; Wilson, 2005). Because of this, to raise their somatic voltage, the SPNs require strong synaptic drive from distributed synaptic inputs across the dendrites, or a localized activation of clustered synapses which generates a plateau potential. With this behavior, SPNs are said to have two states: a “down state”, which requires strong synaptic drive to reach the firing threshold, and an “up state”, where they are more sensitive to incoming synaptic inputs (Plotkin et al., 2011). It has also been suggested that plateau potentials themselves cause the transition from a “down” state to an “up” state (Oikonomou et al., 2014). Plateau potentials in SPNs normally do not cause somatic spiking without additional synaptic inputs, which is why in the simulations we have used elevated background noise.

In this study, synapses are placed on spines, which are not present in the original model in Lindroos and Hellgren Kotaleski (2020). The spines consist of a neck and head with neck and head with lengths and diameters lneck = 0.5μm, lhead = 0.5μm, and dneck = 0.125μm, dhead = 0.5μm, respectively. Axial resistance in all compartments of the model is 150 Ω ⋅ cm, except for the spine neck, where it is 1130 Ω ⋅ cm (Dorman et al., 2018). The spines contain the inwardly rectifying K+ channel, and the same voltage-gated calcium channels as in the dendrites. The spine calcium channel conductances have been manually tuned to match the relative proportions determined in Carter and Sabatini (2004) and Higley and Sabatini (2010), as well as the Ca2+ concentration amplitudes arising from backpropagating action potential (bAP) stimulation as in Fig. 2 of Shindou et al. (2011).

Calcium dynamics and diffusion model

Because calcium from two different sources triggers LTP and LTD in SPNs, we tracked calcium from the two sources separately. Calcium from NMDA channels, when followed by a peak in dopamine, triggers LTP (Shen et al., 2008; Yagishita et al., 2014; Fino et al., 2010; Fisher et al., 2017). Calcium from Cav1.3 channels, when paired with a pause in dopamine, has been shown to be necessary for LTD (Wang et al., 2006; Fino et al., 2010; Shindou et al., 2011; Iino et al., 2020).

Without implementing calcium diffusion explicitly, the accumulation of [Ca]NMDA in a spine shows a strange effect when the synaptic cluster size is gradually increased, which is that the largest plateaus do not correspond to the largest calcium amplitude (Figs. 9A, B). This is simply a result of less NMDAR membrane current flowing in as the plateau’s voltage approaches the NMDAR reversal potential in Eq. 6. Larger plateaus giving rise to lower [Ca]NMDA amplitudes could pose a problem to the learning rule, since the decision when to trigger plasticity and metaplasticity is made by comparing [Ca]NMDA amplitude against the LTP threshold. In a synaptic cluster that is being strengthened, but starts to evoke a smaller [Ca]NMDA amplitude after some strengthening due to the effect in Figs. 9A, B, learning might be prematurely stopped as the [Ca]NMDA amplitude falls below the increased LTP threshold.

Alternative models of calcium accumulation.

(A) Spine [Ca]NMDA without axial diffusion. [Ca]NMDA is described with a pool model similar to that in Eq. 4 (except that instead of [Ca]L-type, this pool accumulates [Ca]NMDA arising from the NMDA current in Eq. 6). (B) Spine [Ca]NMDA + [Ca]VGCC without axial diffusion. Calcium from the NMDA current and all voltage-gated calcium channels (except L-type) accumulates in a pool model. (C) Spine [Ca]NMDA with axial diffusion. Compared to Fig. 2C3, in this plot only calcium from NMDARs diffuses in the neuron, without calcium from other voltage-gated calcium channels. Figure 9—figure supplement 1. Axial diffusion of calcium from NMDARs only.

Because of this, we sought to obtain a monotonic increase of calcium amplitude with increasing plateau voltage (as in Fig. 2C3), for which it was necessary to implement axial diffusion and buffering for [Ca]NMDA. Because tuning a detailed calcium diffusion model to data is a project in its own right, we used the most detailed calcium diffusion model for SPNs available to date (by Dorman et al. (2018)) as a reference. We reuse the same buffer molecules, their amounts and their calcium binding and unbinding rates from Dorman et al. (2018) with small changes, given in Table 2. The rationale for reusing them is that the idealized morphology used in Dorman et al. (2018) should on average represent any other SPN, so the parameters obtained from the optimization procedure in that study should on average also apply to other multicompartment SPN models. The model in Dorman et al. (2018) also includes a plasma-membrane Ca pump (PMCA) in the soma, dendrites and spines, which we added in this model, and a sodium/calcium exchanger in the spines, which we ommited, because with the parameters from Dorman et al. (2018), it did not noticeably alter the spine [Ca]NMDA concentration. In the calcium diffusion model in Dorman et al. (2018), calcium originating from all sources, which in that study are NMDARs and voltage–gated calcium channels (VGCCs), and a minor, 0.1% contribution of the AMPAR current, diffuses throughout the SPN model. Because the calcium buffer parameters are optimized to such a scenario, we also included calcium from the Cav2.1 (Q-type), Cav2.2 (N-type), Cav2.3 (R-type), Cav3.2 and Cav3.3 (T-type) channels in the SPN together with [Ca]NMDA in the diffusion model. Indeed, non-selective blockade of voltage-gated calcium channels with mibefradil has shown that they are also necessary for inducing LTP in SPNs (Fino et al., 2010). Mibefradil has been shown to block T-type, L-type, R-type, N-type and P/Q-type calcium channels in different experimental preparations, so, apart from the L-type channels, we included all of them in the diffusion model (Randall and Tsien, 1997; Viana et al., 1997; Aczél et al., 1998; Eller et al., 2000; Leuranguer et al., 2001). What the role of each of these channels is in LTP in SPNs requires further study with selective blockers.

We also show how diffusion of only [Ca]NMDA looks like in the SPN in Figure 9–figure supplement 1 (and Fig. 9C). The difference with the diffusion model used throughout the study, which includes VGCCs, is minor (Figure 9–figure supplement 1, cf. Figs. 2C, D). The qualitative and quantitative shapes of the voltage and calcium elevations are very similar (Figure 9–figure supplement 1A, cf. Fig. 2C). Regarding the amplitudes of these quantities, there is a 1 mV increase in the evoked somatic amplitude after glutamate spillover when only [Ca]NMDA diffuses (Figure 9–figure supplement 1B1), and an obvious difference in the amplitude of [Ca]NMDA (Figure 9–figure supplement 1B3). Nevertheless, the threshold nonlinearity in the [Ca]NMDA amplitude is still present, and the learning rule would work the same provided the upper plasticity threshold ΘLTP is placed in the region of the nonlinearity.

Even though mibefradil also affects L-type channels, we did not include them in the axial diffusion model together with [Ca]NMDA because L-type calcium channels frequently form calcium microdomains, where calcium entry and diffusion is restricted within a very small, micrometer range (Berridge, 2006; Chen and Sabatini, 2012; Parekh, 2008). In the microdomain, the L-type channels can be physically coupled in supramolecular protein complexes, further contributing to the local effect of the calcium influx in activating the relevant signaling networks, and this is also the case for Cav1.3 channels in SPNs (Olson et al., 2005; Stanika et al., 2015). Because of this, we assume that such highly localized [Ca]L-type signals trigger LTD in the synapse, and calculated [Ca]L-type with a separate, phenomenological pool model, which describes [Ca]L-type in a thin cylindrical shell under the membrane. The implementation follows that of Wolf et al. (2005), according to Eq. 4. In this model [Ca]L-type decays with a time constant τCa, modeling Ca2+ diffusion only phenomenologically, i.e no Ca2+ ions move between dendritic compartments. Compared to the very simplest pool models, this pool model also includes a Ca2+ pump that extrudes intracellular Ca2+ to the extracellular space. F is the Faraday constant, kt is the catalytic activity of the pump, Kd is the pump dissociation constant for Ca2+, d is the depth of the submembrane shell where Ca2+ accumulates, and k and p are phenomenological parameters used in Wolf et al. (2005) used to balance Ca2+ influx and efflux. The values for the parameters in Eq. 4 are given in Table 3.

In this pool model we included calcium from both Cav1.2 and Cav1.3 channels, since there are no pharmacological agents that selectively block only one of the channels, i.e. the compounds used in the experiments which have determined Cav1.3 as necessary for LTD in cortico-striatal synapses would also block Cav1.2 channels in SPNs (Berger and Bartsch (2014); nimodipine in Wang et al. (2006) and Shindou et al. (2011); also, other non-selective L-type channel blockers are used to block LTD in the following studies: nifedipine in Calabresi et al. (1994); nitrendipine in Kreitzer and Malenka (2005); mibefradil in Fino et al. (2010); finally, experiments in Plotkin et al. (2013) have suggested that other VGCCs than Cav1.3 may be involved in activating calcium-induced calcium release, which is necessary for LTD).

Synaptic inputs

There are two different models of synapses used in this study. One model – the dual exponential model – is used in describing the background noise, which consists of glutamatergic and GABAergic synapses, distributed across the dendrites according to the distribution measured in Cheng et al. (1997). These synapses are fixed (they have no plasticity) because they are activated at a low rate of 1 Hz. Compared to the 35 ms activation window in Fig. 3, this rate is low enough so that we decided to neglect any background noise synapses that might be co-active in the same time window and thus become eligible for plasticity together with the feature-carrying synapses. Because there are many synapses representing the background noise, this assumption also lowers the computational cost – if they were plastic, the learning rule would need to be computed for each of them.

The other model, used to describe the synapses carrying the features, is a saturating synapse model. We chose a saturating synapse model because we activate these synapses with 3 spikes each within a 35 ms window, and wanted to avoid any unrealistically high conductance values that could otherwise arise in non-saturating models.

Dual exponential model

The dual exponential synaptic model (a difference of two exponential functions) for a spike arriving at time 0 is represented by: where gmax is the maximal synaptic conductance, and τA and τB are the rise and decay time constants. C is a normalization factor given with: where t⋆ is the position of the maximum of the dual exponential function. This normalization ensures that the time-varying part of the dual exponential reaches a peak value of 1, which is then scaled by the maximal conductance gmax. (The implementation in NEURON contains a synaptic weight parameter, as well. Since in the dual exponential model we have fixed the weight to 1, it is omitted from Eq. 5.)

The synaptic current arising from the conductance gsyn is: Excitatory synapses have both AMPA and NMDA components, and NMDA synapses in addition have a Mg2+ block described by a sigmoidal function: with parameters α = 0.062 and η = 0.38. The parameters of the dual exponential model for the glutamatergic AMPA and NMDA synapses, and the GABAergic synapses are given in Table 4.

Finally, because the SPN model contains less electrical compartments than the number of synapses reported in Cheng et al. (1997), this implies that several background noise synapses will arrive in a single compartment. As such, their synaptic inputs will be integrated in the same voltage variable. Because of this, we have represented such multiple synapses arriving in one compartment with a single synapse instead, whose input frequency is scaled by the number of synapses it represents. This is why a non-saturating synapse model, such as the dual exponential model, was necessary for the background noise synapses. This also greatly increases simulation speed.

Saturating synapse model

The saturating synapse models are taken from Gao et al. (2021), which are a variation of the saturating synapse models in Destexhe et al. (1994). They are kinetic models which operate according to two different kinetic schemes depending on the presence or absence of neurotransmitters. When neurotransmitters arrive at the postsynaptic site, receptor dynamics are described by a kinetic scheme that models switching between closed (C) and open (O) states with rates of α and β: where [T] is the transmitter concentration. After neurotransmitter has been cleared from the synaptic cleft, receptor dynamics evolves according to a kinetic scheme that describes just the closing of the receptor with a rate β: In this model, when a presynaptic spike arrives, neurotransmitter levels always reach a fixed saturating concentration in the synaptic cleft, [T]max, which lasts for a short time Tdur (a short glutamate lifetime in the synaptic cleft of cultured hippocampal synapses has been reported in Clements et al. (1992)). If another presynaptic spike arrives while the neurotransmitter pulse is still “on”, the pulse duration is lengthened by Tdur.

The synaptic conductance gsyn in this model is represented by the fraction of glutamate receptors in the open state, [O]: where gmax is again the maximal synaptic conductance, and w is the synaptic weight which is modified by the learning rule. The parameters in the saturating synapse model are given in Table 5.

Glutamate spillover

To produce plateau potentials which have an all-or-none jump in the voltage, i.e. a threshold non-linearity (as suggested by existing in vitro experiments), we included glutamate spillover. A detailed study of glutamate spillover and its role in triggering plateau potentials is given in Trpevski et al. (2023), and here we briefly describe the model. Glutamate spillover is a phenomenon that occurs during repeated synaptic stimulation, when glutamate reuptake by the surrounding astrocytes is saturated and when the astrocytes may even reverse function from reuptake to glutamate releas (Malarkey and Parpura, 2008). When this happens in a synaptic cluster, glutamate spills over from the synaptic cleft into the extrasynaptic space, activating extrasynaptic NMDARs (eNMDARs). To model glutamate spillover, we place eNMDARs in the dendritic shaft, at the base of each spine in a cluster. Glutamate spillover is activated when a threshold level of clustered synaptic stimulation is reached – the “glutamate threshold”. In this study, the glutamate threshold is reached when 20 synapses with a weight w > 0.5 are activated (with 3 spikes each). Note that this threshold is equivalent to activating 10 synapses with a weight w > 1.0, or 40 synapses with a weight w > 0.25, i. e. the summed weight of all activated clustered synapses should be greater than 10.

When glutamate spills over, i) an activating spike to the eNMDARs is delivered 1.5 ms after the glutamate threshold is reached, ii) it stays much longer around the eNMDARs than in the synaptic cleft (Tdur is much longer), and iii) its concentration around the eNMDARs is much lower than in the synaptic cleft ([T]max is much lower) (Table 5 and Trpevski et al. (2023)). Also, the eNMDARs represent about 64% of the total NMDA conductance in this study (Table 5).

Stimulation protocol

The stimulation protocol in Fig. 3A for presenting the patterns (in random order), which are followed by dopamine feedback, contains several additional details that we list here. First of all, when the distributed or clustered synapses are activated within the 35 ms stimulation window, 3 input spikes arrive to each synapse within this window (as explained in the caption of Fig. 3). This is because at physiological extracellular calcium concentrations, single spikes do not induce plasticity (at least in the spike-timing-dependent protocols in Inglebert et al. (2020)). Furthermore, with the backward rate constants from Dorman et al. (2018) for calcium unbinding from the buffers are much lower than experimentally measured values for calmodulin and calbindin (Faas et al., 2011; Nägerl et al., 2000). To avoid long simulation times on the computing cluster used for this study, 600 ms after a pattern is presented, we reset the values of the buffers, their bound forms to calcium, and the intra- and extracellular calcium concentrations back to their initial values before the first pattern is presented, thus “equilibrating” the system faster and shortening simulation time. As can be seen from Fig. 2C, by 600 ms the plateaus are over, and both [Ca]NMDA and [Ca]L-type are back to their baseline levels, so this abrupt reset of the system does not affect learning and synaptic plasticity. And as can be seen from Fig. 3C, it neither affects the baseline voltage in the soma and dendrites before the arrival of a future pattern.

Performance scores on the FBP and NFBP

In the FBP, a score of 50% means that the SPN is correct only half of the time. This happens if the SPN is silent for both the relevant and irrelevant pattern (correct output only for the irrelevant pattern), or if it spikes for both patterns (correct output only for the relevant pattern). A score of 100% indicates it always spikes for the relevant pattern, and is always silent for the irrelevant pattern, and the FBP is perfectly solved. The threshold score for solving the FBP is 75%, exemplified with the situation where the SPN is always silent for the irrelevant pattern, and spikes at least half of the time for the relevant pattern. A score of 0% would indicate that the SPN’s output is incorrect all the time, being silent for the relevant pattern, and spiking for the irrelevant pattern (this does not happen in the simulations).

In the NFBP, a score of 50% also means that the SPN output is correct only half of the time. An example of such a score is when the SPN is silent for all four patterns, or spikes for all four patterns. Similarly, a score of 100% indicates it always spikes for the relevant patterns, and is always silent for the irrelevant patterns. A score of 75% is exemplified with the situation where the SPN is silent for the two irrelevant patterns, and spikes for only one of the relevant patterns. The threshold score for solving the NFBP is 87.5%, where the SPN would also spike for the other relevant pattern at least half of the time.

eLife Assessment

This computational study constitutes an extension to prior work on biophysical calcium-based synaptic plasticity rules with metaplasticity, investigating how single neurons can learn to perform non-linear pattern classification. This important work presents a significantly simpler solution to the studied problem with potentially broad applicability, there is however incomplete evidence to support the core conclusions.

Reviewer #1 (Public review):

Summary:

This computational modelling study addresses the important question of how neurons can learn non-linear functions using biologically realistic plasticity mechanisms. The study extends the previous related work on metaplasticity by Khodadadi et al. (2025), using the same detailed biophysical model and basic study design, while significantly simplifying the synaptic plasticity rule by removing non-linearities, reducing the number of free parameters, and limiting plasticity to only excitatory synapses. The rule itself is supervised by the presence or absence of a binary dopamine reward signal, and gated by separate calcium-sensitive thresholds for potentiation and depression. The author shows that, when paired with a strong form of dendritic non-linearity called a "plateau potential" and appropriate pre-existing dendritic clustering of features, this simpler learning mechanism can solve a non-linear classification task similar to the classic XOR logic operator, with equal or better performance than the previous publication. The primary claims of this publication are that metaplasticity is required for learning non-linear feature classification, and that simultaneous dynamics in two separate thresholds (for potentiation and depression) are critical in this process. By systematically studying the properties of a biophysically plausible supervised learning rule, this paper adds interesting insights into the mechanics of learning complex computations in single neurons.

Strengths:

The simplified form of the learning rule makes it easier to understand and study than previous metaplasticity rules, and makes the conclusions more generalizable, while preserving biological realism. Since similar biophysical mechanisms and dynamics exist in many different cell types across the whole brain, the proposed rule could easily be integrated into a wide range of computational models specializing in brain regions beyond the striatum (which is the focus of this study), making it of broad interest to computational neuroscientists. The general approach of systematically fixing or modifying each variable while observing the effects and interactions with other variables is sound and brings great clarity to understanding the dynamic properties and mechanics of the proposed learning rule.

Weaknesses:

General notes

(1) The credibility of the main claims is mainly limited by the very narrow range of model parameters that was explored, including several seemingly arbitrary choices that were not adequately justified or explored.

(2) The choice to use a morphologically detailed biophysical model, rather than a simpler multi-compartment model, adds a great deal of complexity that further increases uncertainty as to whether the conclusions can generalize beyond the specific choices of model and morphology studied in this paper.

(3) The requirement for pre-existing synaptic clustering, while not implausible, greatly limits the flexibility of this rule to solve non-linear problems more generally.

(4) In order to claim that two thresholds are truly necessary, the author would have to show that other well-known rules with a single threshold (e.g., BCM) cannot solve this problem. No such direct head-to-head comparisons are made, raising the question of whether the same task could be achieved without having two separate plasticity thresholds.

Specific notes

(1) Regarding the limited hyperparameter search:

(a) On page 5, the author introduces the upper LTP threshold Theta_LTP. It is not clear why this upper threshold is necessary when the weights are already bounded by w_max. Since w_max is just another hyperparameter, why not set it to a lower value if the goal is to avoid excessively strong synapses? The values of w_max and Theta_LTP appear to have been chosen arbitrarily, but this question could be resolved by doing a proper hyperparameter search over w_max in the absence of an upper Theta_LTP.

(b) The author does not explore the effect of having separate learning rates for theta_LTP and theta_LTD, which could also improve learning performance in the NFBP. A more comprehensive exploration of these parameters would make the inclusion of theta_max (and the specific value chosen) a lot less arbitrary.

(c) Figure 4 Supplements 3-4: The author shows results for a hyperparameter search of the learning rule parameters, which is important to see. However, the parameter search is very limited: only 3 parameter values were tried, and there is no explanation or rationale for choosing these specific parameters. In particular, the metaplasticity learning rates do not even span one order of magnitude. If the author wants to claim that the learning rule is insensitive to this parameter, it should be explored over a much broader range of values (e.g., something like the range [0.1-10]).

(2) Regarding the similarity to BCM, the author would ideally directly implement the BCM learning rule in their model, but at the least the author could have shown whether a slight variant of their rule presented here can be effective: for example having a single (plastic, not fixed) Ca-dependent threshold that applies to both LTP and LTD, with a single learning rate parameter.

(3) This paper is extremely similar (and essentially an extension) to the work of Khodadadi et al. (2025). Yet this paper is not mentioned at all in the introduction, and the relation between these papers is not made clear until the discussion, leaving me initially puzzled as to what problems this paper addresses that have not already been extensively solved. The introduction could be reworked to make this connection clearer while pointing out the main differences in approach (e.g., the important distinction between "boosting" nonlinearities and plateau potentials).

(4) The introduction is missing some citations of other recent work that has addressed single-neuron non-linear computation and learning, such as Gidon et al (2020); Jones & Kording (2021).

(5) Figure 1: The figure prominently features mGluR next to the CaV channel, but there is no mention of mGluR in the introduction. The introduction should be updated to include this.

(6) Could the author explain why there is a non-monotonic increase/decrease in the [Ca]_L in Figure 2B_4? Perhaps my confusion comes from not understanding what a single line represents. Does each line represent the [Ca] in a single spine (and if so, which spine), or is each line an average of all the spines in a given stim condition?

(7) Row 124 (page 4): L-type Ca microdomains (in which ions don't diffuse and therefore don't interact with Ca_NMDA) is a critical assumption of this model. The references for this appear only in the discussion, so when reading this paper, I found myself a bit confused about why the same ion is treated as two completely independent variables with separate dynamics. Highlighting the assumption (with citations) a bit more clearly in the results section when describing the rule would help with understanding.

(8) Row 149 (page 5): The current formulation of the update rule is not actually multiplicative. The fact that the update is weight-dependent alone does not make it a multiplicative rule, and judging by equation (1) it appears to simply be an additive rule with a weight regularization term that guarantees weight bounds. For example, a similar weight-dependent update is also a core component of BTSP (Milstein et al. 2021; Galloni et al. 2025), which is another well-known *additive* rule. An actual multiplicative rule implies that the update itself is applied via a multiplication, i.e. w_new = w_old * delta_w

For an example of a genuinely multiplicative rule, see: Cornford et al. 2024, "Brain-like learning with exponentiated gradients"). Multiplicative rules have very different properties to additive rules, since larger weights tend to grow quickly while small weights shrink towards 0.

(9) Equation 1 (page 5): Shouldn't the depression term be written as: (w_min - w)? This term would be negative if w is larger than w_min, leading to LTD. As it is written now, a large w and small w_min would just cause further potentiation instead of depression.

(10) In the introduction, the teaching signal is described in binary terms (DA peak, or DA pause), but in Equation 1, it actually appears to take on 3 different values. Could the author clarify what the difference is between a "DA pause" and the "no DA" condition? The way I read it, pause = absence of DA = no DA

(11) Figure 3: In these experimental simulations, DA feedback comes in 400ms after the stimulus. The author could motivate this choice a bit better and explain the significance of this delay. Clearly, the equations have a delta_t term, but as far as the learning algorithm is concerned, it seems like learning would be more effective at delta_t=0. Is the choice of 400ms mainly motivated by experimental observations? On a related note, is it meaningful that the 200ms delta_t before the next stimulus is shorter than the 400ms pause from the first stimulus? Wouldn't the DA that arrives shortly before a stimulus also have an effect on the learning rule?

(12) Figure 4C: How is it possible that the theta_LTP value goes higher than the upper threshold (dashed line)? Equation 3 implies that it should always be lower.

(13) Row 429 (page 11): The statement that "without metaplasticity the NFBP cannot be solved" is overly general and not supported by the evidence presented. There exist many papers in which people solve similar non-linear feature learning problems with Hebbian or other bio-plausible rules that don't have metaplasticity. A more accurate statement that can be made here is that the specific rule presented in this paper requires metaplasticity.

(14) The methods section does not make any mention of publicly available code or a GitHub repository. The author should add a link to the code and put some effort into improving the documentation so that others can more easily assess the code and reproduce the simulations.

Reviewer #2 (Public review):

Summary:

The manuscript proposes interesting synaptic plasticity rules grounded in experimental data. Its main features are:

(1) plasticity depends on local calcium concentration driven by presynaptic activity and is independent of somatic action potentials,

(2) the rules incorporate metaplasticity, and (3) they demonstrate how a single neuron could address the feature-binding problem at the dendritic level.

The work extends a previous study (https://doi.org/10.7554/eLife.97274.2), to which the author also contributed.

The author models two calcium thresholds (LTP/LTD) from two different calcium sources (NMDA/VGCC), and these thresholds are flexible (metaplasticity rule, similar to BCM), which is claimed to be necessary for successful learning of both FBP and NFBP (linear and nonlinear feature binding problem with 1 or 2 patterns). The role of each threshold seems to be opposite and complementary. One extra condition has been added: an upper threshold for LTP. This threshold serves to stop synaptic strengthening once synapses are strong enough to evoke a plateau. With that, synapses are not strengthened to the maximal value, avoiding strong supralinear integration for irrelevant patterns.

Strengths:

The current model implements not only local synaptic plasticity but also metaplasticity and solves the FBP at the dendrite level. Another strong aspect of the model is that metaplasticity in the LTD threshold protects strengthened synapses from weakening. In this way, as the author mentioned, metaplasticity is able to protect learned patterns from being forgotten or weakened and prevent irrelevant patterns from being stored. This is a nice modelling example of metaplasticity being helpful in preventing the catastrophic interference or forgetting (as has been explicitly discussed in a recent article https://doi.org/10.1016/j.tins.2022.06.002 ). The author might want to briefly mention or emphasize this aspect of the model, which might be interesting also for the AI community.

Weaknesses:

(1) What is novel in the current paper as compared to Khodadadi et al. eLife 2025? That is not completely clear and should be made clearer. Is it only a minor difference related to the fact that the new learning rule has metaplasticity in both calcium thresholds and is simpler? This seems to be just an incremental increase in knowledge/methods. Can the author defend his paper against this point from the „devil's advocate"? How is the conclusion of the author in the abstract that „metaplasticity in both thresholds is necessary" reconcilable with his previous publication (Khodadadi et al. eLife 2025), in which only metaplasticity in one threshold was successful in solving the nonlinear feature binding problem?

(2) As far as I can judge without testing the model, metaplasticity causes thresholds to monotonically increase during systematic pattern presentation, which stabilizes weights and allows pattern separation. Due to the closed-loop nature of the current implementation, where metaplasticity only happens if plasticity happens, this also effectively locks patterns in place. However, flexible learning is an essential mechanism for survival. Imagine a mutation event takes place and bananas suddenly become red and/or strawberries turn yellow. It seems that the current model would be unable to adapt to these new patterns even if rewards were to be shifted. While out of the scope of the study, due to its importance, I feel that pattern shifting/relearning should at least be briefly discussed. How could the model be improved to allow relearning?

Author response:

The following is the authors’ response to the original reviews.

The goal of the study was to demonstrate the role of the two thresholds in learning, i.e. the role of metaplasticity with a learning rule that is both as simple as possible, but also biologically-based (hence the use of the detailed SPN model, as well). The goal was not to solve the FBP and NFBP with a different rule. However, the text likely gives that impression. Therefore, in addition to addressing the reviewers’ comments, the text has been substantially revised to reflect the following:

Main results

(1) Metaplasticity enables synapses that undergo both LTP and LTD to ultimately express just one plasticity outcome (either LTP or LTD).

(2) Metaplasticity in the LTP threshold allows LTD to be expressed, and vice versa. (That is, the threshold regulating one plasticity process allows the expression of the opposite plasticity process.)

Reason for using the FBP and NFBP

The reason why the FBP and NFBP are particularly useful to demonstrate the thresholds’ roles is that the patterns in the tasks share features. (For example, in the FBP, the feature ‘strawberry’ is shared between both ‘red strawberry’ and ‘yellow strawberry’.) The synapses for these shared features experience precisely such competing LTP and LTD processes where the role of metaplasticity becomes evident. Specifically, the role of the LTD threshold is visible in shared synapses that need to be strengthened during learning (such as the shared ‘strawberry’ synapses in the FBP), while the LTP threshold’s role is visible in shared synapses that need to be weakened (there are no such synapses in the FBP, but there are in the NFBP, which is why it is needed in this article). In addition, feature binding is relevant for the striatum (now described in the introduction).

In the first version of the article there are sentences that are likely confusing and misleading regarding the goal of the study, and these have been removed (lines 98-100 in the tracked changes file).

The revised article now contains 4 new main figures and 15 new figure supplements prompted by the reviewers’ comments, thus expanding the scope of the study and hopefully clarifying its content, the mechanisms, and the conclusions.

Finally, the question whether this rule can solve the NFBP with excitatory synapses alone is being addressed in a new, ongoing study, as written at the end of the Discussion section in the first version of this article. The new study tests 100+ SPN models (71 dSPN and 34 iSPN models differing in ion channel composition, with two different morphologies), different locations of the synaptic clusters along the dendrites (from proximal to distal), and two different learning regimes (suprathreshold – where the neuron initially spikes for all patterns, and subthreshold – where the neuron is initially silent for all patterns, as in this study) containing both clustered and distributed synapses in the same setup. Where necessary, preliminary results from this study are attached in the responses.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This computational modelling study addresses the important question of how neurons can learn non-linear functions using biologically realistic plasticity mechanisms. The study extends the previous related work on metaplasticity by Khodadadi et al. (2025), using the same detailed biophysical model and basic study design, while significantly simplifying the synaptic plasticity rule by removing non-linearities, reducing the number of free parameters, and limiting plasticity to only excitatory synapses. The rule itself is supervised by the presence or absence of a binary dopamine reward signal, and gated by separate calcium-sensitive thresholds for potentiation and depression. The author shows that, when paired with a strong form of dendritic non-linearity called a "plateau potential" and appropriate pre-existing dendritic clustering of features, this simpler learning mechanism can solve a non-linear classification task similar to the classic XOR logic operator, with equal or better performance than the previous publication. The primary claims of this publication are that metaplasticity is required for learning non-linear feature classification, and that simultaneous dynamics in two separate thresholds (for potentiation and depression) are critical in this process. By systematically studying the properties of a biophysically plausible supervised learning rule, this paper adds interesting insights into the mechanics of learning complex computations in single neurons.

As mentioned in the introductory response above, the goal of the article was not to solve the NFBP, but to study the role of metaplasticity. The text has now been substantially revised to reflect that. One of the primary claims that “metaplasticity is required for learning non-linear feature classification, and that simultaneous dynamics in two separate thresholds (for potentiation and depression) are critical in this process” is now removed from the abstract. Instead, the abstract now focuses on the main results regarding the role of metaplasticity, as well as the results which arose from addressing the reviewers’ comments. However, I guess I should note that to be able to study the role of metaplasticity using the FBP and the NFBP, it was first necessary to have a rule that solves these tasks – that allows to later observe the effects of systematically fixing or modifying parameters in the rule. Another aspect which may have added to the confusion about the study’s goal may be due to the NFBP being a complex task requiring supralinear dendritic integration. Because of this, plateaus have received significant attention in the article, possibly diverting from the main focus. This is also now clarified in the article.

Strengths:

The simplified form of the learning rule makes it easier to understand and study than previous metaplasticity rules, and makes the conclusions more generalizable, while preserving biological realism. Since similar biophysical mechanisms and dynamics exist in many different cell types across the whole brain, the proposed rule could easily be integrated into a wide range of computational models specializing in brain regions beyond the striatum (which is the focus of this study), making it of broad interest to computational neuroscientists. The general approach of systematically fixing or modifying each variable while observing the effects and interactions with other variables is sound and brings great clarity to understanding the dynamic properties and mechanics of the proposed learning rule.

It would indeed be great if this rule facilitates similar studies in other neuron types. The rule requires at least qualitatively accurate calcium dynamics, which at the moment are not widely available for other neuron types (but will likely appear in the future).

Weaknesses:

General notes

(1) The credibility of the main claims is mainly limited by the very narrow range of model parameters that was explored, including several seemingly arbitrary choices that were not adequately justified or explored.

The parameter range has now been expanded to address all comments below.

(2) The choice to use a morphologically detailed biophysical model, rather than a simpler multicompartment model, adds a great deal of complexity that further increases uncertainty as to whether the conclusions can generalize beyond the specific choices of model and morphology studied in this paper.

Regarding the plasticity rule and the role of metaplasticity, the conclusions should hold if the calcium dynamics of other (simpler or detailed) neuron models is qualitatively similar (i.e. shows the same monotonicity of “stronger synaptic input – stronger calcium”, and not something like Figs. 12A, B for which the rule was not tested).

As to whether the NFBP can be solved, that is being addressed in the new study with 100+ neuron models (71 dSPN models and 34 iSPN models, and two morphologies: one for the dSPNs and another for the iSPNs). So far, the results show that the important ingredient is the threshold nonlinearity provided by the plateau potential (both dSPNs and iSPNs exhibit this nonlinearity, irrespectively of where the synaptic cluster is located, see Author response image 1).

Also, plateaus can be evoked in passive SPN dendrites (as stated in the methods in lines 916-919, by “turning off” ion channels in the computational model as in Fig. S7 of Du et al. (2017) (see list of references provided)). Although plateaus are somewhat modified by active conductances, the main behavior important for the NFBP is given by NMDARs. The most striking contributions of the other ion channels in SPNs are the low resting potential around -85 mV, delayed first spike under current injection and strong inward rectification. Basically, SPNs require more input to be stimulated compared to other neurons that lack these properties. So, other neurons with NMDAR-dependent nonlinearities should in principle be able to solve the same task, but perhaps needing less inputs (provided dendritic integration is sufficiently supralinear – see response to comment (2) under Reviewer #2 (Recommendations for the authors)).

(3) The requirement for pre-existing synaptic clustering, while not implausible, greatly limits the flexibility of this rule to solve non-linear problems more generally.

In my view, the limitation of this rule is that there is no mechanism for structural plasticity with which clusters could be “grown”. For example, Hedrick et al. 2022 demonstrate that during learning in the motor cortex, existing functional clusters are “enlarged” by growing new spines. Devising a biologically realistic rule for structural plasticity is a separate research project, which would be very interesting to do.

In any case, to solve the NFBP with excitatory synapses alone, supralinear voltage elevations are necesssary (Tran-van-Minh et al. 2015). When it comes to NMDAR-dependent supralinear integration, clustering is basically necessary, within a smaller, e.g. 20 micron dendritic stretch, or more loosely, within a single dendritic branch (Losonczy and Magee, 2006; Branco et al. 2010). Also, synaptic clustering is quite present across different brain regions (such as the sensory and motor cortices, pallium, striatum, hippocampus, brainstem, the developing cortex, and across species, e. g. see the introduction of Kirchner and Gjorgjieva, 2021), so I suppose it is not that implausible. Inputs from motor cortex to striatum are also clustered (Hwang et al. 2022).

(4) In order to claim that two thresholds are truly necessary, the author would have to show that other well-known rules with a single threshold (e.g., BCM) cannot solve this problem. No such direct head-to-head comparisons are made, raising the question of whether the same task could be achieved without having two separate plasticity thresholds.

Although the goal of the study was not to solve the NFBP, this is a very interesting question related to the question “what is the use of having two separate thresholds for plasticity”? I have added simulations in which the rule is modified to use a single calcium threshold instead, showing that the FBP, NFBP, and reversal learning are solved just as well (Fig. 12 and Figure 12 – figure supplements 1 – 4). The results show that two calcium thresholds allow for separate control over the LTP and LTD processes (lines 687-706).

Specific notes

(1) Regarding the limited hyperparameter search:

(a) On page 5, the author introduces the upper LTP threshold Theta_LTP. It is not clear why this upper threshold is necessary when the weights are already bounded by w_max. Since w_max is just another hyperparameter, why not set it to a lower value if the goal is to avoid excessively strong synapses? The values of w_max and Theta_LTP appear to have been chosen arbitrarily, but this question could be resolved by doing a proper hyperparameter search over w_max in the absence of an upper Theta_LTP.

Thank you for raising this question – it is closely related to the very important question of when to stop updating the weights. Before providing the response, I should clear up any confusion made due to an error in notation. In this study, wmax=2w―0, and wmax=2w―0, meaning that during learning the weights can increase up to twice (up to 100% of) the initial value, and can decrease down to 1% of the initial value. Currently, it says wmax = 2, and wmin = 0.01, which is a mistake, and possibly a reason why the chosen values seem arbitrary.

The reason for introducing the upper threshold θLTP is most easily seen in the parameter scan for wmax done in the absence of θLTP (Figure 4 – figure supplement 4) – it is very difficult to choose a value for wmax with which all three tasks will be solved (FBP with linear integration, i.e. distributed synapses, the FBP with supralinear integration, i.e. clustered synapses, and the NFBP). Put differently, if there is a range of wmax values which solves all three tasks, it is very narrow. So, θLTP is introduced to solve all three cases without specially tuning wmax.

The following is a description of the results in Figure 4 – figure supplement 4, which is also present in the article in lines 383-396. The reason why a single preset value for wmax does not work is that distributed synapses sum only linearly at the soma, while clustered synapses sum supralinearly. This means that to trigger somatic spiking, the distributed synapses need to be strengthened more compared to clustered synapses. Concretely, this means that a wmax=1.5w―0 is needed to solve the FBP with linear integration (distributed synapses), but it is already too strong to solve the NFBP (Figure 4 – figure supplement 4, column four: top panel shows that FBP with linear integration is solved, while the bottom panel shows the average performance for the NFBP is right at the border – pink line aligns with the dashed line). On the other hand, wmax=1.5w―0 is enough in the FBP with clustered synapses and the NFBP to trigger glutamate spillover and a plateau potential, which drives somatic spiking (Figure 4 – figure supplement 4, column two). However, it is not enough to solve the FBP with linear integration. Larger values for wmax increase the performance on the FBP with linear integration, but lower the performance on the NFBP due to spiking for the irrelevant patterns (especially visible once wmax=1.5w―0). As said, in this parameter scan, only the value max exists which solves all three tasks, it is very narrow.wmax=1.5w―0 solves all three tasks, with borderline performance on the NFBP (it is arguable whether the borderline performance means the NFBP is solved). This suggests that if a range of values for w

With the settings used in the study, the w max threshold for triggering glutamate spillover is a number between 1.2 and 1.3 with many decimals – 1.230769231. A possibility could be to program spillover to always occur at wmax=1.5w―0, but this is also not a general solution. For example, other dSPN models from the model library in Lindroos et al. (2021) have different excitability, and that would mean a different, specific value of wmax would be needed for each neuron model.

θLTP is thus a solution that prevents “over-excitation” in a dendrite. The logic is that synapses in a cluster do not need to be as strong as distributed synapses to be effective in exciting the neuron. θLTP ensures that once plateaus appear, synapses are not strengthened further. This allows using the same value wmax=2w―0 across all scenarios, i.e. avoiding tuning of wmax or tuning the number of synapses representing the features. (Also, preliminary results show this works well in the new study which tests all dSPN and iSPN models in the model library.) As for the value of θLTP, it can be set anywhere in the “blank” region in Fig. 2C3 where the supralinear jump in [Ca]NMDA occurs. What is important is that [Ca]NMDA evoked by a plateau should be above θLTP – then no more strengthening will occur once plateaus can be evoked.

(b) The author does not explore the effect of having separate learning rates for theta_LTP and theta_LTD, which could also improve learning performance in the NFBP. A more comprehensive exploration of these parameters would make the inclusion of theta_max (and the specific value chosen) a lot less arbitrary.

I am not sure I have understood this comment properly, but hopefully this response addresses it adequately. In fact, the computer code does use separate variables for the learning rates for θLTP and θLTP, but for simplicity, they are set to the same value. In this response I have included some simulations in which they are set to different values (Author response image 2), but to understand their effect it is better to read the response to the comment under c) first. The metaplasticity rates determine how fast the synaptic weights are stabilized, i.e. how long the time window is where the synapses are “flexible”. When the thresholds reach the calcium amplitudes, the weights are “locked”.

If the metaplasticity rates are too low, the weights take a longer time to stabilize and solving the tasks takes a long time (Figure 4 – figure supplement 10 from the response to c). If they are too high, weights are quickly stabilized and, in the case of weakened synapses, prematurely so, leading to worse performance on the NFBP due to spiking for the irrelevant patterns. (Figure 4 – figure supplement 11).

One can use different values for the metaplasticity rates, as in the examples in Fig. 2 , and this affects the speed of learning. A large ηθLTD = 20 quickly prevents the LTD process (Figs. 2D1, D2), causing faster synaptic strengthening, but also quickly limiting weakening (Figs. 2B1, B2) – as a result, after learning, the neuron spiked for an irrelevant pattern (Fig. 2A1). Conversely, a large ηθLTD = 20 quickly prevents the LTP process (Figs. 2C3, C4), prolonging synaptic strengthening and allowing faster weakening (Figs. 2B3, B4) – as a result, the NFBP is solved, but takes a longer time. However, this does not remove the need for using an upper threshold θLTP. If I have not understood the comment properly, please let me know so I can address it further.

(c) Figure 4 Supplements 3-4: The author shows results for a hyperparameter search of the learning rule parameters, which is important to see. However, the parameter search is very limited: only 3 parameter values were tried, and there is no explanation or rationale for choosing these specific parameters. In particular, the metaplasticity learning rates do not even span one order of magnitude. If the author wants to claim that the learning rule is insensitive to this parameter, it should be explored over a much broader range of values (e.g., something like the range [0.1-10]).

Thank you for bringing this up! I have expanded the parameter scan for both the learning rate η (Figure 4 – figure supplements 7 and 8) and the metaplasticity rate ηθ (Figure 4 – figure supplements 10 and 11) to cover three orders of magnitude (when taken together with the results already present in the article in Figure 4 – figure supplements 6 and 9).

Figure 4 – figure supplements 7 and 8: The learning rate η was was tested with the values 0.1 and 20. This shows a known result in other plasticity rules, which is that if the learning rate is too fast or too slow, learning is not successful or is too slow. For η = 0.1, learning takes very long (around 2000 patterns for FBP with linear integration), and for the NFBP is at the borderline score (Figure 4 – figure supplement 7E). As the LTD thresholds rise, they will prevent the slowly decreasing synaptic weights from decreasing sufficiently. For η = 20, the weight updates are too large and no pattern is stored because calcium quickly falls below the thresholds as a result of such large weight updates (Figure 4 – figure supplement 8E). This leaves some intermediate range of learning rates that “works”. (The three values {0.4, 0.85, 1.7} that were tested in Figure 4 – figure supplement 6 are roughly a factor of 2 apart.) Results are described in more detail in lines 409-414 of the revised article.

Figure 4 – figure supplements 10 and 11: The metaplasticity rate ηθ was tested with the values 0.2 and 20. This further demonstrates the role of metaplasticity as a “lock” on the learning process. ηθ determines how fast the weights stabilize (i.e. how long they remain flexible for learning). With the value ηθ = 0.2 the weights take a longer time to stabilize, because it takes a long time for the thresholds to reach the calcium amplitudes (Figure 4 – figure supplement 10). This affects the learning time on the NFBP only. In contrast, with the value ηθ = 20, the thresholds quickly reach the calcium amplitudes, and quickly stabilize the weights. While the neuron does learn to evoke plateaus, the weakened synapses are stabilized too early, resulting in spikes for the irrelevant patterns and lowering the performance on the NFBP. (Figure 4 – figure supplement 11). Results are described in more detail in lines 415-429 of the revised article.

(2) Regarding the similarity to BCM, the author would ideally directly implement the BCM learning rule in their model, but at the least the author could have shown whether a slight variant of their rule presented here can be effective: for example having a single (plastic, not fixed) Cadependent threshold that applies to both LTP and LTD, with a single learning rate parameter.

- As mentioned above, this is also a very interesting question, related to the important question of “what is the use of having two separate calcium thresholds?”. I have added simulations according to your second suggestion – using a single calcium threshold that applies to both LTP and LTD. A single threshold can solve the FBP, NFBP, and reversal learning equally well (Fig. 12 and Figure 12

- Figure Supplements 1, 2). This indeed raises the question what are two separate thresholds useful for? Figure 12 – figure supplement 3 shows that when keeping the single calcium threshold fixed, none of the shared synapses stabilize. Comparing this to Figs. 5 and 6, where the LTP and LTD threshold are kept fixed one at a time, it shows that having two separate thresholds allows separate control over the LTP and LTD processes.

I also tried an implementation of the BCM rule using a nonlinear (quadratic) function in the metaplasticity rule for updating the threshold:

where cnorm is a normalization constant set to 80 μM for [Ca]NMDA and 20 μM for [Ca]L-type (close to the maximal concentrations achieved throughout the article). In the BCM rule, a nonlinear function is necessary to prevent runaway growth (or collapse) of the weights. So, in principle, wmax and θLTP should be unnecessary with this quadratic metaplasticity rule. However, even though the weights seem to stabilize (around a high value), they are too strong and plateaus are triggered for all patterns (see Author response image 3). It is possible that only a narrow range of parameters in the quadratic function will allow weights to stabilize at appropriate values, similarly to the narrow range in wmax in the absence of θLTP (as described for comment 1a) above).

Below are some questions about how to “correctly translate” the BCM rule from its formulation for rate-coded inputs to a formulation for sparsely-coded inputs which are used in this article. I did not address these questions in detail. The following is meant for readers who might be interested in pursuing these questions.

Questions when translating the BCM rule for rate-coded inputs to a rule for sparsely-coded inputs:

The similarity of this study’s rule to BCM is only qualitative – a sliding threshold was added that follows an indicator of synaptic activity (in this study it is the amplitude of the calcium concentration). The original BCM rule is defined for rate-based synapses, and its threshold θ is a nonlinear function of the average (synaptic) activity:

where p and c0 are positive constants, and c) is the average activity. The threshold θ is updated as the average activity c) changes (the averaging being done over a time interval) [1]. Changes in the rate-coded inputs drive changes in c) continuously, thus changing θ over time.

(1) What should be chosen as the average activity c̅ ?

In the BCM model, the average activity, and consequently the threshold θ, are global for the whole neuron (which is represented just by its firing rate), meaning that θ is the same for all synapses. In a morphologically realistic neuron model, what should be the average activity? Especially in the case of supralinear integration occuring in a dendrite, one could also consider average activities in dendrites. In the dSPN model, the calcium concentration decays by the time the next pattern arrives, so one probably needs another, low-pass filtered variable of the calcium concentration (or voltage), with a much longer time constant than that of the calcium decay. (I have chosen to slide the thresholds towards the amplitude of the calcium evoked by a pattern, which could be viewed as an approximation of such a long-lived variable. This is one possible “translation” of this aspect of the BCM rule.)

(2) How to update the chosen indicator of activity c̅?

Assuming an adequate indicator is chosen (different from the calcium amplitude that I have chosen), how should it be updated when using sparsely-coded inputs? Should it evolve continuously, or only when the sparsely-coded inputs are active? If updated continuously, it should not decay very fast when synapses are inactive (not activated by a pattern), while still change fast enough to detect changes in synaptic weight occuring due to learning. (I have chosen to slide the thresholds towards the evoked calcium amplitude only when synapses have been activated, so this is also one way of “translating” this second aspect.)

(1) which in the BCM article is replaced with the average over the input’s probability distribution

(3) What nonlinear function to use to update the threshold?

The BCM rule allows any function with p > 1, but how fast the synapses stabilize will likely depend on c0, the normalization constant in the nonlinear function, and the power p. The role of the nonlinear function is to ensure stability of the synaptic weights (prevent runaway growth), meaning no θLTP nor wmax should be needed. However, achieving high performance on all three tasks might still require a specific tuning of these parameters, as wmax does if no θLTP is used. That is, these parameters will need to be chosen such that the threshold quickly goes up once a plateau appears, so weights are not allowed to strengthen further. Having in mind that there is a very narrow range in wmax where all three tasks are solved, there might similarly be only a narrow range in c0 and p.

Since the goal of the article was to study the role of the two thresholds (not to optimize the rule’s parameters on the three tasks), I have only tried a quadratic function for the metaplasticity rule, and did not search for a region of parameters that could work.

(3) This paper is extremely similar (and essentially an extension) to the work of Khodadadi et al. (2025). Yet this paper is not mentioned at all in the introduction, and the relation between these papers is not made clear until the discussion, leaving me initially puzzled as to what problems this paper addresses that have not already been extensively solved. The introduction could be reworked to make this connection clearer while pointing out the main differences in approach (e.g., the important distinction between "boosting" nonlinearities and plateau potentials).

The last paragraph in the introduction now makes this comparison, highlighting the differences in approach, and mainly emphasizing that the goal of the two studies is different. The open question that this study addresses is to pinpoint the roles of the two calcium thresholds (present in dSPNs) in learning. Also see the response to the similar comment (1) from Reviewer #2 (Public review): Weaknesses.

(4) The introduction is missing some citations of other recent work that has addressed single-neuron non-linear computation and learning, such as Gidon et al (2020); Jones & Kording (2021).

Thank you for mentioning this. Since the first version of the article caused confusion regarding the goal of the study (which is to study metaplasticity), I avoided focusing on nonlinear computation in the introduction, and instead added these references to the Discussion (in line 850 the section regarding nonlinear computation).

(5) Figure 1: The figure prominently features mGluR next to the CaV channel, but there is no mention of mGluR in the introduction. The introduction should be updated to include this.

Thank you for noticing this. It is now included in the introduction in lines 50-53.

(6) Could the author explain why there is a non-monotonic increase/decrease in the [Ca]_L in Figure 2B_4? Perhaps my confusion comes from not understanding what a single line represents. Does each line represent the [Ca] in a single spine (and if so, which spine), or is each line an average of all the spines in a given stim condition?

Thank you for noticing this, that information was missing from the figure caption. The line is from a single spine which is placed at a random location on a randomly chosen dendrite. Because each line comes from a different trial with a different number of synapses, the spine it corresponds to is in a different location, resulting in different voltage and different calcium signals when using distributed synapses. (Clustered synapses are placed at approximately the same somatic distance, despite being on a randomly chosen dendrite, so this effect does not appear in Figs. 2C3, C4). The figure caption has been updated accordingly.

(7) Row 124 (page 4): L-type Ca microdomains (in which ions don't diffuse and therefore don't interact with Ca_NMDA) is a critical assumption of this model. The references for this appear only in the discussion, so when reading this paper, I found myself a bit confused about why the same ion is treated as two completely independent variables with separate dynamics. Highlighting the assumption (with citations) a bit more clearly in the results section when describing the rule would help with understanding.

Thank you for pointing this out! This assumption is now clearly stated with citations in lines 174176 when characterizing the plateaus in SPNs, and lines 205-207 when describing the rule (in addition to the existing description in the Methods).

(8) Row 149 (page 5): The current formulation of the update rule is not actually multiplicative. The fact that the update is weight-dependent alone does not make it a multiplicative rule, and judging by equation (1) it appears to simply be an additive rule with a weight regularization term that guarantees weight bounds. For example, a similar weight-dependent update is also a core component of BTSP (Milstein et al. 2021; Galloni et al. 2025), which is another well-known *additive* rule. An actual multiplicative rule implies that the update itself is applied via a multiplication, i.e. w_new = w_old * delta_w

For an example of a genuinely multiplicative rule, see: Cornford et al. 2024, "Brain-like learning with exponentiated gradients"). Multiplicative rules have very different properties to additive rules, since larger weights tend to grow quickly while small weights shrink towards 0.

Thank you for explaining the difference between “additive” and “multiplicative”. I have removed the word “multiplicative” from the description of the rule, as it is not necessary. (In the first version of the manuscript I followed the nomenclature by Gütig et al. (2003), which refers to the weight-dependent update as a multiplicative rule, while an additive rule has no weight dependence in the equation for Δw, which is Eq. 2 in Gütig et al. (2003).)

(9) Equation 1 (page 5): Shouldn't the depression term be written as: (w_min - w)? This term would be negative if w is larger than w_min, leading to LTD. As it is written now, a large w and small w_min would just cause further potentiation instead of depression.

Yes, a minus sign is missing in front of the depression term. Thank you for noticing this! It is now corrected.

(10) In the introduction, the teaching signal is described in binary terms (DA peak, or DA pause), but in Equation 1, it actually appears to take on 3 different values. Could the author clarify what the difference is between a "DA pause" and the "no DA" condition? The way I read it, pause = absence of DA = no DA

Yes, the “no DA” condition should say “baseline DA”. The DA signal does indeed take three different values. In the striatum, there is a baseline tone of dopamine. The DA peaks are transient increases from this tone, and the DA pauses are transient decreases. The basal dopamine level is now introduced in line 113 of the introduction, so that Eq. 1 should be clearer. In the computer code, DA is in effect a ternary signal, with three values, +1 for DA peak, -1 for DA pause, and 0 for baseline DA.

(11) Figure 3: In these experimental simulations, DA feedback comes in 400ms after the stimulus. The author could motivate this choice a bit better and explain the significance of this delay. Clearly, the equations have a delta_t term, but as far as the learning algorithm is concerned, it seems like learning would be more effective at delta_t=0. Is the choice of 400ms mainly motivated by experimental observations? On a related note, is it meaningful that the 200ms delta_t before the next stimulus is shorter than the 400ms pause from the first stimulus? Wouldn't the DA that arrives shortly before a stimulus also have an effect on the learning rule?

These time windows were chosen purely to keep the simulations as short as possible (since a computing cluster was used to run many trials). In reality, two patterns would probably arrive after a longer time window, as an animal should reach for the pattern (and in the case of these fruit patterns, eat it) which would last on the order of seconds. However, the calcium signals that trigger plasticity in these simulations are shorter than that (Figs. 2B3,4, 2C3,4), and only the amplitude of the signal is used to trigger plasticity. This gives an opportinity to not have to wait for several seconds between two patterns arrive (as might occur in reality), but shorten that interval and so shorten the simulation time. (This is now stated in lines 245-247 in the Results section where the rule is described, pointing to the Methods for more detials.)

Hence, the dopamine feedback is given at 400 ms after a pattern appears because the calcium amplitudes would have surely occurred by then (even if a plateau were evoked, see Figs. 2B3,4, 2C3,4). Similarly, the next pattern is given at 600 ms because the calcium signals would have decreased to baseline by then. (This is already explained in the Methods section in lines 1101-1117). The key implicit assumption here is that the calcium amplitude is somehow remembered by the synaptic circuitry. In experiments, dopamine feedback up to 2 seconds after synaptic stimulation can cause LTP, so it is OK to assume this (Yagishita et al., 2014). However, dopamine signaling before synaptic stimulation has no effect (Yagishita et al., 2014). (This assumption is now clearly stated in the Methods in lines 902-906.)

Lastly, in this rule, if the dopamine feedback coincides with the pattern, then whatever calcium levels were reached at that time would be used to check whether the thresholds are crossed, and these will most likely not be the maximal calcium levels (these are reached some time later, as seen in Figs. 2B3,4, 2C3,4).

One last note is that from your comment I got the impression that you may have understood Δt to be the time until dopamine feedback is provided. It is not, it is simply the time step in the simulation (already stated in line 211; now an additional explanation is added in lines 211-214, just in case).

(12) Figure 4C: How is it possible that the theta_LTP value goes higher than the upper threshold (dashed line)? Equation 3 implies that it should always be lower.

This is also just a choice of implementation: θLTP evolves independently of θLTP, reflecting an implicit assumption that different molecular machinery implements θLTP and θLTP. This assumption is similar to the findings in Figs. 4, 5 in Ngezahayo et al. 2000, where in weakened synapses the voltage threshold for LTD can increase (to -20 mV) above the voltage threshold for LTP (at -30 mV). (This assumption is now explicitly written out in lines 316-322.)

(13) Row 429 (page 11): The statement that "without metaplasticity the NFBP cannot be solved" is overly general and not supported by the evidence presented. There exist many papers in which people solve similar non-linear feature learning problems with Hebbian or other bio-plausible rules that don't have metaplasticity. A more accurate statement that can be made here is that the specific rule presented in this paper requires metaplasticity.

Yes, that is what is meant with that statement – that this specific rule requires metaplasticity. The text has been corrected. Thank you for noticing this.

(14) The methods section does not make any mention of publicly available code or a GitHub repository. The author should add a link to the code and put some effort into improving the documentation so that others can more easily assess the code and reproduce the simulations.

The link to the code is given in the Data availability statement. However, this statement might not appear in the manuscript files that you are receiving (it is available online and in the article PDF downloadable from eLife), so I am providing the link here, as well. The Readme.md file now describes the code, and how to reproduce figures in the article.

https://github.com/danieltrpevski/Plasticity/tree/metaplasticity

Reviewer #2 (Public review):

Summary:

The manuscript proposes interesting synaptic plasticity rules grounded in experimental data. Its main features are:

(1) plasticity depends on local calcium concentration driven by presynaptic activity and is independent of somatic action potentials,

(2) the rules incorporate metaplasticity, and

(3) they demonstrate how a single neuron could address the feature-binding problem at the dendritic level.

The work extends a previous study (https://doi.org/10.7554/), to which the author also contributed.

The author models two calcium thresholds (LTP/LTD) from two different calcium sources (NMDA/VGCC), and these thresholds are flexible (metaplasticity rule, similar to BCM), which is claimed to be necessary for successful learning of both FBP and NFBP (linear and nonlinear feature binding problem with 1 or 2 patterns). The role of each threshold seems to be opposite and complementary. One extra condition has been added: an upper threshold for LTP. This threshold serves to stop synaptic strengthening once synapses are strong enough to evoke a plateau. With that, synapses are not strengthened to the maximal value, avoiding strong supralinear integration for irrelevant patterns.

This is summarizes the learning rule well. What I should probably add is that, as mentioned in the introductory response, the article’s first version may cause some confusion about the study’s goal. The goal was to study the role of metaplasticity in the two thresholds, not to solve feature binding in particular. Nevertheless, feature binding is a very useful task to demonstrate the roles of each threshold because the patterns share features: the synapses for the shared features experience both LTP and LTD, and metaplasticity is necessary to express (or to converge to) just one plasticity state (either LTP or LTD). More specifically, the threshold for one plasticity process (e.g. LTD) is necessary to express the opposite process (e.g. LTP). However, to be able to study metaplasticity using feature binding, a prerequisite is to have a rule that can solve it. (So, in that sense, it was first necessary to show that the rule solves the tasks before exploring the roles of metaplasticity. The article has been revised to clearly state this.)

Strengths:

The current model implements not only local synaptic plasticity but also metaplasticity and solves the FBP at the dendrite level. Another strong aspect of the model is that metaplasticity in the LTD threshold protects strengthened synapses from weakening. In this way, as the author mentioned, metaplasticity is able to protect learned patterns from being forgotten or weakened and prevent irrelevant patterns from being stored. This is a nice modelling example of metaplasticity being helpful in preventing the catastrophic interference or forgetting (as has been explicitly discussed in a recent article https://doi.org/10.1016/j.). The author might want to briefly mention or emphasize this aspect of the model, which might be interesting also for the AI community.

This is a great suggestion, thank you! I have instead emphasized a related aspect of the model: that it can solve the plasticity-stability dilemma (when taking away the closed-loop implementation of metaplasticity). I did not focus on catastrophic forgetting, because in machine learning it has a slightly more specific meaning: forgetting that occurs in sequential learning of different tasks. Instead, I exemplified the plasticity-stability dilemma using reversal learning (the reward policy is switched in the middle of the simulation), which can be viewed as a special case of sequential learning of opposite tasks. So, strictly speaking, there is no demonstration that metaplasticity prevents catastrophic forgetting (in the more general sense), but there is a demonstration that it retains (stabilizes) what has been learned (within the same task). The latter does suggest that metaplasticity could help prevent catastrophic forgetting, as well.

Weaknesses:

(1) What is novel in the current paper as compared to Khodadadi et al. eLife 2025? That is not completely clear and should be made clearer. Is it only a minor difference related to the fact that the new learning rule has metaplasticity in both calcium thresholds and is simpler? This seems to be just an incremental increase in knowledge/methods. Can the author defend his paper against this point from the „devil's advocate"? How is the conclusion of the author in the abstract that „metaplasticity in both thresholds is necessary" reconcilable with his previous publication (Khodadadi et al. eLife 2025), in which only metaplasticity in one threshold was successful in solving the nonlinear feature binding problem?

Thank you for raising this question, since the answer was not clear in the first version. The study may be an incremental increase in methods, but is (in my view) a solid increase in knowledge, giving concrete new insights into metaplasticity (an area that is still comparatively little understood). The two main conclusions are:

(1) In synapses that are exposed to both the LTP and the LTD process, metaplasticity is necessary to ultimately express just one of them (converge to either LTP or LTD).

(2) Metaplasticity in the threshold regulating one plasticity process is necessary for expressing the opposite process (metaplasticity in the LTD threshold is necessary for expressing LTP, and vice versa).

The article has been substantially revised to clearly state these main conclusions (as well as the study’s goal), while a full list of all conclusions is given in points 1-7 the Discussion.

The conclusion in the abstract that „metaplasticity in both thresholds is necessary" is stated too generally, and it was meant to refer to this rule only. I have now removed it from the abstract, since the study’s goal was not to solve feature binding. Instead, the focus is on the two conclusions above as well as other conclusions obtained from expanding the study with reversal learning and using a single threshold (prompted by the comments from yourself and reviewer #1).

Lastly, the differences compared to Khodadadi et al. 2025 (eLife) are:

(1) Yes, the rule is simpler and has metaplasticity in both calcium thresholds, instead of just one in Khodadadi et al. 2025 that operates according to a different mechanism (where the threshold, along with the entire LTP plasticity kernel, moves in a direction opposite of the calcium signal).

(2) Plateau potentials are used here, versus the “boosting” nonlinearities in Khodadadi et al. 2025.

(3) The most detailed calcium diffusion model to date for SPNs is implemented in this study (by Dorman et al. (2018)). This was a necessary addition to avoid non-monotonous increases in calcium (shown in Fig. 13A, B). No calcium diffusion between neuronal compartments is implemented in Khodadadi et al., 2025. This is indeed a small methodological difference, but not trivial to implement.

The relation to Khodadadi et al., 2025 is now stated both in the introduction in lines 137-144, and at greater length in the discussion in lines 856-870.

Perhaps I should note that the two learning rules (this one and that in Khodadadi et al. 2025) developed in parallel, this one being based on formulations similar to Gütig et al. 2003 by adding metaplasticity, while the rule in Khodadadi et al. 2025 was inspired from the ideas in Schiess et al. (2016) and Urbanczik and Senn (2014) (see list of references provided). The two studies have different goals (and in that sense one was not meant to be an extension of the other):

- Khodadadi et al. (2025) studies whether the NFBP can be solved by SPNs,

- this article studies the role of metaplasticity in the two calcium thresholds that exist in dSPNs; it uses feature binding as a task because the role of metaplasticity is exposed precisely by the shared features in the task (they cause shared synapses to undergo competing LTP and LTD, and metaplasticity is necessary to direct synapses into just one outcome); importantly, feature binding is also relevant for the striatum (now described in the introduction in lines 126-136)

Finally, to obtain the conclusions in this study about the roles of the two thresholds, it was necessary to devise a suitable rule. They cannot be obtained neither with the rule in Khodadadi et al. 2025, nor with any other rule, because none have a formulation with two separate thresholds for LTP and LTD. Phrased more generally, one needs a new/different rule (new/different assumptions) to obtain new/different conclusions, even if the new rule may be related to existing ones. That said, a rule with a single threshold would have reached conclusion 1. above, as is now shown in the article in Fig. 12 and Figure 12 – figure supplements 1 – 4, but SPNs have two thresholds, prompting the use of a rule with two thresholds. Also, although not demonstrated with the cascade model (by Fusi et al. (2005) and its extensions), conclusion 1. should be obtainable with it, and is implied by the dynamics of the cascade model (which uses one threshold to control transitions between strong and weak synapses). More on the cascade model is given below and in the revised article.

Hopefully that provided a clearer answer. If you have more questions, please let me know so I can try to clarify further.

(2) As far as I can judge without testing the model, metaplasticity causes thresholds to monotonically increase during systematic pattern presentation, which stabilizes weights and allows pattern separation. Due to the closed-loop nature of the current implementation, where metaplasticity only happens if plasticity happens, this also effectively locks patterns in place. However, flexible learning is an essential mechanism for survival. Imagine a mutation event takes place and bananas suddenly become red and/or strawberries turn yellow. It seems that the current model would be unable to adapt to these new patterns even if rewards were to be shifted. While out of the scope of the study, due to its importance, I feel that pattern shifting/relearning should at least be briefly discussed. How could the model be improved to allow relearning?

This is a very important question, and I think it is in fact within the scope of the study, so it is now addressed using reversal learning as a task (the reward policy is switched in the middle of the simulation, after the FBP and NFBP are learned). As you pointed out, the closed-loop formulation of metaplasticity effectively locks the patterns in the dendrites, and reversal learning cannot be solved (Fig. 9). On the other hand, if the conditions for metaplasticity are relaxed by allowing any calcium levels to trigger metaplasticity (i.e. no longer have a closed-loop formulation requiring that [Ca]NMDA or [Ca]L-type be above their thresholds), reversal learning in both the FBP and NFBP is solved (Fig. 10 and Figure 10 – figure supplements 1 – 3). (Note that [Ca]NMDA or [Ca]L-type still need to be above their thresholds for plasticity to occur.)

Recommendations for the authors:

Reviewing Editor Comments:

While the reviewers were very positive, they highlighted limitations on the strength of evidence. Addressing these points could help revise this assessment.

Reviewer #1 (Recommendations for the authors):

(1) Framing the problem as "feature binding" is easy to understand, but it's a slightly narrow view of learning in general. Some mention of how NFBP relates to the XOR problem in the introduction would help you relate your solution to the much broader class of computational problems, since XOR is a more fundamental computational primitive. Non-linear feature learning is very general and relates to all machine learning problems and to computations that occur across every brain region.

Since the goal of the study was not to solve the NFBP or feature binding (but to study metaplasticity), the relation of the NFBP to the XOR is put in the Discussion instead (to avoid further confusion about the study’s goal). Nevertheless, this connection is clearly made in the introduction of the new study that uses 100+ SPN models.

(2) Since the requirement for clustered synapses is one of the main limitations of this learning rule, it would help if there were some discussion of how clustering might occur (e.g., whether there are any other related learning rules or mechanisms that might promote clustering, or if you are assuming that this has to occur entirely through developmental wiring).

This is now discussed in lines 817-827 and related to developing a learning rule that would incorporate structural plasticity.

(3) Page 5, row 157: Another potentially relevant citation here is Bittner et al. 2017, showing that a dendritic plateau potential in silent CA1 neurons drives place field formation without any prior spiking activity.

Thank you, that is indeed a very relevant citation here! It is now added in line 222.

(4) Figure 4C: It would be helpful to add a sentence in the figure legend explaining what the dashed lines represent (even though you also mention it in the main text).

Added, thank you!

(5) Across all figures, the author should consider using color combinations that are more colorblindfriendly (e.g., cyan/red). As a R/G colorblind person, I find it slightly difficult to see which line is which in the panels that compare scores in the NFBP.

The panels showing the NFBP scores have been changed throughout the article, and all other figures have been checked. If there are any more difficult color combinations, please let me know.

(6) The conventional, widely used abbreviation for dopamine is "DA", not "Da".

Now changed.

(7) A few typos I spotted in the paper:

(a) Row 117 (page 3): missing parenthesis around citation

(b) Row 173 (page 5): "uner" --> under

(c) Row 248 (page 7): repeated sentence "...are first seen above..."

(d) Row 255 (page 7): "NBFP" --> NFBP

These are now fixed, thank you for reporting them!

Reviewer #2 (Recommendations for the authors):

(1) How are synapses distributed in the nonlinear integration case? Are pattern combinations branch-specific, or are features distributed randomly across the whole dendritic tree? Does this matter in any way? In any case, it should be clarified.

Yes, they are branch-specific. In the NFBP, the feature combinations from Figure 3 – figure supplement 1 (and later from Figure 10 – figure supplement 1) are placed on two dendrites, chosen at random from 8 dendrites, in a 20-micrometre dendritic stretch starting around 120 micrometers away from the soma. 12 different trials are run for each feature combination, meaning that 12 randomly chosen pairs of dendrites were tested for each feature combination. In the FBP, the clusters are placed in one dendrite chosen at random from 8 dendrites, at the same distance from the soma as in the NFBP. This information is now added in the caption of Fig. 3.

The idea behind placing clusters on the same dendrite is that, after learning, two strong clusters will produce a plateau (e.g. ‘red’ and ‘strawberry’), while one strong and one weak cluster will not (e.g. ‘yellow’ and ‘strawberry’), as in Figs. 3C2, 3C3 (with the weights shown in Figs. 4B2, 4B3). For this the clusters need to be in the same branch, so that two features are “bound” together “into” a larger, plateau-evoking cluster. Also, a strong and a weak cluster should evoke a sufficiently lower somatic amplitude so that any additional noise will not cause somatic spiking. This is assured by the large voltage jump in the plateau (the all-or-none quality of the plateaus, Fig. 2D1).

If the clusters are distributed randomly on different dendrites, it might or it might not work. Fig. 10 of Oikonomou et al. (2012) shows that dendritic spikes (that individually do not evoke somatic spiking) summate sublinearly at the soma in pyramidal neurons: weak + weak cluster = no spiking (Fig. 10A); strong + strong = no spiking (Fig. 10B); but also strong + strong = spiking (Fig. 10C). To reliably solve the NFBP, a large enough supralinearity is necessary instead, so one should test how the soma integrates dendritic spikes/plateaus in SPNs. As the goal of this study was not to solve the NFBP, this is left for the future (perhaps within the new ongoing study focusing on the NFBP).

(2) To produce dendritic plateau potentials, the model (as in the previously published model - Trpevski et al. 2023) implements glutamate spillover activating extrasynaptic NMDARs. This is an interesting mechanism, but is there empirical evidence supporting this way of generating plateau potentials? Are synaptic NMDARs not sufficient? If not, which experiments support the role of nonsynaptic ones?

This is still an ongoing area of research, indicating that glutamate spillover is regulated by reuptake from glial cells, which could even reverse function and excrete glutamate (Rusakov and Stewart, 2021; Malarkey and Parpura, 2014).

Except for the experiments from Szapiro and Barbour, 2007 showing that climbing fibers in the cerebellum signal to molecular-layer interneurons exclusively through glutamate spillover, there are no other “in-vivo-like” experimental conditions which show that spillover activates extrasynaptic NMDARs (eNMDARs). The strongest other in vitro data come from the following experiments:

(1) Chalifoux and Carter (2011), where glutamate reuptake by transporters was blocked with TBOA, which promoted NMDA spikes evoked by synaptic stimulation, suggesting that glutamate spillover activates eNMDARs. A similar experiment was done in Suzuki et al. (2008), where, in addition, synaptic NMDARs were blocked, leaving only eNMDARs available to trigger plateaus.

(2) Glutamate iontophoresis, which ejects glutamate directly into the extrasynaptic space, activates plateaus once the iontophoretic current is strong enough (Oikonomou et al., 2012), with or without blocking glutamate reuptake (Suzuki et al. 2008).

(3) Repetitive synaptic stimulation is thought to produce glutamate spillover (Suzuki et al. 2008; Oikonomou et al., 2012). For example, two synaptic shocks trigger NMDA spikes, while 5 synaptic shocks trigger plateaus in pyramidal neurons (Oikonomou et al., 2012).

On the other hand, studies that employ glutamate uncaging at spines, which should activate eNMDARs much less, evoke NMDA spikes instead (Losonczy and Magee, 2006; Branco et al. 2010). Compared to plateaus, these have much smaller amplitudes, and the size of the supralinearity (the voltage jump) is much smaller than in the plateaus obtained with glutamate iontophoresis (compare Fig. 3 in Losonczy and Magee, (2006), Fig. 3B in Branco et al. (2010) to Fig. 4C in Oikonomou et al. (2012), Fig. 6 in Oikonomou et al. (2014)).

The reason why glutamate spillover was implemented is precisely this robust all-or-none quality of the plateaus, i.e. the large jump in somatic voltage (Fig. 2D1) after a threshold level of stimulation (when the stimulus intensity increases in equal amounts). Without spillover, plateaus are graded in amplitude (and duration, Fig. 2B1 in Trpevski et al. (2023)), and the NFBP cannot be solved (Fig. 6 in Trpevski et al. (2023)).

(3) The check for dependence on initial conditions of theta_LTP/LTD is missing. Especially for the conditions of the thresholds being fixed. How do you decide what fixed value to use? How about having a corresponding limit for weights and not for thresholds? Why is there still an increase in threshold going on even when weights reach the maximum?

The initial conditions for θLTP and θLTD are also an important question. The idea behind starting with low values of the thresholds (or fixing them to low values) is to make the synapses flexible for learning. One can view metaplasticity’s role as a “lock” on the weights, locking them once learning is done, and unlocking them (making them flexible) when something needs to be learned. Which signals determine when to lock or unlock the weights is what the metaplasticity rule implements, and is insufficiently understood experimentally. In any case, it makes sense to start with flexible synapses (low thresholds) at the beginning of learning; otherwise, no/little learning will take place (Figure 9 – figure supplement 2, the panels for “thresholded” metaplasticity). Throughout the article I choose the initial values of the thresholds to be lower than the calcium evoked by weakened synapses, ensuring that any calcium signal will trigger plasticity (but higher values also work).

Instead of doing a scan over the initial conditions of θLTP and θLTD, I chose to show the following:

(1) An example with initial threshold values higher than the calcium amplitudes: these “lock” the weights (Figure 9 – figure supplement 2, panels for “thresholded” metaplasticity) from the start and no learning can occur.

(2) An example initialized with high values for the thresholds, but where metaplasticity can be triggered by any calcium amplitudes (no closed-loop in the metaplasticity rule, Figure 9 – figure supplement 2, panels for “relaxed” metaplasticity): here the thresholds adapt and learning can take place afterwards.

This is supposed to illustrate a mechanism that can “unlock” weights. A scan over the initial conditions of θLTP and θLTD will give the lowest calcium level that the thresholds can be initialized to so that “thresholded” metaplasticity will work. But it is not important to know these precise values to obtain the conclusions in the article, which is why I did not do such a scan. (The above is treated in the article in the section on reversal learning, lines 569-577 and 628-633.)

Also, when the rule is used for learning, it is not meant to have fixed thresholds, as synapses cannot stabilize and nothing will be stored (Figs. 5, 6, and 8). Fixing the thresholds to a low value was meant to show what happens without any metaplasticity (or if metaplasticity were “dysfunctional”). If in reality a mechanism exists to fix the thresholds to a low value or a high value, fixing would be temporary (until necessary to keep the weights flexible or locked, respectively).

I am probably not understanding this part of the comment: “How about having a corresponding limit for weights and not for thresholds?” There are two limits, the maximal and minimal weights in the rule wmax and wmin, and are initialized to random values within the interval [0.3, 0.35], but this is probably not what you mean.

When the weights reach their maximum, the thresholds still increase because they have not reached the calcium amplitudes evoked by the maximal weights. (This is most visible with supralinear integration, where strengthened weights evoke plateaus Fig. 4C2-C4.) The thresholds move towards the calcium amplitudes with a rate ηθ, which if made very high (as in Figure 4 – figure supplement 11), will make the thresholds rise as fast as the weights. But this is not good, as it will “lock” the learning processes in the synapses too soon (i.e. shorten the window where they are flexible). Neither is a too low ηθ good, as then the synapses are flexible for a very long time and learning of the NFBP is prolonged (Figure 4 – figure supplements 10B3, 10B4, 10E3, learning simulations last longer than in Fig. 4).

(4) Calcium modelling: The author writes that the largest voltage plateaus do not correspond to the largest calcium amplitude because of the plateau's voltage approaching the NMDAR reversal potentials, which might be a problem for the plasticity rule. To solve this, the author tried to implement a monotonic increase of calcium with increasing plateau by implementing axial calcium diffusion and buffering. My question is, is the monotonic increase realistic? Is not the NMDAR reversal effect, described above, a realistic scenario in dendrites?

It seems to be realistic. The closest experiment that I know of is in Fig. 2C2, D2 in Oikonomou et al. (2012), where 2 synaptic shocks produce an NMDA spike, and 5 shocks produce a plateau. The calcium dye shows a higher response from the plateau, suggesting a higher calcium amplitude (with a monotonic increase). Moreover, in the detailed calcium model for SPNs by Dorman et al. (2018), larger synaptic clusters produce monotonically higher calcium amplitudes. These two pieces of evidence suggest the calcium amplitudes are not affected by the membrane voltage approaching to the reversal potential. Most likely, this happens because of strong intracellular calcium buffering.

Similarly, the experiments in Figs. 5 and 8 in Losonczy and Magee (2006) with glutamate uncaging at spine heads show a monotonic increase in calcium dye flourescence as the cluster size is increased. However, these experiments evoke NMDA spikes, which might not have approached the NMDAR reversal potential as closely as plateaus do (so, it is theoretically possible that monotonicity does not hold if a plateau were evoked, although it seems unlikely). This is now added to the Methods in lines 957-961.

(5) Function: What is the actual function of the SPN, and how does it relate to FBP/NFBP learning? SPNs are involved in motor control/decision making and generally in sensorimotor tasks. Rather, use an example for that than a visual stimulus, though I understand it is more easily illustrated. Is feature binding what SPNs do and have to solve?

It is not known yet what SPNs do precisely. The initial action selection role of the basal ganglia is currently being challenged or replaced by a role in movement initiation and/or invigoration (with action selection being done by the cortex, see e.g. Thura and Cisek (2017)). In any case, SPNs receive convergent input from sensory, motor, limbic and associative areas, as well as thalamic and neuromodulatory inputs (with topographical projections indicating functional specialization). In that sense, feature binding is particularly relevant for the striatum, as diverse features from different areas would arrive there. The basal ganglia participate in non-motor loops as well, so the striatum may have a role in initiation and termination of cognitive processes such as planning and attention, and in regulating emotional and motivated behavior (Purves et al. 2018).

The role of the striatum and the relevance of feature binding is now stated in the introduction in lines 126-136, and a task with features more suited for the striatum is given in Figure 1 – figure supplement 1 (inspired from similar tasks in Bernklau et al. (2024)). I have kept the example from the visual system in the main text simply because it is widely used throughout the literature and easily recognizable, while clearly stating that other features would be involved in the striatum and pointing to the example in Figure 1 – figure supplement 1. Hopefully this will make a good compromise between ease of reading and relevance to the striatum.

(6) The population of striatal projection neurons can express dopamine D2 receptors and not only D1 receptors. SPNs with different receptors thus undergo synaptic plasticity according to different rules. Would that lead to similar results? A combination of both populations?

This is also being tested in the new study, and preliminary results indicate the answer is “yes” (see Author response image 4). Contrary to the dSPNs, the indirect-pathway SPNs (iSPNs), which predominantly express D2 receptors, are thought to be involved in suppressing competing or related movements. The learning rule is almost the opposite to dSPNs: to trigger LTP, a dopamine pause is needed, and to trigger LTD a dopamine peak is needed (Fig. 4B and Shen et al. (2008)). So, with such a rule, iSPNs will store the irrelevant patterns and spike to them, and be silent to the relevant patterns (which indeed happens after learning in Author response image 4F2, F3). This aligns well with their role to suppress movements – reaching for the irrelevant patterns will be suppressed, while reaching for the relevant patterns will be disinhibited (the indirect pathway inhibits movement initiation).

(7) Other synaptic plasticity & metaplasticity models (possible comparison or discussion): https://pubmed.ncbi.nlm.nih./ https://linkinghub.elsevier.(https://link.springer.com/ For example, in the last model, there is only one modification LTP/LTD threshold, but it is different from the voltage threshold (which would be comparable to calcium thresholds in the current paper - see the voltage threshold different from the modification threshold here: https://doi.org/10.1371/). The modification threshold favors LTP or LTD - depending on the previous spiking activity of the cell and effectively works as an anti-Hebbian firing rate „homeostasis" mechanism. How is the firing rate homeostasis maintained in the current paper/model? Can you explain how this model is related to the previous models of Benuskova and Abraham, and also Clopath (there was also a metaplasticity version of the Clopath model), in terms of firing rate stability? The current model implements not only local synaptic plasticity but also metaplasticity, which is a nice feature, and seems to not only solve the FBP but also maintain firing stability. Is the stability of firing a consequence of the hard bounds for synaptic weights, or is it also a consequence of plasticity and metaplasticity? I like that each synaptic weight is changed independently based on local calcium concentration, with its own LTP and LTD thresholds.

I will first describe what happens to the firing rate in this model, and then compare with the references you mentioned. In the FBP with linear integration, yes, the firing rate stability is only maintained because of the hard bounds for synaptic weights (Figure 4 – figure supplement 5A1, the neuron goes into depolarization block without wmax). Without wmax, the weights would grow until [Ca]NMDA saturates in the spine (which does not happen for most spines within the simulation time in Figure 4 – figure supplement 5B1).

In the FBP with supralinear integration and the NFBP, the weights are prevented from increasing by the upper threshold for LTP, θLTP. If θLTP and wmax are gone, synapses also grow without bound (Figure 4 – figure supplement 5B2). Regardless of how high the synaptic weight becomes, firing rate stability is ensured by the plateau potential, as the maximally achievable voltage is limited by the NMDAR reversal potential (Figure 4 – figure supplement 5A2, a somatic depolarization block does not happen as in Figure 4 – figure supplement 5A1). This is part of the plateaus’ function to provide dynamic range compression, as shown experimentally in Figs. 3E, 4D and 12 in Oikonomou et al. (2012). (Of course, such high weights are unrealistic, but they happen in the model without wmax and without θLTP). This is now described in the text in lines 397-407.

Regarding the other studies, the second link did not work, so I will refer to Jedlicka et al. (2015), Clopath et al. (2010) and Zenke et al. (2013). (The rule in Jedlicka et al. (2015) is the same as in Benuskova and Abraham, (2007) but with a more detailed neuron model.) In these studies, there is a threshold representing a (weighted) average of the postsynaptic activity (postsynaptic voltage or spiking). In Clopath et al. (2010) and Zenke et al. (2013) it modulates only the amount of LTD, while in Jedlicka et al. (2015) it modulates the amounts of both LTP and LTD that occur due to plasticity. The threshold is global for a neuron (i.e. all synapses use the same threshold), and allows for heterosynaptic effects that compensate increased excitation from synaptic strengthening with stronger weakening in depressed synapses (thus maintaining stable firing rates). (However, the rule in Jedlicka et al. (2015) and Benuskova and Abraham, (2007) has only one threshold, there is no separate voltage threshold. Indeed, the threshold is a low-pass filtered variable of the input spike train, so is similar to calcium concentration.)

There are no such effects in the current learning rule. Some compensatory effects are visible when comparing supralinear integration with and without an upper threshold θLTP, as e.g. in Fig. 4B2 and Figure 4—figure supplement 3B1. With θLTP, synapses do not increase to wmax, and without it, they do. As a consequence, the weakened synaptic cluster is weakened less when using θLTP (‘yellow’ synapses in Fig. 4B2), and more without θLTP (‘yellow’ synapses in Figure 4—figure supplement 2B1). The increased weakening in the latter case is because the strengthened ‘strawberry’ synapses have reached wmax, and contribute to spiking for the irrelevant pattern (‘yellow strawberry’). As a result, the weakened ‘yellow’ synapses have to decrease more so no spiking for the irrelevant pattern occurs. In effect, the increased strengthening in the ‘strawberry’ synapses drives increased weakening in the ‘yellow’ synapses. However, this is a result of the task structure, i. e. that the patterns share features, and not from the learning rule.

The above is briefly explained in the Discussion in lines 778-786.

(8) The last sentence in the discussion is a bold claim: SPN can completely solve NFBP alone. Is that statement too strong, or is it a sign of multiple different mechanisms implementing the same function (i.e., degeneracy)?

The sentence was not meant to be a bold claim. It currently says:

“This indicates that ... SPNs might completely solve the task ... Whether this is true will be explored in another study …”

I have now toned this claim down, as the purpose was only to point to the new study that explores this question (using with 100+ SPN models, in two learning regimes and varying cluster location). Nevertheless, the preliminary results in the new study suggest that SPNs can indeed solve the NFBP (if the plateaus are all-or-none), so I agree that it is a sign of multiple mechanisms implementing the same function.

Figures/Table:

(1) Figure 2: The layout of the figure is confusing. B shows distributed input as in upper A, and C shows clustered input as in lower A. This should be illustrated better (e.g., as in Figure 3 or 4). The term cluster size is confusing as well because it also refers to the distributed inputs; better use the number of synapses here. The color choice is difficult, as similar colors are chosen for cluster size and for differentiating linear and nonlinear integration in D.

Thank you for the detailed comments! The figure has been reworked accordingly.

(2) Figure 4: C What are dashed lines?

They indicate the upper threshold for LTP, θLTP.Thank you for noticing this, it is now added in the figure caption, and in other figures where it was missing.

(3) Figure 5 and Figure 6: B is missing.

These figures have been reworked to match the revision of the text describing the role of metaplasticity in the shared synapses. Now, the FBP is described first, followed by the NFBP, so the figures have been merged with their figure supplements (where the FBP results used to be in the first version).

(4) Figure 4 Supplement 1: What do the colors and regions mean in B3? Why show here and not for others?

The main text in lines 311-312, 331-333, and 358-359 refers to these regions when explaining the dynamics of the calcium, the thresholds and the weights (e.g. the pink region shows that [Ca]L-type in the ‘strawberry’ synapses evoked by ‘yellow strawberry’ is much lower than the calcium threshold for ‘strawberry’, protecting these synapses from weakening). They are simply highlighted with different colors, so readers know precisely where to look when reading the main text. The figure caption has been updated to state this.

Also, the figure caption says that the same regions exist in all panels, although they have not been highlighted. I could have chosen any panel to mark these regions, and I chose B3 because it looked like it had the least visual clutter.

(5) Table 1: Typo for LTD

Now fixed.

(6) Line 21: „thehsold"

Now fixed.

(7) Line 50: Closing parentheses missing.

Now fixed.

(5) Line 55: "All metaplasticity models so far have only one modifiable threshold". This is not true as it stands. Multiple sliding thresholds have been proposed before:

- experimentally by Ngezahayo et al. (2000)

- multiple pathways reviewed by Abraham (2008)

- multiple structural LTP thresholds observed experimentally by Ueda et al. (2022)

- models of synaptic states and cascade models can be considered to employ multiple thresholds (depending on state); e.g., Fusi et al. (2005)

Thank you for providing these references! The sentence has been changed (it was supposed to refer only to computational models), and the studies have been added in the text where appropriate.

Importantly, Fusi et al. (2005) is now being discussed throughout the text, as it contains important results and conclusions about metaplasticity which are used as comparison. (In the first version of the article I did not include it because I was not sure about the interpretation of the thresholds – strictly speaking there is only one plasticity threshold (q, determining whether a synapse will switch sign), while the other is a metaplasticity threshold (p, determining whether the synapse will change its metaplastic state). However, the precise number of thresholds is not as important, as the study’s results are relevant to compare with.)

(9) Lines 117/118: Missing parentheses around reference.

Fixed.

(10) Line 140: "a small difference". I feel like "small" is a bit of an understatement here, as from what I can see, the differences are almost a magnitude -- see blue/black traces in Figure 2B4 vs. 2C4.

Thank you for noticing this. It was meant to refer only to the amplitude of the somatic voltage when the cluster size is small (in Fig 2D1). The text has been clarified now.

(11) I think it's unnecessary to rewrite Eq. 1 in Eq. 3.

I decided to keep this.

(12) Lines 181-183: It would probably still be useful to show that NFBP can't be solved by distributed inputs.

It is now shown in Figure 4 – figure supplement 1, and briefly described in the text in lines 298300.

(13) Line 210: Typo, double "(the"

Fixed, thank you.

(14) Lines 248-249: Typo, double partial sentence.

Now fixed.

(15) Line 760: The year in the „STDP rule endowed with the BCM sliding threshold accounts for hippocampal heterosynaptic plasticity" should be corrected to 2007. Also, all other references should be checked for mistakes and missing pages, etc. (see e.g., also line 882).

Thank you for noticing this! The references have been checked.

References

Benuskova, L., Abraham, W.C. STDP rule endowed with the BCM sliding threshold accounts for hippocampal heterosynaptic plasticity. J Comput Neurosci 22, 129–133 (2007). https://doi.org/10.1007/s10827-006-0002-x

Bernklau TW, Righetti B, Mehrke LS, Jacob SN. (2024) Striatal dopamine signals reflect perceived cue–action–outcome associations in mice. Nature Neuroscience 27(4):747–757. doi: 10.1038/

Branco, T., Clark, B. A., & Häusser, M. (2010). Dendritic Discrimination of Temporal Input Sequences in Cortical Neurons. Science, 329(5999), 1671–1675. http://www.jstor.org.focus.lib.kth.se/stable/40803162

Chalifoux JR, Carter AG (2011) Glutamate Spillover Promotes the Generation of NMDA Spikes. J. Neurosci. 31(45):16435–16446. doi:10.1523/JNEUROSCI.2777-11.2011

Clopath, C., Büsing, L., Vasilaki, E. et al. Connectivity reflects coding: a model of voltage-based STDP with homeostasis. Nat Neurosci 13, 344–352 (2010). https://doi.org/10.1038/nn.2479

Dorman, D. B., Jędrzejewska-Szmek, J., Blackwell, K. T. (2018) Inhibition enhances spatiallyspecific calcium encoding of synaptic input patterns in a biologically constrained model. elife 7:e38588 https://doi.org/10.7554/eLife.38588

Du K., Wu Y., Lindroos R., Liu Y., Rózsa B., Katona G., Ding J.B., & Kotaleski J.H. (2017) Celltype–specific inhibition of the dendritic plateau potential in striatal spiny projection neurons, Proc. Natl. Acad. Sci. U.S.A. 114 (36) E7612-E7621, https://doi.org/10.1073/pnas.1704893114.

Gjorgjieva J., Clopath C., Audet J., & Pfister J. (2011) A triplet spike-timing–dependent plasticity model generalizes the Bienenstock–Cooper–Munro rule to higher-order spatiotemporal correlations, Proc. Natl. Acad. Sci. U.S.A. 108 (48) 19383-19388, https://doi.org/10.1073/pnas.1105933108

Gütig R., Aharonov R., Rotter S., Sompolinsky H (2003) Learning Input Correlations through Nonlinear Temporally Asymmetric Hebbian Plasticity. J. Neurosci. 23 (9) 3697-3714; DOI: 10.1523/JNEUROSCI.23-09-03697.2003

Hedrick, N.G., Lu, Z., Bushong, E. et al. (2022) Learning binds new inputs into functional synaptic clusters via spinogenesis. Nat Neurosci 25, 726–737 . https://doi.org/10.1038/s41593-022-01086-6

Hwang F, Roth R, Wu Y et al. (2022) Motor learning selectively strengthens cortical and striatal synapses of motor engram neurons Neuron 110, 2790-2801.e5

Jedlicka P, Benuskova L, Abraham WC. (2015) A Voltage-Based STDP Rule Combined with Fast BCM-Like Metaplasticity Accounts for LTP and Concurrent “Heterosynaptic” LTD in the Dentate Gyrus In Vivo. PLOS Comput Biol 11(11):1–24. https://doi.org/10.1371/journal.pcbi.1004588

Kirchner, J.H., Gjorgjieva, J. (2021) Emergence of local and global synaptic organization on cortical dendrites. Nat Commun 12, 4005. https://doi.org/10.1038/s41467-021-23557-3

Lindroos R, Hellgren Kotaleski J (2021) Predicting complex spikes in striatal projection neurons of the direct pathway following neuromodulation by acetylcholine and dopamine. Eur J Neurosci. 53:2117–2134. https://doi.org/10.1111/ejn.14891

Losonczy A, Magee J (2006) Integrative Properties of Radial Oblique Dendrites in Hippocampal CA1 Pyramidal Neurons. Neuron 50: 291-307

Malarkey EB, Parpura V (2008) Mechanisms of glutamate release from astrocytes, Neurochemistry International 52(1–2): 142-154, doi: 10.1016/j.neuint.2007.06.005.

Oikonomou KD, Short SM, Rich MT and Antic SD (2012) Extrasynaptic Glutamate Receptor Activation as Cellular Bases for Dynamic Range Compression in Pyramidal Neurons. Front. Physio. 3:334. doi: 10.3389/fphys.2012.00334

Oikonomou KD, Singh MB, Sterjanaj EV and Antic SD (2014) Spiny neurons of amygdala, striatum, and cortex use dendritic plateau potentials to detect network UP states. Front. Cell. Neurosci. 8:292. doi: 10.3389/fncel.2014.00292

Purves D, Augustine GJ, Fitzpatrick D, Hall WC, LaMantia AS, White LE, et al. (2018) Neuroscience. 6 ed. New York:Oxford University Press

Rusakov DA, Stewart MG (2021) Synaptic environment and extrasynaptic glutamate signals: The quest continues, Neuropharmacology 195: 108688, https://doi.org/10.1016/j.neuropharm.2021.108688

Schiess M, Urbanczik R, Senn W (2016) Somato-dendritic Synaptic Plasticity and Errorbackpropagation in Active Dendrites. PLoS Comput Biol 12(2): e1004638. https://doi.org/10.1371/journal.pcbi.1004638

Suzuki, T., Kodama, S., Hoshino, C., Izumi, T. and Miyakawa, H. (2008), A plateau potential mediated by the activation of extrasynaptic NMDA receptors in rat hippocampal CA1 pyramidal neurons. European Journal of Neuroscience, 28: 521-534. https://doi.org/10.1111/j.14609568.2008.06324.x

Thura D, Cisek P. (2017) The Basal Ganglia Do Not Select Reach Targets but Control the Urgency of Commitment. Neuron 95(5):1160–1170.e5. doi: 10.1016/j.neuron.2017.07.039)

Tran-Van-Minh A, Cazé RD, Abrahamsson T, Cathala L, Gutkin BS and DiGregorio DA (2015) Contribution of sublinear and supralinear dendritic integration to neuronal computations. Front. Cell. Neurosci. 9:67. doi: 10.3389/fncel.2015.00067

Trpevski D, Khodadadi Z, Carannante I and Hellgren Kotaleski J (2023) Glutamate spillover drives robust all-or-none dendritic plateau potentials—an in silico investigation using models of striatal projection neurons. Front. Cell. Neurosci. 17:1196182. doi: 10.3389/fncel.2023.1196182

Urbanczik R, Senn W (2014) Learning by the Dendritic Prediction of Somatic Spiking. Neuron 81: 521-528

Plateau potentials (A1, B1) and their somatic amplitudes in dSPNs (A) and iSPNs (B) when the location of the synaptic cluster is varied.

Distally evoked plateaus have a smaller somatic amplitude in both dSPNs and iSPNs, bust still exhibit a nonlinear jump in the somatic voltage. In (A2, B2) all 71 dSPN and 34 iSPN models were tested, color coded with respect to their excitability to a synaptic cluster of 10 synapses.

Two learning simulations on the NFBP where different metaplasticity rates for LTP and LTD were used.

(Top) The metaplasticity rate for LTD is higher, stopping the LTD process prematurely (D1, D2), resulting in less weakening of the synapses (B1, B2) and spiking for one irrelevant pattern (A1). (Bottom) The metaplasticity rate for LTP is higher, preventing frequent synaptic strengthening(C1, C2). As a result, it takes a longer time for synapses to strengthen and trigger plateaus (B3, B4).

Learning on the NFBP with a nonlinear (quadratic) metaplasticity rule, which implements the BCM rule.

(A) Somatic and dendritic voltage before and after learning. (B-D) Evolution of synaptic weights (B), LTP (C) nad LTD thresholds (D). Weights in (B) seem to stabilize around a high value once the LTP threshold reaches the maximal [Ca]NMDA; however, the value is too high and triggers plateau potentials for all patterns.

Learning to solve the NFBP in one dSPN and one iSPN model.

(A, B) Schemes describing the requirements for cortico-striatal synaptic plasticity in dSPNs (A) and iSPNs (B). (C) Dopamine peaks are emitted from the midbrain after the relevant patterns, and dopamine pauses after the irrelevant patterns. (D) Stimulation protocol. (E, F) A single learning simulation in a dSPN (E) and an iSPN (F) model. Before learning the neurons spike to all patterns (due to additional feature-unspecific distributed inputs, shown with the third panels in E3, F3). After learning, the dSPN spikes only for the relevant patterns (E2) and the iSPN only for the irrelevant ones (F2). (E3, F3) The evolution of clustered and distributed synptic weights (In dendrite 1 of the iSPN, the synapses for 'red' experienced some strengtheneing after being weakened, but this does not seem to affect learning in this example.) The neurons in th enew study receive less background synaptic input than in the current article, and additional distributed synapses are needed along with a plateau for somatic spiking, making for a more challenging learning scenario.

Version 2

Abstract

Synaptic plasticity has been shown to occur when calcium, flowing into the synapse due to incoming stimuli, surpasses a threshold level. This threshold level is modifiable through a process called metaplasticity. Some neurons, such as the striatal projection neurons, use different sources of calcium as the signal for synaptic strengthening (long-term potentiation, LTP) or weakening (long-term depression, LTD), resulting in them having two thresholds for inducing plasticity. In this study, we show that metaplasticity enables synapses undergoing both LTP and LTD during learning to selectively express just one form of plasticity (either LTP or LTD). To show this, we use the linear and nonlinear feature binding problem (FBP and NFBP) because their input patterns share features, exposing synapses to such competing LTP and LTD processes. In particular, we identify opposite and complementary roles of metaplasticity in the two thresholds for inducing LTP and LTD: metaplasticity in one threshold (e.g. LTD) allows synaptic plasticity of the opposite type (e.g. LTP) to be properly expressed. This happens because metaplasticity in the LTD threshold protects strengthened synapses from weakening, thus allowing them to persistently increase during learning (and encode learned patterns). Similarly, metaplasticity in the LTP threhsold prevents weakened synapses from strengthening, thus allowing them to persistently decrease. Under more general conditions for triggering metaplasticity, reversal learning can also be solved, demonstrating metaplasticity’s importance for solving the plasticity–stability dilemma. Finally, we show that even though a single calcium threshold is sufficient for solving the FBP, NFBP and reversal learning, two calcium thresholds allow separate control over LTP and LTD.

Introduction

Learning is thought to be implemented by several processes in the nervous system, one of which is synaptic plasticity - the modification of synapses based on incoming stimuli to the neuron (Citri and Malenka, 2007; Kandel et al., 2014; Takeuchi et al., 2014; Abraham et al., 2019). Most often, incoming stimuli cause calcium influx in the synapse, an important “second messenger” ion that triggers synaptic plasticity. However, not all levels of calcium can elicit synaptic plasticity - threshold levels of calcium have been reported above which plasticity is triggered (Yang et al., 1999; Nishiyama et al., 2000; Cormier et al., 2001; Shindou et al., 2011). These calcium thresholds for inducing plasticity are themselves modifiable (plastic) through a process called metaplasticity (Abraham, 2008). Metaplasticity regulates the conditions under which plasticity occurs, i.e. under which synapses are modified.

In many neurons, such as hippocampal CA1 and CA3 neurons, Purkinje neurons, cerebellar Golgi cells, and projection neurons from the core of the nucleus accumbens, the calcium signal for synaptic strengthening (long-term potentiation, LTP) and for synaptic weakening (long-term depression, LTD) comes from different sources (Nishiyama et al., 2000; Hunt et al., 2013; Kohda et al., 1995; Binda et al., 2016; Locatelli et al., 2021; Ji and Martin, 2012). In the direct-pathway striatal projection neurons (dSPNs), a model of which is used in this study, calcium from NMDA receptors (NMDARs) triggers LTP when followed by activation of D1 dopamine receptors (D1Rs), while calcium from voltage-gated L-type channels (Cav1.3) triggers LTD when followed by inactivation of D1Rs (Shen et al., 2008; Yagishita et al., 2014; Fisher et al., 2017; Fino et al., 2010; Shindou et al., 2011; Iino et al., 2020). (Note that to trigger LTD, activation of metabotropic glutamate receptors (mGluRs) is also required (Shen et al., 2008). This requirement assures that LTD is only elicited in active synapses where glutamate transmission occurs; Cav1.3 channels alone have an unspecific response – they can be active anywhere where the voltage is high enough.) A calcium threshold for triggering LTD in SPNs has been experimentally quantified (Shindou et al., 2011). Although a similar threshold for LTP in SPNs has not been reported yet, presumably such a threshold exists, because weaker stimulation protocols do not induce LTP (Fino et al., 2005). (However, probably due to the local striatal microcircuitry, a range of different stimulation protocols can induce a range of different plasticity outcomes, as reviewed in Reynolds and Wickens (2002)). A calcium threshold for LTP has been quantified in hippocampal CA1 neurons (Cormier et al., 2001).

By regulating when plasticity can be induced, metaplasticity has been proposed to prepare neural circuits and networks for learning, stabilize synaptic weights (prevent them from reaching very high or very low values) and to contribute to cognitive effects of disease when it is dysregulated (Hulme et al., 2013). Most computational metaplasticity models have only one modifiable threshold (where synaptic activity below the threshold induces LTD, and above it induces LTP) (Yger and Gilson, 2015; Bienenstock et al., 1982; Jedlicka et al., 2015). On the other hand, the important class of cascade models, whose general form can be seen as having two thresholds, has shown that metaplasticity can solve (several versions of) the plasticity-stability dilemma – it protects stored information from being “deleted” (overwritten) by new synaptic activity, while also allowing for flexible relearning when this information is no longer useful (Fusi et al., 2005; Iigaya, 2016; Farashahi et al., 2017; Khorsand and Soltani, 2017; Jedlicka et al., 2022; Benna and Fusi, 2016). The cascade model is a stochastic model of a number of synapses (not placed on any neuron) that randomly experience plasticity events. In this article we study the role of metaplasticity in learning using models closer to biological neurons: a multicompartment model of a dSPN, and a simple, biologically-grounded learning rule describing plasticity of the dSPN cortico-striatal synapse, to which we add metaplasticity in the two thresholds (Fig. 1A). Using the linear and nonlinear feature binding problem (FBP and NFBP), whose structure, as will be explained later, is particularly suitable to emphasize the role of each threshold, we show that the two thresholds have specific, opposite and complementary roles in learning. Further, using reversal learning in the FBP and NFBP, we show that metaplasticity can also solve the plasticity-stability dilemma. Lastly, we show that while a single threshold is sufficient to solve the FBP, NFBP, and reversal learning, two calcium thresholds allow separate control over synaptic strengthening and weakening.

Synaptic plasticity of the cortico-striatal synapse, the tasks for learning, and the dopamine signals for the two types of patterns in the tasks.

(A) Scheme describing the requirements for synaptic plasticity at the cortico-striatal synapse in dSPNs (adapted from Shen et al. (2008) by adding the calcium thresholds, which are the focus of this study). (B) The FBP and the NFBP, illustrated using features common to the visual system. SPNs might instead receive, e.g. sensoryand motor-related features. (C) Dopamine signals from the midbrain are assumed to arrive in the striatum after every pattern. Dopamine peaks are emitted after the relevant patterns, and dopamine pauses after the irrelevant patterns. (The midbrain and its projections are not explicitly modeled in this study, dopamine is programatically provided to the SPN a certain time after pattern arrival.) This figure is adapted from Khodadadi et al. (2025), licensed under CC BY 4.0 (https://creativecommons.org/licenses/by/4.0/). Figure 1—figure supplement 1. An example of the NFBP with features relevant for the striatum.

Before presenting the results, we introduce the FBP and NFBP. The FBP refers to combining (binding) an object’s individual features, such as color, shape, or motion, into a unified object (Roskies, 1999; Treisman, 1999; von der Malsburg, 1999; Hardcastle, 2017). The binding problem arises because features are often processed in different brain regions, and possibly at different times, yet they need to be combined to represent/construct the perception of the object they correspond to. In its simplest form, the linear FBP consists of discriminating between two patterns, meaning that a neuron should spike when activated by one of them, and remain silent when activated by the other (Fig. 1B, top). Each pattern consists of two features, e.g. shape and color, and one of the features can have two values (e.g. ‘red’ and ‘yellow’ for the color feature, as opposed to just ‘strawberry’ for the shape feature in Fig. 1B, top table). Each feature excites the neuron equally, and the term “linear” refers to the computational complexity of the problem, meaning that a neuron able to perform simple linear summation (integration) of incoming stimuli, such as synaptic potentials summated (added) at the soma, can solve the problem. The NFBP consists of discriminating between four patterns: the neuron should spike for two patterns and remain silent for the other two (Fig. 1B, bottom). Each pattern also consists of two features, and in the NFBP both features can have two values (‘strawberry’ or ‘banana’ for shape, and ‘red’ or ‘yellow’ for color, shown in the bottom table in Fig. 1B). Again, each feature excites the neuron equally, and the additional layer of complexity comes from the requirement to spike for two different patterns whose features do not overlap (‘red strawberry’ and ‘yellow banana’). One of the theoretical solutions to the NFBP requires supralinear integration of inputs (Tran-Van-Minh et al., 2015). SPNs exhibit such supralinearities in the form of prolonged dendritic voltage elevations called plateau potentials, demonstrated in the next section. (Supralinear integration of inputs can also solve the simpler FBP.)

As mentioned above, in addition to calcium influx, dopamine signals are also required for plasticity in SPNs. Because of this, we assume that after a pattern is presented, dopamine signals are emitted from the midbrain, and they indicate whether the pattern was “rewarding” or not (Fig. 1C). The patterns for which the SPN should learn to spike are called relevant patterns, and those for which it should be silent are called irrelevant patterns. We assume that dopamine peaks are emitted for relevant patterns, and dopamine pauses for irrelevant patterns (Fig. 1C). After stimulation of the synapse, if the required calcium signal has reached its corresponding threshold, the dopamine signal determines the direction of plasticity. Dopamine peaks are assumed to activate the D1Rs, enabling LTP, and dopamine pauses are assumed to deplete basal dopamine in the striatum low enough so that D1Rs are not activated, enabling LTD (Yapo et al., 2017).

The reason why the FBP and the NFBP are particularly suitable to demonstrate the role of the calcium thresholds is that relevant and irrelevant patterns share features. In the FBP, the shared feature is the ‘strawberry’ shape (Fig. 1B, top). In the NFBP, both features of a pattern are shared, each feature being shared between one relevant and one irrelevant pattern (Fig. 1B, bottom). As a result, during learning, the synapses representing shared features will experience both dopamine signals (peaks and pauses), guiding them towards both strengthening and weakening. As will be shown, metaplasticity is crucial for “choosing” a direction, i.e. directing plasticity into either strengthening or weakening. In particular, we will see that metaplasticity in the LTD threshold is necessary for strengthening, while metaplasticity in the LTP threshold is necessary for weakening, i.e. that the threshold regulating one plasticity outcome is necessary for properly expressing the opposite plasticity outcome.

Lastly, as the input nucleus of the basal ganglia, the striatum integrates convergent cortical inputs from sensory, motor, limbic and associative areas, as well as thalamic and neuromodulatory inputs (Purves et al., 2018). Based on these inputs it initiates and invigorates goal-directed actions, such as reaching for one of the relevant patterns in the FBP and NFBP (Thura and Cisek, 2017). The cortico-striatal synapses are thought to be a major site for learning to initiate such actions, thus forming associations between environmental stimuli, internal states and rewarding outcomes (Filippo et al., 2009). Because of this, despite being suitable tasks for studying metaplasticity, the FBP, NFBP, and reversal learning are also relevant tasks for the striatum’s function. Throughout this article, feature binding is illustrated with a commonly used example from the visual system, although diverse features from the other brain areas would be involved in the striatum (an example is given in Figure 1–figure supplement 1).

Before continuing to the results, we note that we recently participated in a related study with a different goal – whether SPNs can solve the NFBP (Khodadadi et al., 2025). That study used the same neuron model in a regime with much less supralinear integration, and with a different, biologically more detailed learning rule, which incorporated metaplasticity in the LTP threshold only (operating according to a different mechanism and where the evoking calcium signal does not diffuse across neuronal compartments). On the other hand, the goal of this article is to study the role of metaplasticity in the two calcium thresholds in learning, for which we use a simpler, yet biologically-based learning rule where both thresholds are plastic.

Results

Linear and supralinear integration of inputs by SPNs and corresponding calcium elevations

We first demonstrate linear and supralinear integration in the dSPN model, which are the substrates for solving the FBP and NFBP. Figure 2 compares linear integration of synaptic inputs distributed across the dendrites with supralinear integration of clustered synaptic inputs on one dendrite (Fig. 2A). The number of synaptic inputs is varied from 1 to 30 synapses, and the evoked somatic and spine voltage, as well as the spine [Ca]NMDA and [Ca]L-type (calcium concentrations arising from NMDARs and L-type calcium channels, respectively) are shown in Figs. 2B and 2C. A comparison of the amplitudes of these four quantities between linear and supralinear integration is shown in Fig. 2D. With linear integration, the somatic voltage amplitude increases linearly with the number of synapses (Fig. 2B1 and blue line in Fig. 2D1). Conversely, the spine voltage amplitude remains approximately the same, because spines are distributed across the dendrites and are activated in the same way (due to the distance between them, spines cannot influence each other’s voltage, Fig. 2B2 and blue line in Fig. 2D2). The similar amplitudes of the spines’ voltage result in similar [Ca]NMDA and [Ca]L-type elevations (Figs. 2B3, B4 and blue lines in Figs. 2D3, D4).

Linear versus supralinear integration of synaptic inputs by SPNs.

(A) Illustrations of the two scenarios of linear integration of distributed synaptic inputs by the soma (A1), and supralinear integration of clustered synaptic inputs in a dendrite (A2). In both scenarios, each synapse is activated with 3 incoming spikes within a window of 30 milliseconds. (B1-B4) Somatic voltage, spine voltage, spine [Ca]NMDA and spine [Ca]L-type evoked in the linear integration scenario (randomly distributed synapses across the dendrites). Line color indicates the number of distributed synapses. In B2-B4, traces from a single spine are shown, which was randomly placed on a randomly chosen dendrite. (C1-C4) Somatic voltage, spine voltage, spine [Ca]NMDA and spine [Ca]L-type evoked in the supralinear integration scenario (synaptic cluster placed on one dendrite, in a 20-micrometer region approximately 120 micrometers away from the soma). Similarly, line color represents the size of the synaptic cluster, which is varied from 1 to 30 synapses. (D1-D4) The amplitude of the somatic voltage, spine voltage, spine [Ca]NMDA and [Ca]L-type compared between the cases of linear and supralinear integration. Results are averages over 20 trials, and in the case of synaptic clusters, over 8 different dendrites with 20 trials per dendrite (clusters located in a 20-micrometer in a dendritic region starting at approximately between 120 micrometers from the soma).

On the other hand, when synapses are clustered closely together on a dendrite, SPNs can perform supralinear integration of synaptic inputs through plateau potenitals (Du et al., 2017; Plotkin et al., 2011). Plateau potentials are evoked by activating synaptic NMDARs, and experimental studies have suggested that glutamate spillover, which activates extrasynaptic NMDARs, promotes plateaus or is even necessary to evoke them (Suzuki et al., 2008; Chalifoux and Carter, 2011; Oikonomou et al., 2012). We included glutamate spillover because it robustly provides an all-or-none jump in the (somatic) voltage amplitude (Trpevski et al., 2023). This all-or-none supralinearity is important for solving the NFBP (Tran-Van-Minh et al., 2015). Fig. 2C1 shows the somatic voltage elevations when varying the size of the synaptic cluster from 1 to 30 synapses, while Fig. 2C2 shows the spine voltage in one spine of the synaptic cluster. Figures 2C3, C4 show the spine [Ca]NMDA and [Ca]L-type that result from stimulating the synaptic cluster. As explained in the Methods section, calcium from these two different sources is tracked separately in the SPN model because they trigger a different plasticity outcome ([Ca]NMDA is needed for LTP and [Ca]L-type is needed for LTD). Also, [Ca]NMDA diffuses in the neuron, while [Ca]L-type does not, since it is assumed to be restricted to microdomains, a form of highly localized signaling often mediated by L-type calcium channels in neurons (Berridge, 2006; Parekh, 2008; Chen and Sabatini, 2012). As the cluster size is increased, increasingly larger and longer voltage elevations are produced as a result of activating more NMDARs (Figs. 2C1, C2, traces for cluster sizes up to 20 synapses). Once glutamate spillover occurs (here at a cluster size above 20 synapses), an all-or-none plateau potential appears, causing an all-or-none jump in the calcium concentration, as well (Figs. 2C1-C4).

This all-or-none behavior is also visible as a jump in the amplitudes of the somatic and spine voltage, and the spine [Ca]NMDA and [Ca]L-type in Figs. 2D1-D4. The somatic and spine plateau amplitude shows roughly linear increases before and after glutamate spillover, with the largest nonlinearity being caused by the plateau (Figs. 2D1-D2, red lines). The amplitudes of [Ca]NMDA and [Ca]L-type before spillover reflect the voltage dependencies of NMDARs and L-type channels (Figs. 2D3-D4, red lines for up to 20 synapses), and the largest nonlinearity is again caused by the plateau. The jump in all four quantities in Fig. 2D is an indication of a “threshold supralinearity”. When a critical level of stimulation is achieved, the neuron’s somatic voltage amplitude is suddenly much larger, despite it being stimulated with only one additional synapse in the synaptic cluster. This type of supralinear behavior should in principle be sufficient to solve the NFBP (Tran-Van-Minh et al., 2015).

Also, the comparison of linear and supralinear integration in Fig. 2D1 shows that, once a plateau appears, the same number of synapses trigger a much larger somatic voltage elevation when they are clustered compared to when they are randomly distributed across the dendrites. (In the somatic voltage there is also a small difference between linear and supralinear integration when the number of stimulated synapses is less than 10. This arises because the synaptic clusters are always placed at approximately the same distance from the soma, while the distributed synapses can be randomly placed further away.)

Synaptic plasticity rule with metaplasticity solves both the FBP and NFBP

Before demonstrating the role of metaplasticity using the FBP and NFBP, a learning rule that can solve these two tasks is needed. The synaptic plasticity rule is simple, and is given with Eq. 1. The synaptic weight w is increased if, during pattern presentation, [Ca]NMDA reaches a level above the LTP threshold, θLTP, and a dopamine peak arrives afterwards. Similarly, the weight is decreased if [Ca]L-type goes above the LTD threshold, θLTD, and a dopamine pause arrives afterwards. ([Ca]NMDA and [Ca]L-type are tracked separately in the SPN model, since they produce a different plasticity outcome. Also, as mentioned above, [Ca]NMDA diffuses in the neuron, while [Ca]L-type is assumed to be restricted to microdomains, a form of highly localized signaling often mediated by L-type calcium channels in neurons (Berridge, 2006; Parekh, 2008; Chen and Sabatini, 2012).) Synapses can be increased to a maximal level of wmax, and decreased to a minimal level of wmin, and weight updates are proportional to the current weight value. This formulation is very similar to existing plasticity rules (Rubin et al., 2001; Kistler, 2002; Gütig et al., 2003). The weight updates are also proportional to a learning rate η, and the time step for advancing the simulation, Δt. (Note that Δt is neither the timing between pre- and postsynaptic spikes as in spike-timing-dependent rules – this rule does not depend on somatic spiking, nor is it the time until dopamine feedback is provided. It is simply the time step for numerical integration.) This rule is local, meaning that it depends on calcium levels locally at the synapse, and somatic spiking is not necessary for plasticity (similarly to Khodadadi et al. (2025)). This is beneficial for distal synapses, as backpropagating action potentials in general do not affect the local voltage which triggers calcium influx at distant dendritic locations.

Also, at least in hippocampal CA3 pyramidal neurons, it has been reported that local dendritic depolarizations are necessary for plasticity, and somatic spiking is not, consistent with the fact that local calcium levels, driven by local voltage, are the signals triggering plasticity (Brandalise et al., 2016). Similarly, in hippocampal CA1 neurons, local dendritic activity is sufficient for synaptic plasticity (Bittner et al., 2017). (1) Δw=η{(wmax−w)Δt,[Ca2+]NMDA>θLTPandDA↑−(w−wmin)Δt,[Ca2+]L−type>θLTDand DA↓0,basal DA level Metaplasticity is also implemented in a simple way, given with Eq. 2. In most of the article, it only occurs if synaptic plasticity also occurs. During one update step, each threshold moves closer to the amplitude (maximum) of the calcium level evoked by an arriving pattern with a rate ηθ (θLTP follows the amplitude of [Ca]NMDA, and θLTD follows the amplitude of [Ca]L-type). This formulation of metaplasticity is very similar to the sliding threshold in the well-studied BCM learning rule, where the threshold follows a quantity derived from synaptic activity (Bienenstock et al., 1982). When a synapse is modified, both thresholds are updated. (2) ΔθLTP=ηθ([Ca2+]NMDAmax−θLTP)Δt,ΔθLTD=ηθ([Ca2+]L-typemax−θLTD)Δt. When using supralinear integration of inputs, we have added an upper threshold for LTP, ΘLTP (Eq. 3). This threshold serves to stop synaptic strengthening once synapses are strong enough to evoke a plateau. With it, synapses are not strengthened to the maximal value, wmax, thus avoiding strong supralinear integration for the irrelevant patterns. (This is important for solving the NFBP, as will be seen below.) The upper threshold for LTP only plays a role in synaptic plasticity, and is not taken into account in metaplasticity. This means that metaplasticity is performed in the same way as in Eq. 2, and under the conditions in Eq. 1. (3) Δw=η{(wmax−w)Δt,θLTP<[Ca2+]NMDA<ΘLTPand DA↑−(w−wmin)Δt,[Ca2+]L−type>θLTDand DA↓0,basal DA level Such “thresholded” metaplasticity, which requires the same conditions to occur as synaptic plasticity, cannot solve reversal learning, a task used to demonstrate the plasticity–stability dilemma. Because of this, when simulating reversal learning, we relaxed the conditions for metaplasticity so that it can be triggered by any calcium level in activated synapses (i.e. no need for calcium to be above a threshold as in Eq. 1).

In summary, each synapse is updated based on its local calcium levels, and has its own LTP and LTD thresholds. During learning, patterns in the FBP and NFBP are presented in random order, and after some time, each pattern is followed by a dopamine feedback signal (a dopamine peak or pause), during which plasticity occurs (Fig. 3A). (The intervals until reward delivery and the next pattern were chosen to shorten simulations; behavioral timescales would likely be longer; see the Methods section for more details.) The learning rule is always active (always “on”), avoiding separate training and testing phases sometimes used in other plasticity rules.

Learning outcome on the FBP and NFBP.

(A) Learning protocol. Patterns are presented in random order, with each pattern consisting of synaptic inputs arriving within a 30 ms window (3 spikes per synapse). Synaptic updates occur during a 50 ms dopamine signal delivered 400 ms after pattern onset; the next pattern is presented 600 ms after pattern onset (see Methods). (B) Illustration of the setups for the FBP and NFBP. (B1) FBP with linear integration: each feature is represented by 15 synapses distributed across the dendrites. (B2) FBP with supralinear integration: 10 synapses per feature, clustered in one dendrite. (B3) NFBP (only supralinear integration): 10 clustered synapses per feature, and clusters are located on two different dendrites. (C) The somatic and dendritic voltages in the FBP with linear (C1) and supralinear (C2) integration, and the NFBP (C3), before and after learning (gray and black traces, respectively), elicited by the relevant and irrelevant pattern(s). (D) The performance on the three tasks. Lines indicate score averaged over consecutive blocks of 20 patterns; shaded areas indicate standard deviation, and dashed lines indicate the threshold scores for solving the FBP (75%) and NFBP (87.5%) (see Methods for what these threshold scores represent). In the FBP with linear (D1) and supralinear (D2) integration, scores are averaged over 50 trials. In (D2) the cluster location is randomized across 8 dendrites in each trial; the cluster is placed in a 20-micrometer dendritic region, approximately 120 micrometers away from the soma. In (D3), the score is averaged over 12 trials per input configuration, divided into two groups (Figure 3–figure supplement 1), resulting in 216 and 156 trials per group, respectively. In each trial, two dendrites are randomly selected from 8 dendrites to place the clusters, at the same distance as in (D2). Figure 3—figure supplement 1. All input configurations used in the NFBP.

Before studying the effects of each calcium threshold on learning, we first show that the synaptic plasticity and metaplasticity rule can solve the FBP and NFBP (Fig. 3). For the FBP we have tested both linear integration of randomly distributed synaptic inputs (Figs. 3B1-D1) and supralinear integration of clustered synapses (Figs. 3B2-D2), while for the NFBP we tested only supralinear integration of clustered synapses in two different dendrites (Figs. 3B3-D3), since only supralinear integration can solve the NFBP (Tran-Van-Minh et al., 2015). Figs. 3C1-C3 show the voltage before and after learning for each pattern in the three scenarios. The SPN is silent before learning, and at the end of learning it has learned to spike to the relevant pattern(s). With supralinear integration, it does this by learning to evoke plateau potentials for the relevant pattern(s). For example, after learning (black traces), ‘red strawberry’ in Figs. 3C2 and 3C3 elicits a plateau in dendrite 1, which triggers somatic spiking. For the NFBP, the two relevant patterns evoke plateaus on separate dendrites (‘red strawberry’ on dendrite 1 and ‘yellow banana’ on dendrite 2, Fig. 3C3). Figs. 3D1, D2 show the performance on the FBP with linear and supralinear integration averaged over 50 trials, indicating that both modes of integration can solve the task. For the NFBP, in addition to the input configuration in Fig. 3B3, with three features on two dendrites, we also tested all possible configurations of four features on two dendrites which allow storage of the relevant patterns on separate dendrites. (These input configurations are listed in Figure 3–figure supplement 1.) Fig. 3D3 shows that when up to three features are present in one dendrite, the NFBP is successfully solved. When four features are present in one or both dendrites, usually only one relevant pattern is learned. Solving the NFBP in this case requires additional mechanisms such as branch plasticity or inhibition (Legenstein and Maass, 2011; Trpevski et al., 2026).

Before separately examining the effects of each calcium threshold on learning, it is first necessary to understand how the learning rule successfully solves the tasks by modifying the synaptic weights and thresholds. Figure 4 shows how the synaptic weights and the two calcium thresholds change during learning: they show the same qualitative behavior for both the FBP and NFBP. For the FBP, the synapses carrying the features ‘red’ and ‘strawberry’, which together form the relevant pattern, are strengthened (Figs. 4B1, B2). The synapses for ‘yellow’ only weaken, since they are only activated as part of the irrelevant pattern (which is followed by a dopamine pause, a signal for LTD). For the NFBP, where each relevant pattern is stored in a separate dendrite, the outcome is similar: the synapses for ‘red’ and ‘strawberry’ on dendrite 1 and the synapses for ‘yellow’ and ‘banana’ on dendrite 2 are strengthened (Figs. 4B3, B4). The remaining synaptic cluster, belonging to one irrelevant pattern in each dendrite, is weakened (the synapses for ‘yellow’ on dendrite 1 and the synapses for ‘strawberry’ on dendrite 2). In all cases, weakening stops when the evoked [Ca]L-type falls below the LTD threshold. For the FBP with linear integration at the soma, many of the randomly distributed synapses for the relevant pattern are strengthened almost to the maximal value wmax (Fig. 4B1). On the other hand, for the FBP with supralinear integration and the NFBP, the clustered synapses for the relevant pattern(s) are not strengthened to their maximal values. Instead, they stabilize once they are strong enough to evoke a plateau (Figs. 4B2–B4). This is due to the upper threshold for LTP, ΘLTP, shown with dashed lines in Figs. 4C2–C4. (The stabilization takes about twice as long in the NFBP simply because there are four patterns in the task, i.e. twice as many as in the FBP.) When plateaus are evoked by the relevant patterns, they drive somatic spiking (as indicated above in Figs. 3C2, C3). In contrast, an irrelevant pattern stimulates only half of the strengthened synapses on one dendrite, which are the synapses representing the shared feature with the relevant pattern (the ‘strawberry’ feature in dendrite 1 in Figs. 4B2, B3 and the ‘yellow’ feature in dendrite 2 in Fig. 4B4). This is not enough for a plateau potential, and even though a supralinear voltage elevation is elicited, no somatic spiking is triggered (Fig. 3C2, ‘yellow strawberry’ on dendrite 1 for the FBP and Fig. 3C3, ‘yellow strawberry’ on dendrite 1 and ‘red banana’ on dendrite 2 for the NFBP). Nevertheless, these voltage elevations produce high calcium of both types, but, as will be seen below, the shared strengthened synapses are protected from weakening as a result of metaplasticity. Finally, we also show that the NFBP cannot be solved by linear integration – with distributed synapses only one relevant pattern can be stored by the neuron (Figure 4–figure supplement 1).

The synaptic weights and calcium thresholds during learning of the FBP and NFBP.

(A) Illustrations of the setups for the FBP with linear (A1) and supralinear (A2) integration, and the NFBP (A3) (same as in Fig. 3). (B-D) The evolution of synaptic weights (B), LTP thresholds (C) and LTD thresholds (D) in the FBP with linear (B1-D1) and supralinear (B2-D2) integration, and the NFBP (B3-D3 and B4-D4). In (B) all synaptic weights for the features are shown, while in (C) and (D) only one synapse per feature is chosen to show its calcium threshold (solid lines). Dashed lines in (C2–C4) represent the upper threshold for plasticity ΘLTP. Dots in (C) and (D) represent the amplitudes of [Ca]NMDA and [Ca]L-type, respectively, during pattern presentation. For clarity, the calcium amplitudes for some patterns are omitted. (C) and (D) are shown in larger scale in Figure 4–figure supplement 2, with much more detail in the calcium amplitudes. Arrows show the moment in time when the weakened synapses become low enough for their calcium levels to mostly stay below the calcium thresholds. Figure 4—figure supplement 1. The NFBP cannot be solved with linear integration at the soma. Figure 4—figure supplement 2. Calcium thresholds and calcium amplitudes from Fig. 4 in greater detail. Figure 4—figure supplement 3. The effect of having no upper LTP threshold on supralinear integration. Figure 4—figure supplement 4. A parameter scan of wmax using the rule without an upper LTP threshold. Figure 4—figure supplement 5. The maximal weight wmax prevents unbounded weight growth. Figure 4—figure supplement 6. The effect of the learning rate on learning. Figure 4—figure supplement 7. The effect of a very low learning rate on learning (η = 0.1). Figure 4—figure supplement 8. The effect of a very high learning rate on learning (η = 20). Figure 4—figure supplement 9. The effect of the metaplasticity rate on learning. Figure 4—figure supplement 10. The effect of a low metaplasticity rate on learning (ηθ = 0.2). Figure 4—figure supplement 11. The effect of a high metaplasticity rate on learning (ηθ = 20).

The LTP and LTD thresholds for just one synapse from each feature are shown in Figs. 4C and 4D, respectively (the same plots are shown with larger panels in Figure 4–figure supplement 2 for better clarity). As mentioned above, they show the same behavior for both the FBP and NFBP, the only difference being that they stabilize at different values due to the different calcium levels evoked by distributed and clustered synaptic inputs. Also, the LTP and LTD thresholds show the same behavior when compared to each other. For example, looking at the strengthened synapses first, they increase to follow the calcium amplitudes evoked by each pattern presentation, and stabilize at these values towards the end of the simulation (thresholds for ‘red’ and ‘strawberry’ tend to the dots of the same color above them in Figs. 4C1-C3 and 4D1-D3, and the thresholds for ‘yellow’ and ‘banana’ tend to the dots of the same color above them in Figs. 4C4, D4; this is better visible in Figure 4–figure supplement 2 and also highlighted with the light blue regions in Figure 4–figure supplement 2B3). In the case of linear integration they stabilize at low values, because distributed synapses do not have cooperative voltage effects (due to the large distances between them), and low calcium levels are evoked as a result (Figs. 4C1, D1). Instead, with supralinear integration, the plateaus evoked by the clustered synapses cause much larger voltage and calcium elevations, so the thresholds rise to much higher levels (Figs. 4C2-C4 and 4D2-D4). Note that the LTP threshold θLTP can also go above the upper plasticity threshold ΘLTP (θLTP and ΘLTP are independent), reflecting the assumption that different molecular circuitry implements these two thresholds. (For example, experiments in hippocampal CA1 pyramidal neurons have demonstrated that the two different voltage thresholds for LTD and LTP can reverse their relative positions as a result of metaplasticity: the voltage threshold for LTD can rise to -20 mV, going above the voltage threshold for LTP, which reduced to -30 mV (Ngezahayo et al., 2000)).

We next look at the thresholds for the weakened synapses (‘yellow’ synapse in Figs. 4C1-C3 and 4D1-D3 and ‘strawberry’ synapse in Figs. 4C4, D4). At the beginning, the thresholds increase. Later, once the synapses are weakened enough so that the evoked [Ca]L-type falls below the LTD threshold, they reach a stable level. For example, in Figs. 4C2, C3 and 4D2, D3, at the beginning of learning, yellow dots can be seen above the ‘yellow’ thresholds, while later they almost always remain below the thresholds. Similarly, in Figs. 4C4, D4, light red dots are at first seen above the ‘strawberry’ thresholds, while later they are found below them. This transition of the [Ca]L-type amplitudes from above to below the LTD thresholds is marked with the arrows in Figs. 4B2-4-D2-4, corresponding to the moment in time where synapses are weakened enough. (This is more clearly visible in Figure 4–figure supplement 2, and it is also highlighted with the green and brown regions in Figure 4–figure supplement 2B3, and the pink region where yellow dots are absent.) We mention that in the FBP, where the weakened synapses are only activated by the irrelevant pattern, both of their thresholds are only updated during LTD. On the other hand, in the NFBP the weakened synapses and their thresholds can be updated during both LTP and LTD, because these synapses are also activated by a relevant pattern (i.e. are shared with a relevant pattern). (However, in Fig. 4 this only happens at the very beginning, when the weakened synapses also undergo LTP. As they are being weakened, their LTP threshold is raised, so [Ca]NMDA amplitudes later do not go above the LTP threshold.) In summary, once a calcium amplitude falls below its threshold, plasticity and metaplasticity cannot be triggered, and the thresholds, as well as the synapses, are not updated anymore. In other words, once the thresholds reach the evoked calcium amplitudes, synapses are stabilized. In effect, metaplasticity acts as a lock on the synapses: synapses are flexible for learning in the time it takes for the thresholds to reach the calcium amplitudes, and are stabilized once that occurs. Accordingly, we have assumed that at the beginning of learning thresholds start at a low value, so that synapses are initially flexible to store information. All of this will be much more evident in the next section, where we study the role of metaplasticity.

We stress once again that in the FBP with supralinear integration, one of the strengthened synaptic clusters belongs to the shared feature (the ‘strawberry’ cluster in Figs. 4A2, B2), and in the NFBP both strengthened synaptic clusters belong to shared features (the ‘red’ and ‘strawberry’ clusters in dendrite 1 in Figs. 4A3, B3, and the the ‘yellow’ and ‘banana’ clusters in dendrite 2 in Figs. 4A3, B4). Importantly, the LTD threshold for these synapses is raised because of the plateaus (Figs. 4D2–D4). Because of this, the high [Ca]L-type levels in these synapses evoked from an irrelevant pattern are below the LTD threshold and are not enough to weaken them (e.g. light red dots in Figs. 4D2, D3 just above the ‘yellow’ threshold after 100 seconds are far below their ‘strawberry’ threshold, and dark red dots are far below their ‘red’ threshold in Fig. 4D3; similarly yellow and dark yellow dots in Fig. 4D4 located just above the ‘strawberry’ threshold after 100 seconds are far below their ‘yellow’ and ‘banana’ thresholds, respectively; dots are also highlighted in the pink region of Figure 4–figure supplement 2B3). In this way, these strengthened synapses belonging to a shared feature are protected from weakening, allowing them to store one part of the relevant pattern. This effect will also be much more evident in the next sections, where we look at the effects of metaplasticity on learning.

The effects of parameters in the learning rule

We have also examined the effects of parameters in the learning rule. In supralinear integration, the upper plasticity threshold ΘLTP prevents the strengthened synapses from saturating at their maximal levels. Without it, the FBP is still solved, even though weights attain their maximal values (Figure 4–figure supplements 3B1 and 3E1). It is successfully solved because the features innervate only one dendrite, and the saturated weights do not cause somatic spiking when activated by the irrelevant pattern (Figure 4–figure supplement 3A1). However, the NFBP is not solved without an upper threshold ΘLTP, because the features innervate two dendrites, and the maximally strengthened synapses cause supralinear voltage elevations in the dendrites for the irrelevant patterns, which trigger somatic spiking (Figure 4–figure supplements 3A2, 3B2, 3B3 and 3E2). The maximally strengthened synapses also drive more weakening in the weakened synapses, compared to when an upper threshold is present (Figure 4–figure supplement 3B, cf. weakened weights in Fig. 4B2– B4). This is due to the elevated [Ca]L-type evoked when an irrelevant pattern activates the maximally strengthened synapses (shared “strawberry” synapses in Figure 4–figure supplements 3B1, B2 and shared “yellow” synapses in Figure 4–figure supplement 3B3). The elevated [Ca]L-type for the irrelevant patterns is also visible in the elevated LTD thresholds for the weakened synapses (“yellow” synapse in Figure 4–figure supplement 3D1, D2 and “strawberry” synapse in Figure 4–figure supplement 3D3, cf. Figs. 4D2–D4). (The LTP thresholds for the weakened synapses are also elevated, due to higher [Ca]NMDA produced by the maximally strengthened synapses. This is visible in Figure 4–figure supplement 3C when compared to Figs. 4C2–C4).

The upper threshold ΘLTP is necessary to solve the FBP and NFBP because there is no single value of the maximal weight wmax with which both tasks are solved simultaneously, as can be seen from the parameter scan for wmax in Figure 4–figure supplement 4 done in the absence of ΘLTP. The FBP with supralinear integration and the NFBP are first solved for wmax=1.3w¯0 (due to the wmax threshold for glutamate spillover being between 1.2w¯0 and 1.3w¯0 in this study, and spillover evoking plateaus), but this value is not high enough to solve the FBP with linear integration (Figure 4–figure supplement 4, second column). A value of wmax=1.5w¯0 is needed to solve the FBP with linear integration, which, on the other hand, is too high to solve the NFBP (the performance on the NFBP drops to the threshold score for solving it in Figure 4–figure supplement 4, fourth column). All other values for wmax>1.5w¯0 increase the performance on the FBP with linear integration, and cause faster learning of the FBP with supralinear integration, but lower the performance on the NFBP due to spiking for the irrelevant patterns, as shown previously in Figure 4–figure supplement 3A2. This suggests that if a range of wmax values exists that solves all three tasks, it is very narrow, and such a solution is not robust to variations in wmax.

The maximal weight wmax prevents unbounded weight growth, ensuring stability of the neuron’s firing rate in the FBP with linear integration (Figure 4–figure supplement 5). Without wmax, the weights grow until [Ca]NMDA saturates in the spines (which does not occur for most synapses within the simulation time in Figure 4–figure supplement 5B; also the weights grow to unphysiological values, which of course does not happen in reality). Thus, in the FBP with linear integration, the neuron enters depolarization block when stimulated by a pattern after learning (Figure 4–figure supplement 5A1). On the other hand, in the FBP with supralinear integration, firing rate stability is ensured by the plateau potentials, since a plateau’s voltage is limited by the NMDAR reversal potential, regardless of how high the synaptic weights are (Figure 4–figure supplement 5A2). This is part of the plateaus’ function to provide dynamic range compression, as shown experimentally in Figs. 3E, 4D and 12 in Oikonomou et al. (2012).

The learning rate η has an effect on the speed of learning – more pattern presentations are required with lower learning rates, and vice versa (Figure 4–figure supplement 6). However, if the learning rate η is very low or very high, the NFBP cannot be solved: a very low η causes spiking for the irrelevant patterns because synapses are not weakened enough (Figure 4–figure supplement 7), and a very high η causes large updates to the synaptic weights, disrupting learning due to the similarly large changes in calcium, which can easily fall below the thresholds (Figure 4–figure supplement 8). Within a certain range, the rate of threshold adaptation (metaplasticity rate) ηθ has no significant effect on learning to solve the FBP and NFBP (Figure 4–figure supplement 9). If the metaplasticity rate is too low, it takes a long time for the thresholds to reach the calcium amplitudes, and as a result, synapses take a longer time to stabilize in the FBP with linear integration and in the NFBP (Figure 4–figure supplement 10B1, 10B3, 10B4). Synapses stabilize quickly in the FBP with supralinear integration because it is sufficient that one of the clusters is strengthened enough to trigger plateaus (Figure 4–figure supplement 10B2). Learning is slowed down in the NFBP – more patterns are needed before the same performance level is reached (Figure 4–figure supplement 10E3, cf. Fig. 3D3), while learning in the FBP is unaffected (Figure 4–figure supplements 10E1, E2, cf. Figs. 3D1, D2). On the other hand, if the metaplasticity rate is too high, the thresholds reach the respective calcium levels quickly, shortening the period where synapses are flexible and preventing them from being changed very much, i.e. they are quickly stabilized (Figure 4–figure supplement 11B–D). This results in more variable performance on the FBP with linear integration because weights are not strengthened enough (Figure 4–figure supplement 11E1), and in reduced performance on the NFBP because weakened clusters are not weakened enough, causing spiking for the irrelevant patterns (Figure 4–figure supplement 11E3).

Metaplasticity in the LTD threshold prevents weakening of strengthened synapses, protecting learned patterns

The role of each threshold is easily examined with this plasticity and metaplasticity learning rule by turning metaplasticity off, i.e. fixing a threshold’s value during learning. As mentioned in the introduction, the FBP and NFBP are particularly useful for demonstrating the role of each threshold because the patterns in the tasks share features. In the FBP, the relevant and irrelevant pattern (the red and yellow strawberry) share the shape feature (the feature “strawberry”). This means that during learning, the synapses representing this feature experience both strengthening and weakening (e.g. visible at the beginning of learning in Figs. 4B1, B2), and that strengthening is somehow “chosen”. Fixing the LTD threshold (to the initial low value in the simulation) shows that the shared synapses cannot strengthen properly, i.e. cannot stabilize (Figs. 5B1, B2). This happens because there is no metaplasticity in the LTD threshold – it is fixed, i.e. it never rises to follow [Ca]L-type. As a result, whenever an irrelevant pattern arrives, it always causes higher [Ca]L-type than the fixed LTD threshold, and can always weaken the shared synapses, despite them also being strengthened by the relevant pattern (Figs. 5D1, D2 – LTD thresholds stay fixed during learning, and [Ca]L-type is always above θLTD). The other two features are not shared – the feature “red” is only activated by the relevant pattern and, the feature “yellow” only by the irrelevant pattern. As a result, they can only ever experience plasticity in one direction: the feature “red” is strengthened, and the feature “yellow” is weakened. The fixed LTD threshold has no influence on the strengthened feature (“red”). However, in the FBP with supralinear integration, where the cooperative voltage effects of the clustered synapses produce higher calcium elevations, it does influence the weakened feature (“yellow”): because the LTD threshold is fixed to a low value, the evoked [Ca]L-type is high enough to decrease the weakened synapses much more compared to when metaplasticity in the LTD threshold is active (Figs. 5B1, B2, cf. “yellow” synapses in Figs. 4B1, B2).

Learning outcome on the FBP and NFBP without metaplasticity in the LTD threshold.

(A) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for the FBP with linear (A1) and supralinear integration (A2) and the NFBP (A3). (B–D) The evolution of synaptic weights (B), LTP thresholds (C) and LTD thresholds (D). (D) shows the fixed LTD thresholds. In (B) all synapses are shown, and in (C, D) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (C) and [Ca]L-type (D) during pattern presentation, omitting some patterns for clarity. Dashed line in (C2) shows the upper threshold for LTP, ΘLTP. (E) The performance on the FBP with linear (E1) and supralinear integration (E2) and the NFBP (E3) without metaplasticity in the LTD threshold. Performance on the FBP is averaged over 50 trials, and in the NFBP over 12 trials per input configuration (divided into the two groups shown in Figure 3–figure supplement 1), with 216 and 156 trials, respectively).

Regarding the performance, in the FBP with linear integration, despite the shared synapses not stabilizing, the strengthening is enough to drive somatic spiking (Fig. 5A1), and the task is solved but with a lower performance (Fig. 5E1). On the other hand, in the FBP with supralinear integration, because the shared “strawberry” synapses do not stabilize, the clustered synapses are not strong enough to evoke plateaus (Fig. 5A2), and the FBP is not learned (Fig. 5E2).

The effect of the fixed LTD threshold on the NFBP is analogous, but even more prominent because in the NFBP, a relevant pattern shares its features with both irrelevant patterns (e.g. “red strawberry” shares the color feature with “red banana” and the shape feature with “yellow strawberry”). This means that during learning, both features of a relevant pattern experience strengthening and weakening. Hence, when the LTD threshold is fixed, none of their synapses stabilize, and the relevant patterns are not stored in the dendrites (neither “red” and “strawberry” in dendrite 1, Fig. 5B3, nor “yellow” and “banana” in dendrite 2, Fig. 5B4). As a result, they cannot evoke plateaus, and the NFBP is not learned (Figs. 5A3, E3). The LTP threshold follows the level of [Ca]NMDA and also does not stabilize (Figs. 5C3, C4). As with the FBP, this happens because there is no metaplasticity in the LTD threshold, and irrelevant patterns always cause higher [Ca]L-type than the fixed LTD threshold (Figs. 5D3, D4). This means that despite being strengthened, these synapses will always get weakened. Conversely, when metaplasticity in the LTD threshold is active, it follows the increased levels of [Ca]L-type elicited by a plateau (evoked due to prior synaptic strengthening, as shown previously in Figs. 4D2–D4). This elevated LTD threshold protects the strengthened synapses from weakening when they are activated by an irrelevant pattern. As mentioned above, they are protected because an irrelevant pattern, activating only half of the strengthened synapses, does not evoke a plateau, resulting in [Ca]L-type lower than the elevated LTD threshold in Figs. 4D2–D4 (e.g. light red dots in pink region in Figure 4–figure supplement 2B3 are below the elevated “strawberry” threshold).

Without metaplasticity in the LTD threshold, strengthened synapses are not protected from weakening. Also, because the LTD threshold is fixed to a low value, the synapses belonging to the weakened feature in each dendrite are almost weakened to the minimal value wmin. In summary, by protecting synapses from weakening, the threshold for weakening allows proper synaptic strengthening.

Metaplasticity in the LTP threshold prevents strengthening of weakened synapses, promoting the weakening of unnecessary synapses

On the other hand, the LTP threshold has an opposite effect on learning. This effect can be seen in a shared feature which ultimately needs to be weakened, but which experiences a drive to both strengthen and weaken during learning. There is no such feature in the FBP (the only shared feature, “strawberry”, needs to be strengthened), so fixing the LTP threshold has no effect on solving this task (Figs. 6A1–E1 and 6A2–E2, cf. Figs. 3A1,2–C1,2, 4A1,2–D1,2). Nevertheless, there are such features in the NFBP, which is why it is a suitable task to demonstrate the role of the LTP threshold. In fact, each irrelevant pattern in the NFBP also shares its two features with the relevant patterns, and, as a result, they also experience both strengthening and weakening during learning (e.g. the features “yellow” in dendrite 1 and and “strawberry” in dendrite 2 in Fig. 4A3).

Learning outcome on the FBP and NFBP without metaplasticity in the LTP threshold.

(A) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for the FBP with linear (A1) and supralinear integration (A2) and the NFBP (A3). (B–D) The evolution of synaptic weights (B), LTP thresholds (C) and LTD thresholds (D) for the synapses in each dendrite. (C) shows the fixed LTP thresholds, and the dashed lines show the upper LTP threshold, ΘLTP. In (B) all synapses are shown, and in (C, D) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (C) and [Ca]L-type (D) during pattern presentation, omitting some patterns for clarity. Dashed lines in (C2–C4) show the upper threshold for LTP, ΘLTP. (E) The performance on the FBP with linear (E1) and supralinear integration (E2) and the NFBP (E3) without metaplasticity in the LTP threshold. Performance on the FBP is averaged over 50 trials, and in the NFBP over 12 trials per input configuration (divided into the two groups shown in Figure 3–figure supplement 1, with 216 and 156 trials, respectively).

If the LTP threshold is fixed, the features belonging to the irrelevant pattern in each dendrite cannot be weakened properly (“yellow” synapses in dendrite 1 in Fig. 6B3 and “strawberry” synapses in dendrite 2 in Fig. 6B4). Because of the low value of the LTP threshold, whenever an irrelevant pattern activates these synapses, the evoked [Ca]NMDA is still greater than the threshold (Figs. 6C3, C4), triggering synaptic strengthening even if synapses have previously been weakened. This means that when metaplasticity in the LTP threshold is active, by following the levels of [Ca]NMDA (e.g. LTP threshold for “yellow” in Fig. 4C3 and for “strawberry” in Fig. 4C4), it prevents these synapses from strengthening when they are activated by the relevant pattern (which can only be stored fully in the other dendrite). As a result, the synapses for “yellow” in Fig. 4B3 and for “strawberry” in Fig. 4B4 are successfully weakened, while in Figs. 6B3, B4 they are not. (The LTD threshold behaves as usual for the strengthened synapses, protecting them from weakening, as shown in Figs. 6B3, B4 and 6D3, D4. However, for the weakened synapses, it does not stabilize.) Without metaplasticity in the LTP threshold, the NFBP is on average not solved (Fig. 6E3). Despite both relevant patterns being stored by each dendrite, the performance is lowered because the SPN sometimes also spikes for the irrelevant patterns (Fig. 6A3). Also, in the input configurations containing four features in one dendrite, the average performance even drops below 50 % (green trace in Fig. 6E3). In summary, by preventing synapses from strengthening, the threshold for strengthening allows proper synaptic weakening.

Both thresholds need to be updated during one synaptic modification in order to solve the NFBP

The rule is formulated so that during one synaptic modification (e.g. LTP) the thresholds for both LTP and LTD are updated. Alternatively, it could be formulated so that during LTP, only the threshold for LTP is updated, and during LTD, only that for LTD is updated. We test this variant in this section, and the results from the previous two sections already give an intuition how this variant would work. This time we start with the NFBP, since that is the task where the role of both thresholds can be demonstrated. Updating only the corresponding threshold during one synaptic modification means that the other threshold is, in effect, kept fixed during that plasticity update, making metaplasticity only partially active. The effects are seen in Figs. 7C1, C2. Synapses are neither properly strengthened, nor weakened, at least within the length of the simulation. For example, one of the relevant patterns (“red strawberry”) shows a trend of strengthening, but is also being weakened (Fig. 7C1). This is because as the LTD thresholds for “red” and “strawberry” move higher (Fig. 7E1), LTD in these synapses is (sometimes) prevented, and a trend of synaptic strengthening appears. Also, as the LTP threshold for “yellow” increases (Fig. 7D1), it prevents the “yellow” synapses from strengthening, and as a result, they weaken to a degree.

Learning outcome on the NFBP with partial metaplasticity, where the LTP threshold is updated only during LTP, and the LTD threshold is updated only during LTD.

(A) An illustration of the setup for the NFBP (same as in Fig. 3B3). (B) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for all four patterns. (C-E) The evolution of synaptic weights (C), LTP thresholds (D) and LTD thresholds (E) for the synapses in each dendrite. In (C) all synapses are shown, and in (D, E) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (D) and [Ca]L-type (E) during pattern presentation, omitting some patterns for clarity. Dashed lines in (C2–C4) show the upper threshold for LTP, ΘLTP. (F) The performance on the NFBP with partial metaplasticity, averaged over 12 trials per input configuration (divided into the two groups shown in Figure 3–figure supplement 1, with 216 and 156 trials, respectively). Figure 7—figure supplement 1. The effect of partial metaplasticity on solving the FBP.

However, since the LTD threshold is not updated during LTP, it does not follow the [Ca]L-type levels generated by the relevant patterns. It is updated only during LTD, following the [Ca]L-type levels from irrelevant patterns. For example, in dendrite 1, the “red banana” activates only the cluster representing “red”, evoking little [Ca]L-type and, as a result, the LTD threshold for the “red” synapse is lower (Fig. 7E1). On the other hand, the “yellow strawberry” activates two synaptic clusters (Fig. 7A1), causing a larger voltage elevation and consequently larger [Ca]L-type, so the LTD thresholds for these synapses move higher (Fig. 7E1). However, because of the trend of synaptic strengthening in the “red” and “strawberry” synapses, the irrelevant patterns sometimes still evoke [Ca]L-type higher than the LTD thresholds. As a result, the synapses for “red” and “strawberry” in dendrite 1 are not protected from weakening (when activated by the irrelevant patterns). Also, as the LTD threshold for “yellow” increases in the first 70 s of the simulation (Fig. 7E1), the synapses for “yellow” show a trend of increasing in this period, as well, before being decreased afterwards (when the “yellow” LTP threshold is increased). On the other hand, in dendrite 2 the relevant pattern is not strengthened, and it is not clear whether more learning trials will cause both relevant patterns to be stored in the two dendrites. On average, the NFBP is not solved in this case (Fig. 7F).

Finally, this variant is similar to not having metaplasticity at all (Fig. 8). Without metaplasticity, no synapses stabilize, i.e. they are neither strengthened nor weakened (Fig. 8C1, C2).

Learning outcome on the NFBP without metaplasticity (the LTP and the LTD thresholds are fixed).

(A) An illustration of the setup for the NFBP (same as in Fig. 3B3). (B) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for all four patterns. (C) The evolution of synaptic weights in both dendrites. (D, E) The fixed LTP thresholds (D) and LTD thresholds (E) in each dendrite, with calcium amplitudes during pattern presentation shown with dots (omitting some patterns for clarity). (F) The performance on the NFBP without metaplasticity (same number of trials as in Fig. 7F. Figure 8—figure supplement 1. The effect of having no metaplasticity on solving the FBP.

For the FBP, the results in the previous two sections again give an intuition for how partial metaplasticity would work. As shown in Figs. 6A1–E1 and 6A2–E2, even when the LTP threshold is fixed, it does not have an effect on solving the FBP (because there are no shared features in the FBP which ultimately need to be weakened). Similarly, neither does it have an effect when it is partially active: the LTP threshold is merely less elevated than compared to when metaplasticity is always active (Figure 7–figure supplement 1C, cf. Figs. 4C1, C2). On the other hand, partial metaplasticity in the LTD threshold is not enough to protect the shared strengthened synapses from weakening. The LTD threshold does not follow [Ca]L-type fast enough (Figure 7–figure supplement 1D), and as a result the shared “strawberry” synapses are not strengthened properly (even though, for example, their weights stabilize when using supralinear integration, Figure 7–figure supplement 1B). Similarly to Figs. 5A1–E1 and 5A2–E2, this does not affect learning to solve the FBP with linear integration (Figure 7–figure supplement 1A1 and 1E1), but it does affect learning of the FBP with supralinear integration (Figure 7–figure supplement 1A2 and 1E2).

The effects of having no metaplaticity on the FBP with linear integration are similar to those of having partial metaplasticity (Figure 8–figure supplement 1). Again, the synapses encoding the shared feature cannot stabilize (synapses for “strawberry”, Figure 8–figure supplement 1B), only moderately affecting the performance (Figure 8–figure supplement 1A1 and 1E1). No metaplasticity in the FBP with supralinear integration also causes synapses encoding the shared to not stabilize (Figure 8–figure supplement 1A2). However, compared to partial metaplasticity, the unstable shared synapses sometimes evoke plateaus as they oscillate, on average solving the task (Figure 8–figure supplement 1E2). In summary, this shows that in the FBP, metaplasticity is important for robust, consistently high performance on the task.

Reversal learning can be solved when relaxing the conditions for metaplasticity or introducing a signal that resets the thresholds

We have so far demonstrated a situation where the calcium thresholds are initially set to a low value, making the synapses “flexible” for learning in the time it takes for the thresholds to reach the [Ca]NMDA and [Ca]L-type amplitudes. Once the thresholds have elevated and reach the calcium amplitudes, the synapses are effectively “locked”, providing stable information storage. Conversely, fixing the calcium thresholds to study metaplasticity (in Figs. 5 and 6) made the synapses permanently flexible for learning and no information could be stored. This illustrates the role of metaplasticity as a lock on the plasticity process, and that it plays a role in solving the plasticity–stability dilemma, which means that synapses should “both be stable enough to store information and be plastic enough to accommodate new learning” (Jedlicka et al., 2022). Reversal learning is a straightforward example of the plasticity–stability dilemma applied to the FBP and NFBP: after the tasks have been learned, we switch the reward policy so that relevant patterns become irrelevant (and are followed by dopamine pauses), and vice versa, irrelevant patterns become relevant (and are followed by dopamine peaks). To solve this task in the (N)FBP, the neuron should keep the learned changes in (one of the) the shared strengthened feature(s) (e.g. the shared “strawberry” is necessary for both the “red strawberry” and “yellow strawberry” patterns), and change the weights of the other features (e.g. “red” should be weakened and “yellow” should be strenghtened to reverse the neuron’s response from “red strawberry” to “yellow strawberry”).

We first show that “thresholded” metaplasticity cannot solve reversal learning (Fig. 9). The reward policy is reversed in the middle of the simulation, and all three tasks are solved before the reversal (blue voltage traces in Fig. 9A, and score in Fig. 9E before reversal). After reversal, because metaplasticity is only activated if a calcium signal is above its threshold (and followed by dopamine feedback), this means that the strengthened features cannot be weakened properly once their calcium levels fall below their LTD thresholds (strengthened synapses in Figs. 9B1–B4 weaken only slightly because their [Ca]L-type levels fall below the LTD thresholds after weakening, shown in Figs. 9D1–D4; this fall in [Ca]L-type is very prominent with supralinear integration because after weakening, no plateaus can be evoked). Neither can the weakened features be strengthened in most of the cases, because the previosuly irrelevant patterns, which are now rewarded and relevant, rarely cause [Ca]NMDA above the LTP thresholds (calcium dots in Figs. 9C1–C4 largely stay below the LTP threshold of the weakened synapses). The FBP and NFBP are successfully learned before the point of reward policy reversal, while after reversal, the synapses cannot adapt to the new situation. As a result, in the end there is no spiking for the initially irrelevant patterns, nor is there spiking for the initially relevant patterns that were learned before reversal (Figs. 9A1–A3, black traces, and Figs. 9E1–E3). As expected, “thresholded” metaplasticity permanently locks the patterns in the dendrites, making the rule inflexible to learn new pattern-outcome associations if the reward policy changes.

Reversal learning in the FBP and NFBP cannot be solved with “thresholded” metaplasticity.

(A) The dendritic and somatic voltage before learning (gray traces), before reversal (blue traces) and after reversal (black traces) for the FBP with linear (A1) and supralinear integration (A2) and the NFBP (A3). (B–D) The evolution of synaptic weights (B), LTP thresholds (C) and LTD thresholds (D) for the synapses in each dendrite. In the middle of the simulation, the reward policy is reversed. Dashed lines in (C) show the upper LTP threshold, ΘLTP. In (B) all synapses are shown, and in (C, D) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (C) and [Ca]L-type (D) during pattern presentation, omitting some patterns for clarity. Dashed lines in (C2–C4) show the upper threshold for LTP, ΘLTP. (E) The performance on reversal learning in the FBP with linear (E1) and supralinear integration (E2) and the NFBP (E3). Performance on the FBP is averaged over 50 trials, and in the NFBP over 12 trials per input configuration (divided into the two groups shown in Figure 3–figure supplement 1), with 216 and 156 trials, respectively). Figure 9—figure supplement 1. “Relaxed” metaplasticity also solves the FBP and NFBP. Figure 9—figure supplement 2. “Relaxed” metaplasticity can overcome initially high calcium thresholds, unlocking synapses and allowing learning.

To solve reversal learning, we demonstrate two different approaches. One is to allow the thresholds to be lowered again, i.e. prevent them from stabilizing at an elevated level forever. To do this, we simply relax the conditions for metaplasticity so that any calcium levels in the active synapses can cause metaplasticity, i.e. calcium no longer needs to be above a threshold level for metaplasticity to occur (note that those conditions still hold for synaptic plasticity, as in Eq. 1). This approach has been taken in an extension of the cascade model, which in its original form cannot solve reversal learning. (In the generalized model the threshold is allowed to move in both directions (Farashahi et al., 2017).) We first show that such “relaxed” metaplasticity can solve the FBP and NFBP (Figure 9–figure supplement 1). In this case, the thresholds always follow any calcium levels evoked when a synapse is activated by presynaptic input, allowing them to lower in value. When the variability in the calcium amplitudes is small, the fluctuations in the thresholds are also small, as occurs in the FBP with linear integration (Figure 9–figure supplement 1C1, D1), the strengthened “red” synaptic cluster in the FBP with supralinear integration (Figure 9–figure supplement 1C2, D2, dark red traces) and the weakened synaptic clusters in supralinear integration (Figure 9–figure supplement 1C3, 4, D3,4). Conversely, when the variability in the calcium amplitudes is large, as occurs in the shared synapses that need to be strengthened when using supralinear integration, the thresholds show large fluctuations (Figure 9–figure supplement 1B2–4–D2–4). Strictly speaking, with “relaxed” metaplasticity the thresholds never stabilize, which is particularly evident for the shared strengthened clustered synapses with large variability in the calcium amplitudes (Figure 9–figure supplement 1C2–4, D2–4). Despite the fluctuations, the thresholds perform their function equally well: the LTD thresholds are elevated enough to protect the strengthened synapses from weakening, and the LTP thresholds follow [Ca]NMDA in the the weakened synapses to prevent them from strengthening. Note that because of the relaxed metaplasticity conditions, the LTD thresholds for the weakened synapses are not elevated as with “thresholded” metaplasticity (Figure 9–figure supplement 1D2–D4, cf. Fig. 5D2–D4). Hence, the weakened synapses decrease almost to the minimal level wmin (Figure 9–figure supplement 1B2–B4). Finally, “relaxed” metaplasticity allows to overcome initially high calcium thresholds set before learning (Figure 9–figure supplement 2). With “thresholded” metaplasticity, if the thresholds are initialized to values higher than the calcium signals, the synapses will remain locked and no learning will occur (Figure 9–figure supplement 2B1, B2). “Relaxed” metaplasticity can instead adapt the thresholds, thus unlocking the synapses and allowing learning (Figure 9–figure supplement 2B3, B4).

Relaxing the conditions for metaplasticity allows the rule to solve reversal learning in both the FBP and the NFBP (Fig. 10). The key to solving it with this approach is that calcium thresholds which have been raised while learning the FBP and NFBP are free to be lowered again after reversing the reward policy. In the FBP, after reversal, the LTD thresholds for the features “red” and “strawberry” are lowered, so these two features are weakened (Figs. 10B1, B2 and 10D1, D2, these weights and thresholds are weakened immediately after reversal). Simultaneously, “yellow strawberry” evokes higher [Ca]NMDA than the “yellow” LTP threshold because in the initial stages of reversal the feature “strawberry” is still strong, despite being weakened (Figs. 10C1, C2, LTP threshold for “yellow” rises to follow the evoked [Ca]NMDA). When the feature “yellow” is strengthened enough, it elicits strong [Ca]NMDA together with the “strawberry” feature, causing the latter to strengthen again (Figs. 10B1, B2, “yellow” synapses increase after reversal, and “strawberry” synapses follow after some time). The LTD thresholds for “yellow” and “strawberry” become elevated and protect the two strengthened features from weakening (Figs. 10D1–D3). Conversely, the LTP thresholds for “red” follow [Ca]NMDA and prevent this feature from strengthening. In this way, in the FBP, the initially relevant pattern “red strawberry” stops evoking spiking (through linear integration at the soma or via a plateau), and the neuron starts responding to the initially irrelevant pattern “yellow strawberry”, thus solving the task (blue voltage traces before reversal learning and black traces after reversal learning in Fig. 10A1, A2, and performance scores in Fig. 10E1, E2).

Reversal learning in the FBP and NFBP is solved when the conditions for metaplasticity are relaxed.

(A) The dendritic and somatic voltage before learning (gray traces), before reversal (blue traces) and after reversal (black traces) for the FBP with linear (A1) and supralinear integration (A2) and the NFBP (A3). (B–D) The evolution of synaptic weights (B), LTP thresholds (C) and LTD thresholds (D) for the synapses in each dendrite. In the middle of the simulation, the reward policy is reversed. Dashed lines in (C) show the upper LTP threshold, ΘLTP. In (B) all synapses are shown, and in (C, D) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (C) and [Ca]L-type (D) during pattern presentation, omitting some patterns for clarity. (E) The performance on reversal learning in the FBP with linear (E1) and supralinear integration (E2) and the NFBP (E3). Performance on the FBP is averaged over 50 trials. In (E2), due to the thresholds changing slowly, it often occurs that after reward policy reversal, calcium never goes above the thresholds to trigger new learning, thus lowering the performance (blue trace). Increasing the metaplasticity rate to ηθ = 4 can solve this (brown trace). For the NFBP, 12 trials per input configuration are used (divided into the two groups shown in Figure 3–figure supplement 1, with 216 and 156 trials, respectively). Figure 10—figure supplement 1. All input configurations in reversal learning with the NFBP that can store one initially relevant pattern and both irrelevant patterns. Figure 10—figure supplement 2. Relaxed metaplasticity also solves the NFBP in input configurations that can store both (initially) irrelevant patterns. Figure 10—figure supplement 3. Reversal learning in the NFBP on one input configuration that can store both relevant and both irrelevant patterns.

The same occurs in the NFBP – the neuron stops responding to the two initially relevant patterns (Fig. 10A3, blue voltage traces before reversal show a plateau potential in each dendrite evoked by a relevant pattern, and black traces after reversal show no plateaus and no somatic spiking for these two patterns). However, since there is no structural plasticity, the available synaptic connectivity in this example (Fig. 3A3) allows for only one of the initially relevant patterns to be stored (in both dendrites, as seen from the synaptic weights in Figs. 10B3, B4; also, black voltage traces after reversal show plateaus in both dendrites for “yellow strawberry” in Fig. 10A3). This is also reflected in the performance, which shows that before reversing the reward policy, the NFBP is solved, while after the reversal the neuron responds to only one pattern (Figs. 10E3). To show that both initially irrelevant patterns can be learned in reversal learning, we also simulated the task on input configurations that are innervated by only one initially relevant pattern, but both irrelevant patterns (Figure 10–figure supplements 1 and 2). An example of such a configuration is given in Figure 10– figure supplement 2A, where only one initially relevant pattern (“red strawberry”) can be stored in both dendrites, but both initially irrelevant patterns (“yellow strawberry” and “red banana”) innervate dendrites 1 and 2, respectively. During learning, the neuron first learns to respond to “red strawberry”, and after reward policy reversal spikes to both “yellow strawberry” and “red banana”, elicited via plateau potentials in dendrites 1 and 2, respectively (Figure 10–figure supplement 2B, blue traces before reversal, and black traces after reversal, and Figure 10–figure supplement 2E, score of 75% before reversal, and close to 100% after reversal). Finally, there are a total of four input configurations in group 1 in Figure 3–figure supplement 1 and Figure 10–figure supplement 2 (marked with a star) which can store both relevant patterns and both irrelevant patterns. Reversal learning using (one of) these is shown in Figure 10–figure supplement 3, demonstrating that given the appropriate structural connectivity, reversal learning from both relevant to both irrelevant patterns can be done.

Another approach to solve reversal learning is to assume that another part of the brain detects the change in the reward policy and sends a signal to temporarily “reset” the permanently stabilized thresholds to a low value, thus “unlocking” the synapses, i.e. making the weights flexible again to relearn the new pattern-reward associations (Fig. 11). (This approach has also been taken in another extension of the original cascade model (Iigaya, 2016).) The noradrenergic system, particularly the locus coeruleus, has been implicated in detecting unexpected changes in environmental conditions and facilitating behavioral adaptation, including during reversal learning (Jordan, 2024; Sara, 2009; Pajkossy et al., 2023). The striatum receives only limited projections from the locus coeruleus, but SPNs express adrenergic receptors which modulate phosphorylation of DARPP-32, an important protein regulating cortico-striatal synaptic plasticity (Hara et al., 2010), and hence could be affected by such signals about environmental changes.

Threshold resetting solves reversal learning.

A signal which resets the thresholds is assumed to arrive from a different brain region detecting the change in reward policy. (A) The dendritic and somatic voltage before learning (gray traces), before reversal (blue traces) and after reversal (black traces) for the FBP with linear (A1) and supralinear integration (A2) and the NFBP (A3). (B–D) The evolution of synaptic weights (B), LTP thresholds (C) and LTD thresholds (D) for the synapses in each dendrite. In the middle of the simulation, the reward policy is reversed. Dashed lines in (C) show the upper LTP threshold, ΘLTP. In (B) all synapses are shown, and in (C, D) only one synapse per feature is used to show its calcium threshold. Dots represent the amplitudes of [Ca]NMDA (C) and [Ca]L-type (D) during pattern presentation, omitting some patterns for clarity. (E) The performance on reversal learning in the FBP with linear (E1) and supralinear integration (E2) and the NFBP (E3). Performance on the FBP is averaged over 50 trials, and in the NFBP over 12 trials per input configuration (divided into the two groups shown in Figure 3–figure supplement 1, with 216 and 156 trials, respectively).

Two calcium thresholds provide separate control over LTP and LTD

In this section we ask the question what is the use of two separate calcium thresholds for LTP and LTD? The FBP, NFBP and reversal learning can be solved just as well with a single plasticity threshold (Fig. 12 and Figure 12–figure supplements 1 and 2). To show this we simulated a scenario where calcium from both NMDA and L-type channels is tracked with a single variable, [Ca] (rather than two separate ones), and diffuses axially across the spines and dendrites, and the same [Ca] signal is responsible for LTP and LTD when it crosses a single calcium threshold (Eqs. 4 and 5). This is not biologically realistic in dSPNs, but allows us to compare the functionality of two thresholds versus just one. As before, the dopamine signal determines the direction of synaptic plasticity:

A single plasticity threshold is sufficient to solve the FBP and NFBP.

“Thresholded” metaplasticity is used here. (A) The dendritic and somatic voltage before (gray traces) and after learning (black traces) for the FBP with linear (A1) and supralinear integration (A2) and the NFBP (A3). (B, C) The evolution of synaptic weights (B) and the single calcium threshold (C) for the synapses in each dendrite. Dashed lines in (C) show the upper LTP threshold, ΘLTP. In (B) all synapses are shown, and in (C) only one synapse per feature is used to show its calcium threshold. Dots represent the [Ca] amplitudes during pattern presentation, omitting some patterns for clarity. Dashed lines in (C2–C4) show the upper threshold for LTP, ΘLTP. (D) The performance on the FBP with linear (D1) and supralinear integration (D2) and the NFBP (D3). Performance on the FBP is averaged over 50 trials, and in the NFBP over 12 trials per input configuration (divided into the two groups shown in Figure 3–figure supplement 1, with 216 and 156 trials, respectively). Figure 12—figure supplement 1. A single plasticity threshold is sufficient to solve the FBP and NFBP. when the conditions for metaplasticity are relaxed. Figure 12—figure supplement 2. A single plasticity threshold is sufficient to solve reversal learning when the conditions for metaplasticity are relaxed. Figure 12—figure supplement 3. Learning outcome without metaplasticity in the single threshold scenario (the threshold is fixed). Figure 12—figure supplement 4. [Ca]NMDA and [Ca]L-type are highly correlated and carry essentially the same information.

(4) Δw=η{(wmax−w)Δt,θ<[Ca2+]<ΘLTPandDA↑−(w−wmin)Δt,[Ca2+]>θandDA↓0,basal DA level, (5) Δθ=ηθ([Ca2+]max−θ)Δt, and we use both “thresholded” metaplasticity (θ is updated only when [Ca] goes above it, Fig. 12) and the relaxed condition for metaplasticity (θ is updated for any [Ca] level in active synapses, Figure 12–figure supplement 1). In both cases, the dynamics of the weights and thresholds are analogous to those when two separate calcium thresholds are used. As before, no metaplasticity (fixing the threshold to a low value) results in the synapses representing shared features not stabilizing (Figure 12–figure supplement 3). With one calcium threshold, in the case of the NFBP none of the synapses stabilize (as occurs when both thresholds are fixed in Fig. 8). Comparing Figure 12–figure supplement 3B3, B4 to Figs. 5B3, B4 and 6B3, B4 where just one of the two calcium thresholds is fixed, we see that having two separate thresholds allows for separate control over synaptic strengthening and weakening.

The reason why a single plasticity threshold works as well as two separate ones on the FBP, NFBP and reversal learning is that [Ca]NMDA and [Ca]L-type are driven by the same variable synaptic voltage. Both their amplitudes, which are used to compare against the thresholds and the full [Ca]NMDA and [Ca]L-type time-courses are highly correlated (Figure 12 – figure supplement 4), meaning that the two signals carry the same information, and it is not necessary to have two separate calcium variables to track that information. In this study, we have triggered metaplasticity in both thresholds using the same signal (a calcium source should be above its threshold, followed by a dopamine signal; or simply a dopamine signal with “relaxed” metaplasticity). If metaplasticity in the LTP and LTD thresholds is triggered under different conditions for each, then two separate thresholds could be useful.

Discussion

In this article we studied the effects of metaplasticity in two separate calcium thresholds for inducing synaptic plasticity, one of which is a threshold for LTP, and the other for LTD. Plasticity occurs when calcium levels from different sources, implicated in synaptic plasticity in SPNs, reach levels above the thresholds. We used the FBP and NFBP because the patterns in these tasks share their features, and the role of each threshold is clearly demonstrated on the synapses representing these shared features. Metaplasticity slides the thresholds towards the calcium amplitudes evoked by arriving patterns. Two different metaplasticity rules were tested: i) one where calcium levels need to be above the thresholds to trigger metaplasticity (“thresholded” metaplasticity), and ii) one where any calcium levels trigger metaplasticity (relaxed conditions). The results showed that:

The FBP and NFBP can be solved with this simple plasticity and metaplasticity rule, which is required in order to demonstrate the role of each calcium threshold. With this rule, metaplasticity is necessary for solving the NFBP, whereas in the FBP it ensures robust and consistently high task performance.

Metaplasticity enables synapses undergoing both LTP and LTD during learning to selectively express (converge to) only one form of plasticity (either LTP or LTD).

Metaplasticity in the LTD threshold protects learned patterns from being weakened, allowing the LTP process to properly strengthen (encode) the patterns. In effect, the threshold for weakening allows proper synaptic strengthening.

Metaplasticity in the LTP threshold prevents unnecessary features from being strengthened, allowing the LTD process to persistently weaken them. In effect, the threshold for strengthening allows proper synaptic weakening.

For successful learning of the NFBP, metaplasticity in both thresholds needs to occur during a synaptic modification, irrespective of whether the modification is LTP or LTD.

Relaxing the conditions for metaplasticity so that it is triggered by any calcium levels allows reversal learning in the FBP and NFBP to be solved, demonstrating metaplasticity’s importance for solving the plasticity–stability dilemma. Alternatively, assuming a signal exists that can reset the calcium thresholds, reversal learning can be solved without relaxing the conditions for metaplasticity.

Using a single calcium threshold instead of two separate ones also solves the FBP, NFBP and reversal learning, but does not allow for separate control over synaptic strengthening and weakening.

The role of metaplastcity in this and other studies

In the original BCM rule, metaplasticity prevents runaway growth or collapse of the synaptic weights, which is one of its most commonly cited roles (Bienenstock et al., 1982; Abraham, 2008; Hulme et al., 2013). This is achieved by using a nonlinear function in the update rule for the threshold. Conversely, here we used a linear function in the metaplasticity rule for sliding the thresholds towards the calcium amplitudes, so metaplasticity by itself cannot prevent the weights from unbounded growth (unless that growth produces saturation in the evoked calcium amplitude). Instead, metaplasticity has a different role in this study – it enables synapses exposed to competing LTP and LTD drives to converge to one stable plasticity state, i.e. either LTP or LTD – without metaplasticity, the synapses never stabilize. Although not explicitly demonstrated, the same role of metaplasticity can be inferred from the dynamics of the cascade model (Fusi et al., 2005; Farashahi et al., 2017). In fact, the calcium thresholds in the current rule are a possible mechanistic implementation of the abstract hidden metaplastic states in the cascade model: threshold sliding is analogous to moving up and down the cascade of states, making synaptic plasticity easier or more difficult to occur. (Hence the similar role of metaplasticity in both studies.)

More specifically, this study shows that the two calcium thresholds have opposite and complementary roles in regulating the strengthening and weakening of synapses. The ultimate effects of metaplasticity are unexpected, at least at first glance, since the threshold for one plasticity type (e.g. LTD) allows proper expression of the other plasticity type (e.g. LTP). In addition to separately controlling the two plasticity processes, two thresholds allow implementing a different update rule for each threshold (a case which we have not studied here). Generally speaking, two thresholds would be useful if the different calcium concentrations that they track are uncorrelated, i.e. exhibit separate dynamics, which does not seem to be the case for the NMDA and L-type channels in the dSPNs. They could also be useful if metaplasticity in each threshold is triggered by a different signal (a case which we also did not study here, but only touched upon with partial metaplasticity). Having two thresholds might also contribute to increased robustness of the machinery for synaptic plasticity across different conditions, a common feature of biological systems (Kitano, 2004; Wollman, 2018).

Finally, in a range of studies using the same threshold for all synapses, derived from the neuron’s activity (similarly to Bienenstock et al. (1982)), metaplasticity provides a homeostatic, heterosynaptic effect which helps to keep a neuron’s firing rate stable (Jedlicka et al., 2015; Zenke et al., 2013). The learning rule in this article does not incorporate such effects. As a result, without an upper bound on the synaptic weight, synapses would increase indefinitely in the case of FBP with linear integration, causing a depolarization block in the neuron. Similarly, without the upper threshold for LTP, they would also increase indefinitely when using supralinear integration. However, in the latter case the plateau potential provides stability of the neuronal firing rate, since the dendritic voltage is limited by the NMDAR reversal potential.

Directions for future studies

This study also provides a basis for exploring other metaplasticity rules. Here we only considered metaplasticity evoked in the same synapses activated by presynaptic spikes (homosynaptic metaplasticity). However, heterosynaptic metaplasticity mechanisms also exist, where the evoking stimuli originate from different synapses or different neurons (Abraham, 2008; Abraham et al., 2019; Hulme et al., 2014). Furthermore, as detailed calcium models for other neurons are developed, the contributions of different calcium sources in triggering synaptic plasticity can be studied using this simple learning rule as a basis. Regarding SPNs specifically, more realistic metaplasticity rules can also be developed once future experimental findings appear.

Some examples of different metaplasticity mechanisms are known from other brain areas. In the Schaffer collateral synapses to the CA1 hippocampal area, prior activation of NMDARs (which does not induce long-term plasticity) temporarily increases the LTP threshold and prevents LTP, while prior activation of mGluRs facilitates LTP (Huang et al., 1992; Cohen and Abraham, 1996). Additionally, weakening these synapses lowers the LTP threshold and raises the LTD threshold, while weakening does the opposite, i.e. lowers the LTD threshold and likely raises the LTP threshold (Ngezahayo et al., 2000). In the synapses from the perforant path to the dentate gyrus, prior stimulation at theta frequency (typical for the hippocampus) lowers the threshold for both LTP and LTD (Christie et al., 1995; Christie and Abraham, 1992). Additionally, prior stimulation at a high frequency also lowers the LTD threshold in both unchanged and in previously strengthened synapses (Holland and Wagner, 1998). Adapting the learning rule to incorporate the metaplasticity mechanisms in different neurons will allow to study whether metaplasticity supports other specific functions in various brain regions.

In addition to signals arising from the local neuronal microcircuitry, it is interesting to see if global signals reaching many brain areas might also have a role in triggering metaplasticity. The noradrenergic neuromodulatory system is involved in detecting unexpected changes in learned pattern-outcome associations, and has been proposed to transiently interrupt or reset ongoing neural processing (Jordan, 2024; Sara, 2009; Pajkossy et al., 2023; Yu and Dayan, 2005; Dayan and Yu, 2006). Noradrenaline is known to modulate synaptic plasticity through adrenergic receptors, such as the phosphorylation level of DARPP-32 (Hara et al., 2010), an important protein for corticostriatal synaptic plasticity. Hence, it is interesting to ask whether noradrenergic signaling can act as a global metaplastic signal that transiently modifies plasticity thresholds.

An important direction for future work is to develop a learning rule that also models structural plasticity. Since NMDAR-dependent supralinear integration relies on synaptic clustering within a short dendritic segment, or more widely, within a whole branch (Losonczy and Magee, 2006; Branco et al., 2010), a learning rule would ideally implement mechanisms that can give rise to such clusters. Synaptic clustering occurs already in development, and learning in the motor cortex, for example, builds on existing functional clusters by growing new spines into them (Winnubst and Lohmann, 2012; Hedrick et al., 2022). Motor learning also strengthens the synaptic connections from the motor cortex to the striatum, which are themselves clustered (Hwang et al., 2022). A rule incorporating structural plasticity could then also include thresholds specific to structural plasticity (Ueda et al., 2022).

Assumptions in the learning rule

We have made two assumptions when formulating the learning rule. The first is that metaplasticity operates in cortico-striatal synapses, which is not a strong assumption, since metaplasticity has been reported for many synapses. The second – a major assumption – is the upper threshold for synaptic strengthening, which means that high [Ca]NMDA during LTP conditions (a dopamine peak) prevents synaptic plasticity (blocks LTP). It is an open question whether this assumption is true. Nevertheless, there are some molecular mechanisms which could form the basis for such an upper threshold. In the neuromodulatory branch of the signaling network in Fig. 1 (the branch activated by dopamine), the enzyme adenylyl cyclase 5, which transduces the dopamine signal from the D1R, is inhibited by calcium. This indicates that high [Ca]NMDA might prevent synaptic strengthening already very early in the signaling network. Further downstream in the signaling network, calcium activates PP2B through calmodulin, and PP2B inactivates PKA, an enzyme necessary for synaptic strengthening (Church et al., 2021; Yagishita et al., 2014). This is another route through which high [Ca]NMDA could prevent LTP. Due to the complexity of the signaling network, the influence of high [Ca]NMDA on synaptic plasticity requires further study, and kinetic models of the cortico-striatal synaptic signaling network which incorporate the many molecular interactions might provide some answers (Lindskog et al., 2006; Nakano et al., 2010; Nair et al., 2015). Finally the astrocytically secreted S100B protein could be involved in limiting the amount of LTP within certain bounds (Jones, 2015).

Insights into nonlinear computation by neurons

Finally, this study gives insights into nonlinear computation by single neurons, which have a variety of mechanisms for nonlinear input processing, giving them computational abilities similar to those of multilayer artificial neural networks (Gidon et al., 2020; Poirazi et al., 2003; Jones and Kording, 2021; Major et al., 2013; Antic et al., 2010; Tran-Van-Minh et al., 2015). Metaplasticity allows dSPNs to perform the NFBP, a computationally difficult task that relies on supralinear integration by the clustered synapses. In the simpler FBP, which can be solved with either linear or supralinear integration of synaptic inputs, metaplasticity ensures robust, consistently high performance on the task.

We have recently participated in another study with a different goal – whether SPNs can solve the NFBP (Khodadadi et al., 2025). (The NFBP is related to the XOR problem, the canonical minimal example of a linearly non-separable task requiring nonlinear computation beyond a single-layer perceptron.) Khodadadi et al. (2025) uses a different learning rule that incorporates more biological complexity/detail and focused on dendritic integration which was much less supralinear, i.e. the dendrites were in a regime where they produced so-called “boosting” nonlinearities, that are likely common physiologically. Completely solving the NFBP required both excitatory and inhibitory synaptic plasticity, regardless of whether the synaptic clusters were closer or further away from the soma. Although here we study the role of metaplasticity, the results also add some information related to the solution of the NFBP. The all-or-none plateau potentials in this study achieve a very high score on the NFBP with excitatory plasticity alone. This indicates that with a threshold nonlinearity instead, the SPNs might be able to completely solve the task without requiring inhibitory synapses and even for very distally located synaptic clusters. Whether this is true will be explored in another study, as theoretical considerations suggest that the NFBP can be solved with excitatory synapses only (Tran-Van-Minh et al., 2015).

In summary, at least when it comes to solving the NFBP, the proposed learning rule highlights the importance of deciding when to stop updating the synaptic weights. The two processes (mechanisms) that regulate when synaptic plasticity occurs are the upper plasticity threshold for LTP and metaplasticity. Without the upper plasticity threshold, all strengthened synapses would reach their maximally attainable synaptic weight. Without metaplasticity, the synaptic weights would not stabilize.

Methods

Plasticity rule

We highlight again that, similar to Khodadadi et al. (2025), the learning rule is:

local: only signals at the local synaptic site are used to trigger plasticity,

calcium- and reward-based: local calcium levels trigger plasticity, the type of which is decided by the dopamine reward signal, and

always “on”: no separate training and testing phases are needed because learning is regulated by metaplasticity.

Differently from Khodadadi et al. (2025), the learning rule has metaplasticity in both calcium thresholds, and is also (much) simpler, making it suitable for studying the thresholds’ roles in synaptic plasticity.

The parameters used in the plasticity rule are given in Table 1. The values of wmax and wmin are not varied throughout the simulations. They describe the maximal and minimal values that the synaptic weight can take on in the model. Whether these values are reached depends on the metaplasticity mechanisms (the values of the calcium thresholds), and, in the case of supralinear integration, the upper plasticity threshold ΘLTP, which is a fixed parameter that can be set at any value in the region of the supralinearity (the jump) in [Ca]NMDA in Fig. 2D3 (we used 40 μM). The learning rate η and the metaplasticity rate ηθ are also kept fixed throughout all simulations, except in Figure 4–figure supplements 6–11, where they are varied to study their effect on learning. Finally, the LTP and LTD thresholds, θLTP and θLTD, are initialized to low values, θLTP, 0 and θLTD, 0, which are also the minimal values that these thresholds can take on. They are initialized to low values so that synapses are flexible (“unlocked”) for learning. Otherwise, if they are set to high values, they may act to protect synapses from weakening (high θLTD, 0) or prevent them from strengthening (high θLTP, 0), as is shown with Figs. 5, 6 and Figure 9 – figure supplement 2. The synaptic weights are initialized uniformly at random from the interval [0.3, 0.35].

Fianlly, the rule uses the amplitudes of [Ca]NMDA and [Ca]L-type to trigger plasticity (and metaplasticity, in the scenarios with “thresholded” metaplasticity), implicitly assuming that such values can be remembered by the synapse’s molecular circuitry while dopamine feedback arrives. At least within the stimulation protocol in Fig. 3A, this assumption is justified, since dopamine feedback up to 2 seconds after synaptic stimulation can trigger LTP in experiments (Yagishita et al., 2014).

Neuron model

We use a morphologically realistic, multicompartmental model of a direct-pathway striatal projection neuron from the collection of 71 such models in Lindroos and Hellgren Kotaleski (2020) (index 34 in the collection). As detailed in Lindroos and Hellgren Kotaleski (2020), the ionic conductances in all models in the collection have been selected with a search process to produce frequencycurrent (F-I) responses and dendritic calcium elevations arising from backpropagating action potentials that match experimental data, and in addition produced a suitable number of spikes to injected current, no voltage oscillations leading up to the spikes, proper afterhyperpolarization behavior, and an F-I slope within three standard deviations of the experimentally determined mean.

The main contributor to supralinear integration in the SPN model is the activation of NMDARs. This has been shown, for example, in Fig. S7 in Du et al., 2017, where supralinear integration with clustered synapses is similar in an active and a passive dendrite in the SPN. This means that the NMDA synapses, and not the SPN’s ion channels, are causing it. SPN–specific ion channels do have a large effect on the SPN’s elecrophysiological behavior – the lowered resting potential of around -85 mV is a result of the strongly inwardly rectifying K+ channels (Nisenbaum and Wilson, 1995; Wilson, 2005). Because of this, to raise their somatic voltage, the SPNs require strong synaptic drive from distributed synaptic inputs across the dendrites, or a localized activation of clustered synapses which generates a plateau potential. With this behavior, SPNs are said to have two states: a “down state”, which requires strong synaptic drive to reach the firing threshold, and an “up state”, where they are more sensitive to incoming synaptic inputs (Plotkin et al., 2011). It has also been suggested that plateau potentials themselves cause the transition from a “down” state to an “up” state (Oikonomou et al., 2014). Plateau potentials in SPNs normally do not cause somatic spiking without additional synaptic inputs, which is why in the simulations we have used elevated background noise.

In this study, synapses are placed on spines, which are not present in the original model in Lindroos and Hellgren Kotaleski (2020). The spines consist of a neck and head with neck and head with lengths and diameters lneck = 0.5μm, lhead = 0.5μm, and dneck = 0.125μm, dhead = 0.5μm, respectively. Axial resistance in all compartments of the model is 150 Ω ⋅ cm, except for the spine neck, where it is 1130 Ω ⋅ cm (Dorman et al., 2018). The spines contain the inwardly rectifying K+ channel, and the same voltage-gated calcium channels as in the dendrites. The spine calcium channel conductances have been manually tuned to match the relative proportions determined in Carter and Sabatini (2004) and Higley and Sabatini (2010), as well as the Ca2+ concentration amplitudes arising from backpropagating action potential (bAP) stimulation as in Fig. 2 of Shindou et al. (2011).

Calcium dynamics and diffusion model

Because calcium from two different sources triggers LTP and LTD in SPNs, we tracked calcium from the two sources separately. Calcium from NMDA channels, when followed by a peak in dopamine, triggers LTP (Shen et al., 2008; Yagishita et al., 2014; Fino et al., 2010; Fisher et al., 2017). Calcium from Cav1.3 channels, when paired with a pause in dopamine, has been shown to be necessary for LTD (Wang et al., 2006; Fino et al., 2010; Shindou et al., 2011; Iino et al., 2020).

Without implementing calcium diffusion explicitly, the accumulation of [Ca]NMDA in a spine shows a strange effect when the synaptic cluster size is gradually increased, which is that the largest plateaus do not correspond to the largest calcium amplitude (Figs. 13A, B). This is simply a result of less NMDAR membrane current flowing in as the plateau’s voltage approaches the NMDAR reversal potential in Eq. 8. Larger plateaus giving rise to lower [Ca]NMDA amplitudes could pose a problem to the learning rule, since the decision when to trigger plasticity and metaplasticity is made by comparing the [Ca]NMDA amplitude against the LTP threshold. In a synaptic cluster that is being strengthened, but starts to evoke a smaller [Ca]NMDA amplitude after some strengthening due to the effect in Figs. 13A, B, learning might be prematurely stopped as the [Ca]NMDA amplitude falls below the increased LTP threshold.

Alternative models of calcium accumulation.

(A) Spine [Ca]NMDA without axial diffusion. [Ca]NMDA is described with a pool model similar to that in Eq. 6 (except that instead of [Ca]L-type, this pool accumulates [Ca]NMDA arising from the NMDA current in Eq. 8). (B) Spine [Ca]NMDA + [Ca]VGCC without axial diffusion. Calcium from the NMDA current and all voltage-gated calcium channels (except L-type) accumulates in a pool model. (C) Spine [Ca]NMDA with axial diffusion. Compared to Fig. 2C3, in this plot only calcium from NMDARs diffuses in the neuron, without calcium from other voltage-gated calcium channels. Figure 13—figure supplement 1. Axial diffusion of calcium from NMDARs only.

Because of this, we sought to obtain a monotonic increase of calcium amplitude with increasing plateau voltage (as in Fig. 2C3), which seems to also occur in experiments where plateaus are evoked (Fig. 2C2, D2 in Oikonomou et al. (2012)). Increasing cluster size when evoking NMDA spikes also produces monotonic increases in calcium amplitude, at least for the voltages achieved by NMDA spikes (Figs. 5 and 8 in Losonczy and Magee (2006)). To obtain such monotonic increase, it was necessary to implement axial diffusion and buffering for [Ca]NMDA. Because tuning a detailed calcium diffusion model to data is a project in its own right, we used the most detailed calcium diffusion model for SPNs available to date (by Dorman et al. (2018)) as a reference. We reuse the same buffer molecules, their amounts and their calcium binding and unbinding rates from Dorman et al. (2018) with small changes, given in Table 2. The rationale for reusing them is that the idealized morphology used in Dorman et al. (2018) should on average represent any other SPN, so the parameters obtained from the optimization procedure in that study should on average also apply to other multicompartment SPN models. The model in Dorman et al. (2018) also includes a plasmamembrane Ca pump (PMCA) in the soma, dendrites and spines, which we added in this model, and a sodium/calcium exchanger in the spines, which we ommited, because with the parameters from Dorman et al. (2018), it did not noticeably alter the spine [Ca]NMDA concentration. In the calcium diffusion model in Dorman et al. (2018), calcium originating from all sources, which in that study are NMDARs and voltage–gated calcium channels (VGCCs), and a minor, 0.1% contribution of the AMPAR current, diffuses throughout the SPN model. Because the calcium buffer parameters are optimized to such a scenario, we also included calcium from the Cav2.1 (Q-type), Cav2.2 (N-type), Cav2.3 (R-type), Cav3.2 and Cav3.3 (T-type) channels in the SPN together with [Ca]NMDA in the diffusion model. Indeed, non-selective blockade of voltage-gated calcium channels with mibefradil has shown that they are also necessary for inducing LTP in SPNs (Fino et al., 2010). Mibefradil has been shown to block T-type, L-type, R-type, N-type and P/Q-type calcium channels in different experimental preparations, so, apart from the L-type channels, we included all of them in the diffusion model (Randall and Tsien, 1997; Viana et al., 1997; Aczél et al., 1998; Eller et al., 2000; Leuranguer et al., 2001). What the role of each of these channels is in LTP in SPNs requires further study with selective blockers.

We also show how diffusion of only [Ca]NMDA looks like in the SPN in Figure 13–figure supplement 1 (and Fig. 13C). The difference with the diffusion model used throughout the study, which includes VGCCs, is minor (Figure 13–figure supplement 1, cf. Figs. 2C, D). The qualitative and quantitative shapes of the voltage and calcium elevations are very similar (Figure 13–figure supplement 1A, cf. Fig. 2C). Regarding the amplitudes of these quantities, there is a 1 mV increase in the evoked somatic amplitude after glutamate spillover when only [Ca]NMDA diffuses (Figure 13– figure supplement 1B1), and an obvious difference in the amplitude of [Ca]NMDA (Figure 13–figure supplement 1B3). Nevertheless, the threshold nonlinearity in the [Ca]NMDA amplitude is still present, and the learning rule would work the same provided the upper plasticity threshold ΘLTP is placed in the region of the nonlinearity.

Even though mibefradil also affects L-type channels, we did not include them in the axial diffusion model together with [Ca]NMDA because L-type calcium channels frequently form calcium microdomains, where calcium entry and diffusion is restricted within a very small, micrometer range (Berridge, 2006; Chen and Sabatini, 2012; Parekh, 2008). In the microdomain, the L-type channels can be physically coupled in supramolecular protein complexes, further contributing to the local effect of the calcium influx in activating the relevant signaling networks, and this is also the case for Cav1.3 channels in SPNs (Olson et al., 2005; Stanika et al., 2015). Because of this, we assume that such highly localized [Ca]L-type signals trigger LTD in the synapse, and calculated [Ca]L-type with a separate, phenomenological pool model, which describes [Ca]L-type in a thin cylindrical shell under the membrane. The implementation follows that of Wolf et al. (2005), according to Eq. 6. In this model [Ca]L-type decays with a time constant τCa, modeling Ca2+ diffusion only phenomenologically, i.e no Ca2+ ions move between dendritic compartments. Compared to the very simplest pool models, this pool model also includes a Ca2+ pump that extrudes intracellular Ca2+ to the extracellular space. F is the Faraday constant, kt is the catalytic activity of the pump, Kd is the pump dissociation constant for Ca2+, d is the depth of the submembrane shell where Ca2+ accumulates, and k and p are phenomenological parameters used in Wolf et al. (2005) used to balance Ca2+ influx and efflux. The values for the parameters in Eq. 6 are given in Table 3.

(6) d[Ca]L-typedt=[Ca]∞−[Ca]L-typeτCa+k2Fd∑x∈{1.2,1.3}ICavx−pkt[Ca]L-type[Ca]L-type+Kd In this pool model we included calcium from both Cav1.2 and Cav1.3 channels, since there are no pharmacological agents that selectively block only one of the channels, i.e. the compounds used in the experiments which have determined Cav1.3 as necessary for LTD in cortico-striatal synapses would also block Cav1.2 channels in SPNs (Berger and Bartsch (2014); nimodipine in Wang et al. (2006) and Shindou et al. (2011); also, other non-selective L-type channel blockers are used to block LTD in the following studies: nifedipine in Calabresi et al. (1994); nitrendipine in Kreitzer and Malenka (2005); mibefradil in Fino et al. (2010); finally, experiments in Plotkin et al. (2013) have suggested that other VGCCs than Cav1.3 may be involved in activating calcium-induced calcium release, which is necessary for LTD).

Synaptic inputs

There are two different models of synapses used in this study. One model – the dual exponential model – is used in describing the background noise, which consists of glutamatergic and GABAergic synapses, distributed across the dendrites according to the distribution measured in Cheng et al. (1997). These synapses are fixed (they have no plasticity) because they are activated at a low rate of 1 Hz. Compared to the 30 ms activation window in Fig. 3, this rate is low enough so that we decided to neglect any background noise synapses that might be co-active in the same time window and thus become eligible for plasticity together with the feature-carrying synapses. Because there are many synapses representing the background noise, this assumption also lowers the computational cost – if they were plastic, the learning rule would need to be computed for each of them.

The other model, used to describe the synapses carrying the features, is a saturating synapse model. We chose a saturating synapse model because we activate these synapses with 3 spikes each within a 30 ms window, and wanted to avoid any unrealistically high conductance values that could otherwise arise in non-saturating models.

Dual exponential model

The dual exponential synaptic model (a difference of two exponential functions) for a spike arriving at time 0 is represented by: (7) gsyn(t)=gmax1CτAτBτB−τA(exp⁡(−t−t0τB)−exp⁡(−t−t0τA)) where gmax is the maximal synaptic conductance, and τA and τB are the rise and decay time constants. C is a normalization factor given with: C=τAτBτB−τA(exp⁡(−t∗τB)−exp⁡(−t∗τA))t⋆=τAτBτB−τAln⁡τBτA where t⋆ is the position of the maximum of the dual exponential function. This normalization ensures that the time-varying part of the dual exponential reaches a peak value of 1, which is then scaled by the maximal conductance gmax. (The implementation in NEURON contains a synaptic weight parameter, as well. Since in the dual exponential model we have fixed the weight to 1, it is omitted from Eq. 7.)

The synaptic current arising from the conductance gsyn is: Isyn=gsyn(V−Erev) Excitatory synapses have both AMPA and NMDA components, and NMDA synapses in addition have a Mg2+ block described by a sigmoidal function: (8) INMDA=gNMDA(V−Erev)σ;σ=11+ηexp⁡(−αV), with parameters α = 0.062 and η = 0.38. The parameters of the dual exponential model for the glutamatergic AMPA and NMDA synapses, and the GABAergic synapses are given in Table 4.

Finally, because the SPN model contains less electrical compartments than the number of synapses reported in Cheng et al. (1997), this implies that several background noise synapses will arrive in a single compartment. As such, their synaptic inputs will be integrated in the same voltage variable. Because of this, we have represented such multiple synapses arriving in one compartment with a single synapse instead, whose input frequency is scaled by the number of synapses it represents. This is why a non-saturating synapse model, such as the dual exponential model, was necessary for the background noise synapses. This also greatly increases simulation speed.

Saturating synapse model

The saturating synapse models are taken from Gao et al. (2021), which are a variation of the saturating synapse models in Destexhe et al. (1994). They are kinetic models which operate according to two different kinetic schemes depending on the presence or absence of neurotransmitters. When neurotransmitters arrive at the postsynaptic site, receptor dynamics are described by a kinetic scheme that models switching between closed (C) and open (O) states with rates of α and β: C+T⇌αβO where [T] is the transmitter concentration. After neurotransmitter has been cleared from the synaptic cleft, receptor dynamics evolves according to a kinetic scheme that describes just the closing of the receptor with a rate β: C←βO In this model, when a presynaptic spike arrives, neurotransmitter levels always reach a fixed saturating concentration in the synaptic cleft, [T]max, which lasts for a short time Tdur (a short glutamate lifetime in the synaptic cleft of cultured hippocampal synapses has been reported in Clements et al. (1992)). If another presynaptic spike arrives while the neurotransmitter pulse is still “on”, the pulse duration is lengthened by Tdur.

The synaptic conductance gsyn in this model is represented by the fraction of glutamate receptors in the open state, [O]: gsyn=wgmax[O] where gmax is again the maximal synaptic conductance, and w is the synaptic weight which is modified by the learning rule. The parameters in the saturating synapse model are given in Table 5.

The fraction of NMDAR current that contributes to calcium influx is 17.5%, between the values of 15.9% for receptors with GluN1-GluN2A subunit composition (Table 18 in Traynelis et al. (2010)) and 18.5% (Schneggenburger, 1996).

Glutamate spillover

To produce plateau potentials which have an all-or-none jump in the voltage, i.e. a threshold nonlinearity (as suggested by existing in vitro experiments), we included glutamate spillover. A detailed study of glutamate spillover and its role in triggering plateau potentials is given in Trpevski et al. (2023), and here we briefly describe the model. Glutamate spillover is a phenomenon that occurs during repeated synaptic stimulation, when glutamate reuptake by the surrounding astrocytes is saturated and when the astrocytes may even reverse function from reuptake to glutamate releas (Malarkey and Parpura, 2008). When this happens in a synaptic cluster, glutamate spills over from the synaptic cleft into the extrasynaptic space, activating extrasynaptic NMDARs (eNMDARs). To model glutamate spillover, we place eNMDARs in the dendritic shaft, at the base of each spine in a cluster. Glutamate spillover is activated when a threshold level of clustered synaptic stimulation is reached – the “glutamate threshold”. In this study, the glutamate threshold is reached when 20 synapses with a weight w > 0.5 are activated (with 3 spikes each). Note that this threshold is equivalent to activating 10 synapses with a weight w > 1.0, or 40 synapses with a weight w > 0.25, i. e. the summed weight of all activated clustered synapses should be greater than 10.

When glutamate spills over, i) an activating spike to the eNMDARs is delivered 1.5 ms after the glutamate threshold is reached, ii) it stays much longer around the eNMDARs than in the synaptic cleft (Tdur is much longer and depends on the synaptic weight w), and iii) its concentration around the eNMDARs is much lower than in the synaptic cleft ([T]max is much lower) (Table 5 and Trpevski et al. (2023)). Also, the eNMDARs represent about 64% of the total NMDA conductance in this study (Table 5).

Stimulation protocol

The stimulation protocol in Fig. 3A for presenting the patterns (in random order), which are followed by dopamine feedback, contains several additional details that we list here. First of all, when the distributed or clustered synapses are activated within the 30 ms stimulation window, 3 input spikes arrive to each synapse within this window (as explained in the caption of Fig. 3). This is because at physiological extracellular calcium concentrations, single spikes do not induce plasticity (at least in the spike-timing-dependent protocols in Inglebert et al. (2020)). Furthermore, the backward rate constants from Dorman et al. (2018) for calcium unbinding from the buffers are much lower than experimentally measured values for calmodulin and calbindin, making full calcium unbinding and equilibration of the calcium dynamics very long (Faas et al., 2011; Nägerl et al., 2000). To avoid long simulation times on the computing cluster used for this study, 600 ms after a pattern is presented, we reset the values of the buffers, their bound forms to calcium, and the intra- and extracellular calcium concentrations back to their initial values before the first pattern is presented, thus “equilibrating” the system faster and shortening simulation time. As can be seen from Fig. 2C, by 600 ms the plateaus are over, and both [Ca]NMDA and [Ca]L-type are back to their baseline levels, so this abrupt reset of the system does not affect learning and synaptic plasticity. And as can be seen from Fig. 3C, it neither affects the baseline voltage in the soma and dendrites before the arrival of a future pattern.

Performance scores on the FBP and NFBP

In the FBP, a score of 50% means that the SPN is correct only half of the time. This happens if the SPN is silent for both the relevant and irrelevant pattern (correct output only for the irrelevant pattern), or if it spikes for both patterns (correct output only for the relevant pattern). A score of 100% indicates it always spikes for the relevant pattern, and is always silent for the irrelevant pattern, and the FBP is perfectly solved. The threshold score for solving the FBP is 75%, exemplified with the situation where the SPN is always silent for the irrelevant pattern, and spikes at least half of the time for the relevant pattern. A score of 0% would indicate that the SPN’s output is incorrect all the time, being silent for the relevant pattern, and spiking for the irrelevant pattern (this does not happen in the simulations).

In the NFBP, a score of 50% also means that the SPN output is correct only half of the time. An example of such a score is when the SPN is silent for all four patterns, or spikes for all four patterns. Similarly, a score of 100% indicates it always spikes for the relevant patterns, and is always silent for the irrelevant patterns. A score of 75% is exemplified with the situation where the SPN is silent for the two irrelevant patterns, and spikes for only one of the relevant patterns. The threshold score for solving the NFBP is 87.5%, where the SPN would also spike for the other relevant pattern at least half of the time. Alternatively, a score of 75% can be achieved by the SPN spiking for both relevant patterns, but being silent for only one irrelevant pattern. Similarly, the threshold score of 87.5% can be reached by being silent for the other irrelevant pattern at least half of the time.